---
title: "Inference from linear models"
subtitle: "Analysis of Ecological and Environmental Data<br>QERM 514"
author: "Mark Scheuerell"
date: "10 April 2026"
output:
  ioslides_presentation:
    wide: true
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(faraway)
```

## Goals for today

* Understand the use of *F*-tests for hypothesis testing

> - Understand how to estimate confidence intervals


## Inferential methods

Once we've estimated the model parameters and their variance, we might want to draw conclusions from our analysis


## Comparing models

Imagine we had 2 linear models of varying complexity: 

1. a model with one predictor

2. a model with five predictors

It would seem logical to ask whether the complexity of (2) is necessary?


## Hypothesis test to compare models

Recall our partitioning of sums-of-squares, where

$$
SSTO = SSR + SSE
$$

We might prefer the more complex model (call it $\Theta$) over the simple model (call it $\theta$) if 

$$
SSE_{\Theta} < SSE_{\theta}
$$

or, more formally, if

$$
\frac{SSE_{\theta} - SSE_{\Theta}}{SSE_{\Theta}} > \text{a constant}
$$


## Hypothesis test to compare models

If $\Theta$ has $k_{\Theta}$ parameters and $\theta$ has $k_{\theta}$, we can scale this ratio to arrive at an $F$-statistic that follows an $F$ distribution

$$
F = \frac{ \left( SSE_{\theta} - SSE_{\Theta} \right)  / (k_{\Theta} - k_{\theta})}{ SSE_{\Theta}  / (n - k_{\Theta})} \sim F_{k_{\Theta} - k_{\theta}, n - k_{\Theta}}
$$


## $F$-distribution

The $F$-distribution is the ratio of two random variates, each with a $\chi^2_{n}$ distribution

If $A \sim \chi_{df_{A}}^{2}$ and $B \sim \chi_{df_{B}}^{2}$ are independent, then

$$
\frac{\left( \frac{A}{df_{A}} \right) }{ \left( \frac{B}{df_{B}} \right) }  \sim F _{df_{A},df_{B}}
$$


## $F$-distribution

```{r f_dist, fig.height=4, fig.width=6, fig.align="center"}
par(mai = c(0.5,0.5,0.1,0.1), omi = rep(0,4))
curve(df(x, df1=2, df2=1), from=0, to=4, lwd = 2, col = "orange",
      ylab = "", xlab = expression(italic(x)), las = 1)
curve(df(x, df1=4, df2=10), from=0, to=4, lwd = 2, col = "red", add = TRUE)
curve(df(x, df1=6, df2=25), from=0, to=4, lwd = 2, col = "blue", add = TRUE)
curve(df(x, df1=8, df2=50), from=0, to=4, lwd = 2, col = "purple", add = TRUE)
legend(3, 0.9, legend = c("df1 = 2, df2 = 1",
         "df1 = 4, df2 = 10",
         "df1 = 6, df2 = 25",
         "df1 = 8, df2 = 50"),
       col = c("orange", "red", "blue", "purple"),
       lwd = 1.1, bty = "n", xjust = 0.5, y.intersp = 1.2)
```


## Test of *all* predictors in a model

Suppose we wanted to test whether the collection of predictors in a model were better than simply estimating the data by their mean.

$$
\Theta: \mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{e} \\
\theta: \mathbf{y} = \boldsymbol{\mu} + \mathbf{e} \\
$$

We write the null hypothesis as

$$
H_0: \beta_1 = \beta_2 = \dots = \beta_k = 0
$$

and we would reject $H_0$ if $F > F^{(\alpha)}_{k_{\Theta} - k_{\theta}, n - k_{\Theta}}$


## Hypothesis test to compare models

$$
SSE_{\Theta} = \left( \mathbf{y} - \mathbf{X} \boldsymbol{\beta} \right)^{\top} \left( \mathbf{y} - \mathbf{X} \boldsymbol{\beta} \right) = \mathbf{e}^{\top} \mathbf{e} = SSE \\
SSE_{\theta} = \left( \mathbf{y} - \bar{y} \right)^{\top} \left( \mathbf{y} - \bar{y} \right) =  SSTO \\
\Downarrow \\
F = \frac{ \left( SSTO - SSE \right)  / (k - 1) } { SSE  / (n - k)}
$$


## Predictors of plant diversity

Later in lab we will work with the `gala` dataset<sup>$\dagger$</sup> in the **faraway** package, which contains data on the diversity of plant species across 30 Galapagos islands

For now let's hypothesize that

<center>diversity = $f$(area, elevation, distance to nearest island)</center>

<br><br>

<sup>$\dagger$</sup>From Johnson & Raven (1973) *Science* 179:893-895


## Testing one predictor

We might ask whether any one predictor could be dropped from a model

For example, can $\text{nearest}$ be dropped from ourf full model?

$$
\text{species}_i = \alpha + \beta_1 \text{area}_i + \beta_2 \text{elevation}_i + \beta_3 \text{nearest}_i + \epsilon_i
$$


## Testing one predictor

One option is to fit these two models and compare them via our $F$-test with $H_0: \beta_3 = 0$

<br>

$$
\begin{aligned}
\text{species}_i &= \alpha + \beta_1 \text{area}_i + \beta_2 \text{elevation}_i + \beta_3 \text{nearest}_i + \epsilon_i \\
~ \\ 
\text{species}_i &= \alpha + \beta_1 \text{area}_i + \beta_2 \text{elevation}_i + \epsilon_i
\end{aligned}
$$


## Testing one predictor

Another option is to estimate a $t$-statistic as

$$
t_i = \frac{\widehat{\beta}_i}{\text{SE} \left( \widehat{\beta}_i \right)}
$$

and compare it to a $t$-distribution with $n - k$ degrees of freedom


## Testing 2+ predictors

Sometimes we might want to know whether we can drop 2+ predictors from a model

For example, can we drop both $\text{elevation}$ and $\text{nearest}$ from our full model?

$$
\begin{aligned}
\text{species}_i &= \alpha + \beta_1 \text{area}_i + \beta_2 \text{elevation}_i + \beta_3 \text{nearest}_i + \epsilon_i \\
~ \\ 
\text{species}_i &= \alpha + \beta_1 \text{area}_i + \epsilon_i
\end{aligned}
$$

$H_0 : \beta_2 = \beta_3 = 0$


## Testing a subspace

Some tests cannot be expressed in terms of the inclusion or exclusion of predictors

Consider a test of whether the areas of the current and adjacent island could be added together and used in place of the two separate predictors

$$
\text{species}_i = \alpha + \beta_1 \text{area}_i + \beta_2 \text{adjacent}_i + \dots + \epsilon_i \\
~ \\
\text{species}_i = \alpha + \beta_1 \text{(area + adjacent)}_i + \dots + \epsilon_i
$$

$H_0 : \beta_{\text{area}} = \beta_{\text{adjacent}}$


## Testing a subspace

What if we wanted to test whether a predictor had a specific (non-zero) value?

For example, is there a 1:1 relationship between $\text{species}$ and $\text{elevation}$ after controlling for the other predictors?

$$
\text{species}_i = \alpha + \beta_1 \text{area}_i + \underline{1} \text{elevation}_i + \beta_3 \text{nearest}_i + \epsilon_i
$$

$H_0 : \beta_2 = 1$


## Testing a subspace

We can also modify our $t$-test from before and use it for our comparison by including the hypothesized $\beta_{H_0}$ as an offset

$$
t_i = \frac{(\widehat{\beta_i} - \beta_{H_0})}{\text{SE} \left( \widehat{\beta}_i \right)}
$$


## Caveats about hypothesis testing

Null hypothesis testing (NHT) is a slippery slope

> - $p$-values are simply the probability of obtaining a test statistic as large or greater than that observed

> - $p$-values are **not** weights of evidence

> - "Critical" or "threshold" values against which to compare $p$-values must be chosen *a priori*

> - Be aware of "$p$ hacking" where researchers make *many* tests to find significance


# QUESTIONS?


## Confidence intervals for $\beta$

We can also use confidence intervals (CI's) to express uncertainty in $\widehat{\beta}_i$

They take the form

$$
100(1 - \alpha)\% ~ \text{CI}: \widehat{\beta}_{i} \pm t_{n-p}^{(\alpha / 2)} \operatorname{SE}(\widehat{\beta})
$$

where here $\alpha$ is our *predetermined* Type-I error rate


## Bootstrap confidence intervals

The $F$- and $t$-based CI's we have described depend on the assumption of normality

The bootstrap<sup>$\dagger$</sup> method provides a way to construct CI's without this assumption

<br><br>

<sup>$\dagger$</sup>Efron (1979) *The Annals of Statistics* 7:1–26


## Bootstrap procedure

1. Fit your model to the data  

2. Calculate $\mathbf{e} = \mathbf{y} - \mathbf{X} \widehat{\boldsymbol{\beta}}$ 

3. Do the following *many* times:
    * Generate $\mathbf{e}^*$ by sampling *with replacement* from $\mathbf{e}$  
    * Calculate $\mathbf{y}^*$ = $\mathbf{X} \widehat{\boldsymbol{\beta}} + \mathbf{e}^*$
    * Estimate $\widehat{\boldsymbol{\beta}}^*$ from $\mathbf{X}$ & $\mathbf{y}^*$)  

4. Select the $\tfrac{\alpha}{2}$ and $(1 - \tfrac{\alpha}{2})$ percentiles from the saved $\widehat{\boldsymbol{\beta}}^*$


## Confidence interval for new predictions

Given a fitted model $\mathbf{y} = \mathbf{X} \widehat{\boldsymbol{\beta}} + \mathbf{e}$, we might want to know the uncertainty around a new estimate $\mathbf{y}^*$ given some new predictor $\mathbf{X}^*$


## CI for the mean response

Suppose we wanted to estimate the uncertainty in the *average* response given by

$$
\widehat{\mathbf{y}}^* = \mathbf{X}^* \widehat{\boldsymbol{\beta}}
$$

Recall that the general formula for a CI on a quantity $z$ is

$$
100(1 - \alpha)\% ~ \text{CI}: \text{E}(z) ~ \pm ~ t^{(\alpha / 2)}_{df}\text{SD}(z)
$$

So we would have

$$
\widehat{\mathbf{y}}^* ~ \pm ~ t^{(\alpha / 2)}_{df} \sqrt{\text{Var} \left( \widehat{\mathbf{y}}^* \right)}
$$


## CI for the mean response

We can calculate the SD of our expectation as

$$
\begin{aligned}
\text{Var} \left( \widehat{\mathbf{y}}^* \right) &= \text{Var} \left( \mathbf{X}^* \widehat{\boldsymbol{\beta}} \right)  \\
  &= {\mathbf{X}^*}^{\top} \text{Var}\left( \widehat{\boldsymbol{\beta}} \right) \mathbf{X}^* \\
  &= {\mathbf{X}^*}^{\top} \left[ \sigma^2 (\mathbf{X}^{\top} \mathbf{X})^{-1} \right] \mathbf{X}^* \\
  &\Downarrow \\
\text{SD} \left( \widehat{\mathbf{y}}^* \right) &= \sigma \sqrt{ {\mathbf{X}^*}^{\top} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^* } 
\end{aligned}
$$


## CI for the mean response

So our CI on the mean response is given by

$$
\widehat{\mathbf{y}}^* \pm ~ t^{(\alpha / 2)}_{df} \sigma \sqrt{ {\mathbf{X}^*}^{\top} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^* } 
$$


## CI for a specific response

What about the uncertainty in a *specific* prediction?

In that case we need to account for our additional uncertainty owing to the error in our relationship, which is given by

$$
\widehat{\mathbf{y}}^* = \mathbf{X}^* \widehat{\boldsymbol{\beta}} + \mathbf{e}
$$

## CI for a specific response

The SD of the new prediction is given by

$$
\begin{aligned}
\text{Var} \left( \widehat{\mathbf{y}}^* \right) &= {\mathbf{X}^*}^{\top} \text{Var}\left( \widehat{\boldsymbol{\beta}} \right) \mathbf{X}^* + \text{Var} \left( \mathbf{e} \right) \\
  &= {\mathbf{X}^*}^{\top} \left[ \sigma^2 (\mathbf{X}^{\top} \mathbf{X})^{-1} \right] \mathbf{X}^* + \sigma^2\\
  &= \sigma^2 \left( {\mathbf{X}^*}^{\top} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^* + 1 \right) \\
  &\Downarrow \\
\text{SD} \left( \widehat{\mathbf{y}}^* \right) &= \sigma \sqrt{1 + {\mathbf{X}^*}^{\top} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^* } 
\end{aligned}
$$

## CI for a specific response

So our CI on the new prediction is given by

$$
\widehat{\mathbf{y}}^* \pm ~ t^{(\alpha / 2)}_{df} \sigma \sqrt{1 + {\mathbf{X}^*}^{\top} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^* } 
$$

<br>

This is typically referred to as the *prediction interval*



