---
title: "More on linear models"
subtitle: "Analysis of Ecological and Environmental Data<br>QERM 514"
author: "Mark Scheuerell"
date: "6 April 2026"
output:
  ioslides_presentation:
    wide: true
    css: ../lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Goals for today

* Understand how to represent a linear model with matrix notation

> - Understand the concept, assumptions & practice of least squares estimation for linear models

> - Understand the concept of identifiability


## Simple regression model

<br>

$$
y = \text{intercept} + \text{slope} \times x + \text{error}
$$


## Simple regression model

```{r ex_regr, fig.align = 'center', fig.height = 5, fig.width = 7}
set.seed(514)

## predictor
xx <- runif(30, 0, 10)
## normal errors
ee <- rnorm(30, 0, 2)
## simulated data
yy <- 4 + 2 * xx + ee


par(mai = c(0.8,0.8,0.1,0.1))
plot(xx, yy, pch = 16,
     ylim = c(0, ceiling(max(yy))),
     cex = 1.5, cex.lab = 1.5, cex.axis = 1.2,
     xlab = expression(italic(x)), ylab = expression(italic(y)))
abline(a = 4, b = 2, lwd = 2)
## intercept
segments(0, 0, 0, 4, lwd = 3, col = "blue")
text(0, 2, "intercept = 4", pos = 4, cex = 1.5, col = "blue")
## slope
segments(4, 12, 4, 16, lwd = 3, col = "blue")
segments(4, 16, 6, 16, lwd = 3, col = "blue")
text(4, 14, "slope = 2", pos = 2, cex = 1.5, col = "blue")
```


## Linear models in matrix form | Simple regression

$$
y_i = \alpha + \beta x_{i} + e_i \\
$$

The $i$ subscript indicates one of a total $N$ observations


## Linear models in matrix form | Simple regression

Let's switch the notation for our intercept and slope

$$
y_i = \alpha + \beta x_{i} + e_i \\
\Downarrow \\
y_i = \beta_0 + \beta_1 x_{i} + e_i
$$

<br>

$^*$The reason for this notation switch will become clear later


## Linear models in matrix form | Simple regression

Let's make this general statement more specific

$$
y_i = \beta_0 + \beta_1 x_{i} + e_i \\
\Downarrow \\
y_1 = \beta_0 + \beta_1 x_{1} + e_1 \\
y_2 = \beta_0 + \beta_1 x_{2} + e_2 \\
\vdots \\
y_N = \beta_0 + \beta_1 x_{N} + e_N \\
$$


## Linear models in matrix form | Simple regression

Let's now make the implicit "1" multiplier on $\beta_0$ explicit 

$$
y_1 = \beta_0 \underline{1} + \beta_1 x_{1} + e_1 \\
y_2 = \beta_0 \underline{1} + \beta_1 x_{2} + e_2 \\
\vdots \\
y_N = \beta_0 \underline{1} + \beta_1 x_{N} + e_N \\
$$


## Linear models in matrix form | Simple regression

Let's next gather the common terms into column vectors

$$
y_1 = \beta_0 1 + \beta_1 x_{1} + e_1 \\
y_2 = \beta_0 1 + \beta_1 x_{2} + e_2 \\
\vdots \\
y_N = \beta_0 1 + \beta_1 x_{N} + e_N \\
$$


## Linear models in matrix form | Simple regression

Maybe something like this?

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
\beta_0 \\ \beta_0 \\ \vdots \\ \beta_0
\end{bmatrix}
\begin{bmatrix}
1 \\ 1 \\ \vdots \\ 1
\end{bmatrix}
+
\begin{bmatrix}
\beta_1 \\ \beta_1 \\ \vdots \\ \beta_1
\end{bmatrix}
\begin{bmatrix}
x_1 \\ x_2 \\ \vdots \\ x_N
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$


## An aside on linear algebra

We refer to the dimensions of matrices in a row-by-column manner

$[rows \times columns]$


## An aside on linear algebra

When adding matrices, the dimensions must match

$[m \times n] + [m \times n]$ &nbsp; <b>&#10003;</b>

$[m \times n] + [m \times p]$ &nbsp; <b>X</b>


## An aside on linear algebra

When multiplying 2 matrices, the inner dimensions must match

$[m \times \underline{n}] [\underline{n} \times p]$ &nbsp; <b>&#10003;</b>

$[m \times \underline{n}] [\underline{p} \times n]$ &nbsp; <b>X</b>


## An aside on linear algebra

When multiplying 2 matrices, the dimensions of the product are

[rows-of-first $\times$ columns-of-second]

$[\underline{m} \times n] [n \times \underline{p}] = [m \times p]$


## Linear models in matrix form | Simple regression

Let's check the dimensions

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
\beta_0 \\ \beta_0 \\ \vdots \\ \beta_0
\end{bmatrix}
\begin{bmatrix}
1 \\ 1 \\ \vdots \\ 1
\end{bmatrix}
+
\begin{bmatrix}
\beta_1 \\ \beta_1 \\ \vdots \\ \beta_1
\end{bmatrix}
\begin{bmatrix}
x_1 \\ x_2 \\ \vdots \\ x_N
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$


## Linear models in matrix form | Simple regression

Let's check the dimensions

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
\beta_0 \\ \beta_0 \\ \vdots \\ \beta_0
\end{bmatrix}
\begin{bmatrix}
1 \\ 1 \\ \vdots \\ 1
\end{bmatrix}
+
\begin{bmatrix}
\beta_1 \\ \beta_1 \\ \vdots \\ \beta_1
\end{bmatrix}
\begin{bmatrix}
x_1 \\ x_2 \\ \vdots \\ x_N
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$

$$
[N \times 1] = \underbrace{[N \times 1] [N \times 1]}_{\text{OOPS!}} + \underbrace{[N \times 1] [N \times 1]}_{\text{OOPS!}} + [N \times 1]
$$


## An aside on linear algebra

When multiplying a scalar times a vector/matrix, it's just element-wise

$$
a
\begin{bmatrix}
X \\ Y \\ Z
\end{bmatrix}
=
\begin{bmatrix}
aX \\ aY \\ aZ
\end{bmatrix}
$$


## Linear models in matrix form | Simple regression

So this looks better

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\beta_0
\begin{bmatrix}
1 \\ 1 \\ \vdots \\ 1
\end{bmatrix}
+
\beta_1
\begin{bmatrix}
x_1 \\ x_2 \\ \vdots \\ x_N
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$

$$
[N \times 1] = [N \times 1] + [N \times 1] + [N \times 1]
$$


## Linear models in matrix form | Simple regression

This is nice, but can we make $\beta_0$ and $\beta_1$ more compact?

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\beta_0
\begin{bmatrix}
1 \\ 1 \\ \vdots \\ 1
\end{bmatrix}
+
\beta_1
\begin{bmatrix}
x_1 \\ x_2 \\ \vdots \\ x_N
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$


## Linear models in matrix form | Simple regression

What if we move $\beta_0$ & $\beta_1$ to the other side of the predictors...

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
1 \\ 1 \\ \vdots \\ 1
\end{bmatrix}
\beta_0
+
\begin{bmatrix}
x_1 \\ x_2 \\ \vdots \\ x_N
\end{bmatrix}
\beta_1
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$


## Linear models in matrix form | Simple regression

...and group the predictors and parameters into matrices

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
1 & x_1 \\ 1 & x_2 \\ \vdots & \vdots \\ 1 & x_N
\end{bmatrix}
\begin{bmatrix}
\beta_0 & \beta_1
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$


## Linear models in matrix form | Simple regression

Let's check the dimensions

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
1 & x_1 \\ 1 & x_2 \\ \vdots & \vdots \\ 1 & x_N
\end{bmatrix}
\begin{bmatrix}
\beta_0 & \beta_1
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$

$$
[N \times 1] = \underbrace{[N \times 2] [1 \times 2]}_{\text{OOPS!}} + [N \times 1]
$$


## An aside on linear algebra

Matrix multiplication works on a row-times-column manner

$$
\begin{aligned}
\begin{bmatrix}
a & b \\ c & d
\end{bmatrix}
\begin{bmatrix}
X \\ Y
\end{bmatrix}
&=
\begin{bmatrix}
aX + bY \\ cX + dY
\end{bmatrix} \\
~ \\
[2 \times 2] [2 \times 1] &= [2 \times 1]
\end{aligned}
$$


## Linear models in matrix form | Simple regression

Let's transpose the parameter vector $[\beta_0 ~~ \beta_1]^{\top}$

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
1 & x_1 \\ 1 & x_2 \\ \vdots & \vdots \\ 1 & x_N
\end{bmatrix}
\begin{bmatrix}
\beta_0 \\ \beta_1
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$


## Linear models in matrix form | Simple regression

and check the dimensions

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
1 & x_1 \\ 1 & x_2 \\ \vdots & \vdots \\ 1 & x_N
\end{bmatrix}
\begin{bmatrix}
\beta_0 \\ \beta_1
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix}
$$

$$
\begin{align}
[N \times 1] &= [N \times 2] [2 \times 1] + [N \times 1] \\
&= [N \times 1] + [N \times 1] \\
&= [N \times 1]
\end{align}
$$


## Linear models in matrix form | Simple regression

Lastly, we can write the model in a more compact notation

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
1 & x_1 \\ 1 & x_2 \\ \vdots & \vdots \\ 1 & x_N
\end{bmatrix}
\begin{bmatrix}
\beta_0 \\ \beta_1
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix} \\
\Downarrow \\
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{e}
$$


## Linear models in matrix form | Multiple regression

The matrix form is generalizaable to multiple predictors

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
1 & x_{1,1} & x_{2,1} \\ 1 & x_{1,2} & x_{2,2} \\ \vdots & \vdots \\ 1 & x_{1,N} & x_{2,N} 
\end{bmatrix}
\begin{bmatrix}
\beta_0 \\ \beta_1 \\ \beta_2
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix} \\
\Downarrow \\
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{e}
$$


# QUESTIONS?


## Ordinary least squares

In general, we have something like

$$
DATA = MODEL + ERRORS
$$


## Ordinary least squares

In general, we have something like

$$
DATA = MODEL + ERRORS
$$

<br>

Ideally we have something like

$$
DATA \approx MODEL
$$

and hence

$$
ERRORS \approx 0
$$


## Ordinary least squares

From this it follows that

$$
\text{Var}(DATA) = \text{Var}(MODEL) + \text{Var}(ERRORS)
$$


## Ordinary least squares

From this it follows that

$$
\text{Var}(DATA) = \text{Var}(MODEL) + \text{Var}(ERRORS)
$$

<br>

Our hope is that

$$
\text{Var}(DATA) \approx \text{Var}(MODEL)
$$

and hence

$$
\text{Var}(ERRORS) \approx 0
$$


## Ordinary least squares

Let's say our model for the data is

$$
y_i = f(\text{predictors}_i) + e_i
$$

and our estimate of $y_i$ is given by

$$
\hat{y}_i = f(\text{predictors}_i)
$$


## Ordinary least squares

Let's say our model for the data is

$$
y_i = f(\text{predictors}_i) + e_i
$$

and our estimate of $y_i$ is given by

$$
\hat{y}_i = f(\text{predictors}_i)
$$

The errors (residuals) are then given by

$$
y_i = \hat{y}_i + e_i \\
\Downarrow \\
e_i = y_i - \hat{y}_i \\
$$


## Ordinary least squares

Given our expression for the errors (residuals)

$$
e_i = y_i - \hat{y}_i \\
$$

we want to minimize each of the $e_i$

<br>

Specifically, we want to minimize the sum of their squares

$$
\min \sum_{i = 1}^{N}e_i^2 ~ \Rightarrow ~ \min \sum_{i = 1}^{N}(y_i - \hat{y}_i)^2
$$


## Ordinary least squares

For our linear regression model, we have

$$
\min \sum_{i = 1}^{N}(y_i - \hat{y}_i)^2 \\
\Downarrow \\
\min \sum_{i = 1}^{N} \left(y_i - (\hat{\beta}_0 + \hat{\beta}_1 x_i) \right)^2
$$


## An aside on linear algebra

Recall that matrix multiplication works in a row-by-column manner

$$
\begin{bmatrix}
a & b \\ c & d
\end{bmatrix}
\begin{bmatrix}
X \\ Y
\end{bmatrix}
=
\begin{bmatrix}
aX + bY \\ cX + dY
\end{bmatrix}
$$


## An aside on linear algebra

If $\mathbf{v}$ is an $[n \times 1]$ column vector & $\mathbf{v}^{\top}$ is its $[1 \times n]$ transpose,

multiplying $\mathbf{v}^{\top} \mathbf{v}$ gives the sum of the squared values in $\mathbf{v}$

$$
\begin{aligned}
\begin{bmatrix}
a & b & c
\end{bmatrix}
\begin{bmatrix}
a \\ b \\ c
\end{bmatrix}
&=
\begin{bmatrix}
a^2 + b^2 + c^2
\end{bmatrix} \\
~ \\
[1 \times n] [n \times 1] &= [1 \times 1]
\end{aligned}
$$


## Ordinary least squares

Writing our linear regression model in matrix form, we have

$$
\mathbf{y} = \mathbf{X} \hat{\boldsymbol{\beta}} + \mathbf{e} \\
\Downarrow \\
\mathbf{e} = \mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}}
$$

so the sum of squared errors (SSE) is

$$
\mathbf{e}^{\top} \mathbf{e} = (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}})^{\top} (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}})
$$


## Finding the minimum

For example, at what value of $x$ does this parabola reach its minimum?

$$
y = 2 x^2 - 3 x + 1
$$


## Finding the minimum

For example, at what value of $x$ does this parabola reach its minimum?

$$
y = 2 x^2 - 3 x + 1
$$

<br>

Recall from calculus that we can

1. differentiate $y$ with respect to $x$
2. set the result to 0
3. solve for $x$

## Finding the minimum

For example, at what value of $x$ does this parabola reach its minimum?

$$
y = 2 x^2 - 3 x + 1
$$

<br>

1. differentiate $y$ with respect to $x$: $~~~ \frac{dy}{dx} = 4 x - 3$

2. set the result to 0: $~~~ 4 x - 3 = 0$

3. solve for $x$: $~~~ x = \tfrac{3}{4}$


## Ordinary least squares

We want to minimize the sum of squared errors (SSE)

$$
\begin{align}
\mathbf{e}^{\top} \mathbf{e} &= (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}})^{\top} (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}}) \\
  &= \mathbf{y}^{\top} \mathbf{y} - \mathbf{y}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{y} + \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}}
\end{align}
$$


## Ordinary least squares

We want to minimize the sum of squared errors (SSE)

$$
\begin{align}
\mathbf{e}^{\top} \mathbf{e} &= (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}})^{\top} (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}}) \\
  &= \mathbf{y}^{\top} \mathbf{y} - \mathbf{y}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{y} + \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}}
\end{align}
$$

<br>

which implies we want the derivative of $\mathbf{e}^{\top} \mathbf{e}$, but with respect to what?


## Ordinary least squares

We want to minimize the sum of squared errors (SSE)

$$
\begin{align}
\mathbf{e}^{\top} \mathbf{e} &= (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}})^{\top} (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}}) \\
  &= \mathbf{y}^{\top} \mathbf{y} - \mathbf{y}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{y} + \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}}
\end{align}
$$

<br>

which implies we want the derivative of $\mathbf{e}^{\top} \mathbf{e}$, but with respect to what?

The data ($\mathbf{y}, \mathbf{x}$) are fixed but we can change $\boldsymbol{\beta}$


## Ordinary least squares

We want to minimize the sum of squared errors (SSE)

$$
\begin{align}
\mathbf{e}^{\top} \mathbf{e} &= (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}})^{\top} (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}}) \\
  &= \mathbf{y}^{\top} \mathbf{y} - \mathbf{y}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{y} + \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}}
\end{align}
$$

<br>

We want the partial derivative of $\mathbf{e}^{\top} \mathbf{e}$ with respect to $\hat{\boldsymbol{\beta}}$

$$
\frac{\partial}{\partial \hat{\boldsymbol{\beta}}} \left( \mathbf{y}^{\top} \mathbf{y} - \mathbf{y}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{y} + \hat{\boldsymbol{\beta}}^{\top} \mathbf{X}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}} \right)
$$


## Ordinary least squares

$1.$ differentiate $\mathbf{e}^{\top} \mathbf{e}$ with respect to $\hat{\boldsymbol{\beta}}$ (via several steps) 

$$\frac{\partial (\mathbf{e}^{\top} \mathbf{e})}{\partial \hat{\boldsymbol{\beta}}}  = -2 \mathbf{X}^{\top} \mathbf{y} + 2 \mathbf{X}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}}$$

$2.$ set the result to 0

$$-2 \mathbf{X}^{\top} \mathbf{y} + 2 \mathbf{X}^{\top} \mathbf{X} \hat{\boldsymbol{\beta}} = 0$$

$3.$ solve for $x$

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^{\top} \mathbf{y}$$


## Ordinary least squares

Returning to our estimate of the data, we have

$$
\begin{align}
\hat{\mathbf{y}} &= \mathbf{X} \hat{\boldsymbol{\beta}} \\
  &= \mathbf{X} \left( (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^{\top} \mathbf{y} \right) \\
  &= \underbrace{\mathbf{X} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^{\top}}_{\mathbf{H}} \mathbf{y} \\
  &= \mathbf{H} \mathbf{y}
\end{align}
$$


## Ordinary least squares

```{r model_geom, fig.height=4, fig.width=6, fig.align="center"}
pgram <- data.frame(x = c(0, 3, 10, 7, 0), y = c(0, 3, 3, 0, 0))

## empty plot space
par(mai = rep(0,4), omi = rep(0, 4))
plot(c(0,10), c(0,5), type = "n", xaxt = "n", yaxt = "n",
     bty = "n", xlab = "", ylab = "")
## parallelogram
lines(pgram[,1], pgram[,2])
text(1, 0.5, expression(paste("subspace spanned by ",bold(X))),
     pos = 4, cex = 1.2, col = "darkgray")
## perpendicular
lines(c(6.6,6.6),c(2,2.3), col = "gray")
lines(c(6.6,7),c(2.3,2.3), col = "gray")
## vectors
## data
arrows(3, 2, 7, 4, length = 0.15, angle = 10, lwd = 2, col ="purple")
text(6.5, 4, expression(bold(y)), cex = 1.5)
## yhat
arrows(3, 2, 7, 2, length = 0.15, angle = 10, lwd = 2, col ="blue")
text(6, 1.6, expression(paste(bold(hat(y))," = ",bold(Hy))), cex = 1.5)
## errors
arrows(7, 2, 7, 4, length = 0.15, angle = 10, lwd = 2, col ="red")
text(7.3, 3.5, expression(bold(e)), cex = 1.5)
```

$\mathbf{H}$ is called the "hat matrix" because it maps $\mathbf{y}$ onto $\hat{\mathbf{y}}$ ("y-hat")


## Ordinary least squares

Consider for a moment what it means if

$$
\hat{\mathbf{y}} = \mathbf{X} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^{\top} \mathbf{y} 
$$


## Ordinary least squares

Consider for a moment what it means if

$$
\hat{\mathbf{y}} = \mathbf{X} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^{\top} \mathbf{y} 
$$

**We can estimate the data without any model parameters!**


## Ordinary least squares | Key assumptions

* Model is linear in parameters$^*$

> - Observations $y_i$ are a random sample from the population    

> - The predictor(s) is/are known without measurement error  

> - The predictor(s) is/are independent of the response

> - If 2+ predictors, they are independent of each other

> - Errors are IID: $e_i \sim \text{N}(0, \sigma^2)$; $\text{Cov}(e_i, e_j) = 0$

$^*$parameters are not multiplied or divided by other parameters, nor do they appear as an exponent

## Independent & identically distributed

How do we know if our errors are IID?

* Knowledge of the problem/design  
* Examine residual plots  
* Tests of model fits

<br>

We will discuss this more in later lectures


## Ordinary least squares

What can we say about $\hat{\boldsymbol{\beta}}$ when estimated this way?

1. It's the *maximum likelihood estimate* (MLE)

2. It's the *best linear unbiased estimate* (BLUE)

**NOTE**: these properties only hold if the errors ($e_i$) are *independent and identically distributed* (IID)


## Identifiability

Recall the solution for $\hat{\boldsymbol{\beta}}$ where

$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^{\top} \mathbf{y}
$$

If the quantity $\mathbf{X}^{\top} \mathbf{X}$ is not invertible, then $\hat{\boldsymbol{\beta}}$ is partially unidentifiable. 

This occurs when the columns of $\mathbf{X}$ are not independent<br>(ie, $\mathbf{X}$ is not of "full rank")


## Lack of identifiability | When does it arise?

* analysis of designed experiments (more later)

> - two predictors are perfectly correlated (eg, temperature entered in both degrees F & degrees C)

> - predictors are subsets of one another (eg, counts of trees in 3 categories: DBH $\geq$ 10 cm, DBH $\geq$ 20 cm, DBH $\geq$ 30 cm)

> - number of parameters equals or exceeds the observations  
    $p = n$: model is *saturated*  
    $p \geq n$: model is *supersaturated*








