---
title: "Data and model summaries in R"
author: "<br>Mark Scheuerell"
date: "3 April 2026"
output:
  html_document:
    theme: 
      bootswatch: journal
      primary: "#32006e"
    highlight: textmate
    css: ../lecture_inst.css
    code_download: true
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

This is meant to be a helpful resource for R functions commonly used when working with data, fitting models, and interpreting output.

***

# Data summaries

Here are some functions that are helpful for summarizing data contained in a vector `x`.

`mean(x)`: computes the arithmetic mean (average) of `x`

`sd(x)`: computes the standard deviation of `x`

`var(x)`: computes the variance of `x`

`quantile(x)`: returns estimates of underlying distribution quantiles of `x`

`hist(x)`: creates a histogram of `x`


***

# Random numbers

It's common practice to use some form of randomization when estimating and testing models.

`sample()`: draws a sample of a specified size from a group, with or without replacement

```{r ex_sample}
## make all of this reproducible
set.seed(514)

## a sequence from 1 to 10
xx <- seq(1, 10)

## shuffle the order
sample(xx, size = length(xx), replace = FALSE)

## grab 3 numbers at random
sample(xx, size = 3, replace = FALSE)
```

## Distributional forms

The functions for generating random data from a statistic distribution in R all start with `r` (for "random").

### Continuous values

Continuous values can occur anywhere on the entire real number line from -&infin; to &infin;, but some have a necessary lower bound at zero (e.g., densities, concentrations).

#### Normal

`rnorm()`: draw random samples from a normal distribution

```{r, ex_rnorm}
## 1000 values from a normal distribution with mean = 25 and variance = 5
xx <- rnorm(n = 1000, mean = 25, sd = sqrt(5))

## estimated mean and variance
mean(xx); var(xx)

## histogram of values
hist(xx, main = "",
     breaks = seq(floor(min(xx)), ceiling(max(xx))))
```

<br>

#### Logistic

`rlogis()`: draw random samples from a logistic distribution

#### Uniform

`runif()`: draw random samples from a uniform distribution

<br>

### Discrete values

Discrete values are integers that typically only take on non-negative values (e.g., presence/absence, counts).

#### Bernoulli & Binomial

`rbinom()`: draw random samples from a binomial distribution; can be used for Bernoulli distribution as well

<u>Simulated "coin toss" via Bernoulli</u>

```{r ex_rbern}
## flip a coin 1000 times
xx <- rbinom(n = 1000, size = 1, prob = 0.5)

## count tails (0) and heads (1)
table(xx)
```

<u>Simulated survival</u>

```{r ex_rbinom}
## 1357 tagged smolts; probability of survival is 0.05
xx <- rbinom(n = 1357, size = 1, prob = 0.05)

## number of mortalities (0) and survivors (1)
table(xx)  # p ~= 0.0528
```

#### Poisson

`rpois()`: draw random samples from a Poisson distribution

```{r, ex_rpois}
## 1000 values from a Poisson distribution with mean = variance = 25
xx <- rpois(n = 1000, lambda = 25)

## estimated mean and variance
mean(xx); var(xx)

## histogram of values
hist(xx, main = "",
     breaks = seq(floor(min(xx)), ceiling(max(xx))))
```

<br>

#### Negative binomial

`rnbinom()`: draw random samples from a negative binomial distribution

```{r, ex_rnbinom}
## 1000 values from a negative binomial distribution with mean = 25 and variance = 50
xx <- rnbinom(n = 1000, mu = 25, size = 25)

## estimated mean and variance
mean(xx); var(xx)

## histogram of values
hist(xx, main = "",
     breaks = seq(floor(min(xx)), ceiling(max(xx))))
```


***

# Data in models

Sometimes you want or need to be very explicit about how you want R to treat some aspect of your data. For example, when used inside of a model-fitting function, `a + b` will include both `a` and `b` as predictors in a model, but `I(a + b)` will include the sum of `a` and `b` as a single predictor in a model.

`I()`: change the class of an object to indicate that it should be treated ‘as is’

```{r ex_I, eval = FALSE}
lm(y ~ I(x + z))
```

`offset()`: assume a model predictor has a known coefficient of 1 rather than an unknown coefficient to be estimated

For example, if we counted 6 plants inside a 2 m<sup>2</sup> quadrat, we might express the density of plants as 3 plants per m<sup>2</sup>. As you will see later, models for counts often assume a log-linear relationship, such that 

$$
\log (\text{density}) = \alpha + \beta x \\
\Downarrow \\
\log(\text{number} ~/~ \text{m}^2) = \alpha + \beta x \\
\log(\text{number}) - \log(\text{m}^2) = \alpha + \beta x \\
\log(\text{number}) = \alpha + \beta x + \log(\text{m}^2)\\
$$

We might model this relationship in R with something like

```{r ex_offset, eval = FALSE}
## log-linear regression
lm(log(number) ~ x + offset(log(area)))
```

`scale()`: shifts and/or scales a vector to have a mean of 0 and variance or 1

```{r}
## random variate with mean of 5 and variance of 2
xx <- rnorm(100, mean = 5, sd = sqrt(2))

## scaled values with mean = 0 and variance = 1
zz <- scale(xx)
mean(zz); var(zz)
```

***

# Model summaries

You'll see a variety of functions for fitting different types of linear models in this course. The following functions are useful for extracting information about a fitted model, such as estimated coefficients, standard errors, and p-values.

`anova()`: computes analysis of variance (or deviance) table for a fitted model object

`coefficients()`: extracts model coefficients from objects returned by modeling functions

`residuals()`: extracts model residuals from objects returned by modeling functions

`summary()`: returns a list of summary statistics of the fitted linear model


***

# Monte Carlo

There is a technique known as a _Monte Carlo_ method (or simulation or experiment), whose name comes from the Monte Carlo Casino in Monaco. The idea is to estimate the uncertainty in a stochastic variable or process via repeated sampling. For example, say you wanted to estimate the probability that a coin would come up "heads" when flipped into the air and allowed to land freely. To do so, you could flip the coin 1000 times and after each flip record a 1 if heads or a 0 if tails. At the end count up the number of 1's and divide the sum by 1000. Here's a way we might do that in R.

```{r monte_carlo_ex}
## number of monte carlo experiments
nn <- 1000

## empty vector to record our results
rr <- rep(NA, nn)

## make this repeatable
set.seed(514)

## conduct the experiment
for(i in 1:nn) {
  ## flip the coin; rbinom(1, 1, 0.5) will return a 0 or 1 with probability 0.5
  flip <- rbinom(1, 1, 0.5)
  ## record the result
  rr[i] <- flip
}

## estimate probability of heads
sum(rr) / nn
```



