---
title: "Data and model summaries in R"
author: "Mark Scheuerell"
date: "3 April 2026"
output:
  html_document:
    theme: 
      bootswatch: journal
      primary: "#32006e"
    highlight: textmate
    css: ../lecture_inst.css
    code_download: true
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

This is meant to be a helpful resource when working with data, fitting models, and interpreting output.

# Data summaries

Here are some functions that are helpful for summarizing data.

`mean()`: computes the arithmetic mean (average) of a vector

`sd()`: computes the standard deviation of a vector

`var()`: computes the variance of a vector

`quantile()`: returns estimates of underlying distribution quantiles

`hist()`: creates a histogram of the data in a vector


# Data for models

Sometimes you want or need to be very explicit about how you want R to treat some aspect of your data. For example, when used inside of a model-fitting function, `a + b` will include both `a` and `b` as predictors in a model, but `I(a + b)` will include the sum of `a` and `b` as a single predictor in a model.

`I()`: change the class of an object to indicate that it should be treated ‘as is’

`offset()`: assume a model predictor has a known coefficient of 1 rather than an unknown coefficient to be estimated

`scale()`: shifts and/or scales a vector to have a mean of 0 and variance or 1


# Model summaries

You'll see a variety of functions for fitting different types of linear models in this course. The following functions are useful for extracting information about a fitted model, such as estimated coefficients, standard errors, and p-values.

`anova()`: computes analysis of variance (or deviance) table for a fitted model object

`coefficients()`: extracts model coefficients from objects returned by modeling functions

`residuals()`: extracts model residuals from objects returned by modeling functions

`summary()`: returns a list of summary statistics of the fitted linear model


# Data distributions

It's common practice to use some form of randomization when estimating and testing models.

`sample()`: draws a sample of a specified size from a group, with or without replacement

The functions for generating random data from a statistic distribution in R all start with `r` (for "random").

## Continuous values

`rnorm()`: draw random samples from a normal distribution

`rlogis()`: draw random samples from a logistic distribution

`runif()`: draw random samples from a uniform distribution

## Discrete values

`rbinom()`: draw random samples from a binomial distribution; can be used for Bernoulli distribution as well

`rpois()`: draw random samples from a Poisson distribution

`rnbinom()`: draw random samples from a negative binomial distribution


# Monte Carlo

There is a technique known as a _Monte Carlo_ method (or simulation or experiment), whose name comes from the Monte Carlo Casino in Monaco. The idea is to estimate the uncertainty in a stochastic variable or process via repeated sampling. For example, say you wanted to estimate the probability that a coin would come up "heads" when flipped into the air and allowed to land freely. To do so, you could flip the coin 100 times and after each flip record a 1 if heads or a 0 if tails. At the end count up the number of 1's and divide the sum by 100. Here's a way we might do that in R.

```{r monte_carlo_ex}
## number of monte carlo experiments
nn <- 100

## empty vector to record our results
rr <- rep(NA, nn)

## make this repeatable
set.seed(514)

## conduct the experiment
for(i in 1:nn) {
  ## flip the coin; rbinom(1, 1, 0.5) will return a 0 or 1 with probability 0.5
  flip <- rbinom(1, 1, 0.5)
  ## record the result
  rr[i] <- flip
}

## estimate probability of heads
sum(rr) / nn
```



