---
title: "Homework assignments"
output: html_document
---

```{r child = "_timeline.Rmd", include=FALSE}
```

```{r child = "_list_of_hw.Rmd", include=FALSE}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Students will use **R Markdown** to complete their homework assignments, which will allow you to execute **R** code and combine the output with additional text and equations into a single pdf file. Each homework assignment will have its own Markdown template for students to fill in and submit. Completed homework files should be named

`[lastname]_qerm514_hw_[#].Rmd`

where `[lastname]` is replaced with the student's last name and the `[#]` is replaced with the actual number of the homework assignment (eg, `fisher_qerm514_hw_1.Rmd`). Students should attach their completed homework file to an email message and send it to [qerm514@uw.edu](mailto:qerm514@uw.edu).

### Homework will be assigned each Friday of weeks 1-8 and will be due by 11:59 PM on the following Friday (weeks 2-9).

```{r week_hw, include = FALSE}
## empty list for labs
week_hw <- vector("list", 8)
names(week_hw) <- paste0("L", seq(8))
## hw topics
week_hw$L1 <- markdown
week_hw$L2 <- linear_models
```

```{r topics, include = FALSE}
topics <- lapply(week_hw, function(x) { if(is.list(x)) {c(x$topics)} else {""} })
## topics formatted for html
topics_html <- sapply(lapply(topics, paste, "<br>", collapse = ""), c)
names(topics_html) <- NULL
```

```{r assignment_links, include = FALSE}
# assignment_links <- lapply(week_hw, function(x) {if(is.list(x)) {x$pdf} else {""} })
# ## topics formatted for html
# assignment_links_html <- sapply(lapply(assignment_links, paste, "<br>", collapse = ""), c)
# names(assignment_links_html) <- NULL

assignment_links <- lapply(week_hw, function(x) { if(is.list(x)) {c(x$pdf)} else {""} })
assignment_links_html <- sapply(lapply(assignment_links, function(x) {
  if(length(x)>1) {
    tmp1 <- paste0("[", as.character(fontawesome::fa("file-pdf", height = 20, fill = "#eb6864")), " (", 1:length(x), ")](")
    } else {
      tmp1 <- paste0("[", as.character(fontawesome::fa("file-pdf", height = 20, fill = "#eb6864")), "](")
      }
  tmp2 <- ")"
  tmp <- paste0(tmp1, x, tmp2)
  tmp[x==""] <- ""
  tmp <- paste(tmp, collapse="<br>"); tmp}), c)
```

```{r template_links, include = FALSE}
template_links <- lapply(week_hw, function(x) { if(is.list(x)) {c(x$rmd)} else {""} })
template_links_html <- sapply(lapply(template_links, function(x) {
  if(length(x)>1) {
    tmp1 <- paste0("[", as.character(fontawesome::fa("file-code", height = 20, fill = "#eb6864")), " (", 1:length(x), ")](")
    } else {
      tmp1 <- paste0("[", as.character(fontawesome::fa("file-code", height = 20, fill = "#eb6864")), "](")
      }
  tmp2 <- ")"
  tmp <- paste0(tmp1, x, tmp2)
  tmp[x==""] <- ""
  tmp <- paste(tmp, collapse="<br>"); tmp}), c)
```

```{r data_links, include = FALSE}
data_links <- lapply(week_hw, function(x) { if(is.list(x)) {x$data} else {""} })
## topics formatted for html
# data_links_html <- sapply(lapply(data_links, function(x) {
#   tmp1 <- paste("[",x,"](")
#   tmp2 <- rep(")",length(x))
#   tmp1[x==""] <- ""
#   tmp2[x==""] <- ""
#   tmp <- paste0(tmp1, x, tmp2)
#   tmp <- paste(tmp, collapse = "<br>")
#   tmp}), c)
# names(data_links_html) <- NULL
## topics formatted for html
data_links_html <- sapply(lapply(data_links, function(x) {
  if(length(x)>1) {
    tmp1 <- paste0("[", as.character(fontawesome::fa("table", height = 20, fill = "#eb6864")), " (", 1:length(x), ")](")
    } else {
      tmp1 <- paste0("[", as.character(fontawesome::fa("table", height = 20, fill = "#eb6864")), "](")
      }
  tmp2 <- ")"
  tmp <- paste0(tmp1, x, tmp2)
  tmp[x==""] <- ""
  tmp <- paste(tmp, collapse="<br>"); tmp}), c)
```

```{r key_links, include = FALSE}
key_links <- lapply(week_hw, function(x) { if(is.list(x)) {c(x$key)} else {""} })
key_links_html <- sapply(lapply(key_links, function(x) {
  if(length(x)>1) {
    tmp1 <- paste0("[", as.character(fontawesome::fa("key", height = 20, fill = "#eb6864")), " (", 1:length(x), ")](")
    } else {
      tmp1 <- paste0("[", as.character(fontawesome::fa("key", height = 20, fill = "#eb6864")), "](")
      }
  tmp2 <- ")"
  tmp <- paste(tmp1, x, tmp2)
  tmp[x==""] <- ""
  tmp <- paste(tmp, collapse = " ")
  tmp}), c)
names(key_links_html) <- NULL
```

```{r generate_table, echo=FALSE}
## data frame with dates & topics
syl_txt <- data.frame(
  date = sub("0(\\d)", "\\1", format(fri[1:8], "%d %B")),
  topics = topics_html,
  assignment_links = assignment_links_html,
  template_links = template_links_html,
  data_links = data_links_html,
  key_links = key_links_html
)
colnames(syl_txt) <- c("Date", "Topic(s)", "Assignment", "Template", "Data", "Answer key")
rownames(syl_txt) <- NULL

## generate table
kable(syl_txt, format = "html", caption = " ",
      align = "clccc", escape = FALSE) %>%
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE,
                position = "left") %>%
  column_spec(1, width = "6em") %>%
  column_spec(c(3,4,5), width = "7em") %>%
  column_spec(6, width = "8em")
```

