---
title: "Design Matrices - Part 2"
author: "Sarah Converse"
date: "23 April 2025"
output:
  ioslides_presentation
subtitle: Analysis of Ecological and Environmental Data<br>QERM 514

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(faraway)
library(MASS)
library(plot3D)
```


```{css, echo = FALSE}

.title-slide hgroup h1 {
  font-size: 40px;
}
.title-slide hgroup h2 {
  font-size: 30px;
}
h2 {
  font-size: 40px;
}
slides > slide {
  font-size: 20px;
}
```


```{r, echo = FALSE, results = FALSE}
#dataset 
set.seed(1)

years <- 5
year <- rep(c(1990:1994)-1989,3)
temp <- rep(runif(5,15,29),3)
rainfall <- rep(runif(5,75,125),3)
region <- c(rep(1,5),rep(2,5),rep(3,5))

region.dummy <- matrix(NA,nrow=years*3,ncol=3); region.dummy[,1] <- c(rep(1,5),rep(0,10));region.dummy[,2] <- c(rep(0,5),rep(1,5),rep(0,5)); region.dummy[,3] <- c(rep(0,10),rep(1,5))
b0 <- 110
b1 <- -1.5
b2 <- 0.03
b3 <- 0.01
b4 <- -0.02
b5 <- 4
b6 <- -5
sigma <- 3
creepers <- b0 + b1*year + b2*temp + b3*rainfall + b4*temp*rainfall + b5*region.dummy[,2] + b6*region.dummy[,3] + rnorm(years*3,0,sigma) 

KW <- data.frame(creepers,year,temp,rainfall,region)
KW$creepers <- round(KW$creepers,dig=1)
KW$temp <- round(KW$temp,dig=1)
KW$rainfall <- round(KW$rainfall,dig=1)
KW$region <- as.factor(KW$region)
KW$temp_scale <- scale(KW$temp)[1:15]
KW$rainfall_scale <- scale(KW$rainfall)[1:15]
KW$yrfac <- as.factor(KW$year)

write.csv(KW,"Wallcreeper.csv",row.names=FALSE)
rm(list=ls(all=TRUE))

KW <- read.csv("Wallcreeper.csv")

KW$region <- as.factor(KW$region)
KW$yrfac <- as.factor(KW$yrfac)


``` 


## Objectives for today

* Understand advanced design matrix coding, including design matrices with multiple continuous or categorical covariates, both continuous and categorical covariates, interactions, and nested effects

* Be able to interpret the $\boldsymbol{\beta}$ parameters dependent on the design matrix 


## Design matrices

* We've seen how to code categorical covariates and continuous covariates 

* Now we'll move on to some more complex models including: 

  - multiple continuous or categorical covariates 
 
  - continuous and categorical covariates in the same model
 
  - interactions and nesting 

* Let's return to our wallcreeper data: we have estimates of occupancy for 3 regions in 5 years, and we have temperature and rainfall 

## Design matrices 

Our data: 

$$
\begin{bmatrix}
Region_1,year_1 \\ Region_1,year_2 \\ Region_1,year_3 \\ Region_1,year_4 \\ Region_1,year_5 \\ Region_2,year_1 \\ Region_2,year_2 \\ Region_2,year_3 \\ Region_2,year_4 \\ Region_2,year_5 \\ Region_3,year_1 \\ Region_3,year_2 \\ Region_3,year_3 \\ Region_3,year_4 \\ Region_3,year_5 \\ \end{bmatrix}
\rightarrow
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
$$


## Multiple continuous covariates 

What is the design matrix for this model: `creepers ~ temp + rainfall`? 


## Multiple continuous covariates 

What is the design matrix for this model: `creepers ~ temp + rainfall`? 

```{r, echo = FALSE, results = TRUE}

#run model and look at design matrix 
temp.rain <- lm(creepers ~ temp + rainfall, KW)
model.matrix(temp.rain)[1:15,]
```


## Multiple continuous covariates 

What is the design matrix for this model: `creepers ~ temp + rainfall`? 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
\color{red}{1} & `r KW$temp[1]` & `r KW$rainfall[1]`\\ 
\color{red}{1} & `r KW$temp[2]` & `r KW$rainfall[2]`\\ 
\color{red}{1} & `r KW$temp[3]` & `r KW$rainfall[3]`\\ 
\color{red}{1} & `r KW$temp[4]` & `r KW$rainfall[4]`\\ 
\color{red}{1} & `r KW$temp[5]` & `r KW$rainfall[5]`\\ 
\color{red}{1} & `r KW$temp[1]` & `r KW$rainfall[1]`\\ 
\color{red}{1} & `r KW$temp[2]` & `r KW$rainfall[2]`\\ 
\color{red}{1} & `r KW$temp[3]` & `r KW$rainfall[3]`\\ 
\color{red}{1} & `r KW$temp[4]` & `r KW$rainfall[4]`\\ 
\color{red}{1} & `r KW$temp[5]` & `r KW$rainfall[5]`\\ 
\color{red}{1} & `r KW$temp[1]` & `r KW$rainfall[1]`\\ 
\color{red}{1} & `r KW$temp[2]` & `r KW$rainfall[2]`\\ 
\color{red}{1} & `r KW$temp[3]` & `r KW$rainfall[3]`\\ 
\color{red}{1} & `r KW$temp[4]` & `r KW$rainfall[4]`\\ 
\color{red}{1} & `r KW$temp[5]` & `r KW$rainfall[5]`\\ 
\end{bmatrix}
\begin{bmatrix}
\hat{\beta_0} \\ \hat{\beta_1} \\ \hat{\beta_2}
\end{bmatrix}
$$


## Multiple continuous covariates 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?


## Multiple continuous covariates 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?

   - $\hat{\beta_0}$ is the predicted occupancy when both temperature and rainfall are 0. Note that we don't observe this in our data 
   
   - $\hat{\beta_1}$ is the predicted change in occupancy for each 1 degree change in temperature 

   - $\hat{\beta_2}$ is the predicted change in occupancy for each 1 cm change in rainfall  


## Multiple continuous covariates 

* It's hard to visualize a 3-D relationship, so let's plot our regression for one predictor while holding the other constant 

   - For the red line, rainfall is held at 75 cm 
   
   - For the blue line, rainfall is held at 125 cm
   
```{r, echo = FALSE, results = FALSE}
plot(KW$temp,KW$creepers, ylab = "Percent occupancy", xlab = "Temperature",main = "")

lines(seq(18,27,1),predict(temp.rain,newdata = data.frame(temp = seq(18,27,1), rainfall = rep(75,10))),col="red")


lines(seq(18,27,1),predict(temp.rain,newdata = data.frame(temp = seq(18,27,1), rainfall = rep(125,10))),col="blue")

```


## Multiple continuous covariates 

* What is the design matrix for this model: `creepers ~ scale(temp) + scale(rainfall)`? 


## Multiple continuous covariates 

* What is the design matrix for this model: `creepers ~ scale(temp) + scale(rainfall)`? 

```{r, echo = FALSE, results = TRUE}
#run model and look at design matrix
temp.rain.scale <- lm(creepers ~ temp_scale + rainfall_scale, KW)
model.matrix(temp.rain.scale)[1:15,]
```


## Multiple continuous covariates 

* What is the design matrix for this model: `creepers ~ scale(temp) + scale(rainfall)`? 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
\color{red}{1} & `r scale(KW$temp)[1]` & `r scale(KW$rainfall)[1]`\\ 
\color{red}{1} & `r scale(KW$temp)[2]` & `r scale(KW$rainfall)[2]`\\ 
\color{red}{1} & `r scale(KW$temp)[3]` & `r scale(KW$rainfall)[3]`\\ 
\color{red}{1} & `r scale(KW$temp)[4]` & `r scale(KW$rainfall)[4]`\\ 
\color{red}{1} & `r scale(KW$temp)[5]` & `r scale(KW$rainfall)[5]`\\ 
\color{red}{1} & `r scale(KW$temp)[1]` & `r scale(KW$rainfall)[1]`\\ 
\color{red}{1} & `r scale(KW$temp)[2]` & `r scale(KW$rainfall)[2]`\\ 
\color{red}{1} & `r scale(KW$temp)[3]` & `r scale(KW$rainfall)[3]`\\ 
\color{red}{1} & `r scale(KW$temp)[4]` & `r scale(KW$rainfall)[4]`\\ 
\color{red}{1} & `r scale(KW$temp)[5]` & `r scale(KW$rainfall)[5]`\\ 
\color{red}{1} & `r scale(KW$temp)[1]` & `r scale(KW$rainfall)[1]`\\ 
\color{red}{1} & `r scale(KW$temp)[2]` & `r scale(KW$rainfall)[2]`\\ 
\color{red}{1} & `r scale(KW$temp)[3]` & `r scale(KW$rainfall)[3]`\\ 
\color{red}{1} & `r scale(KW$temp)[4]` & `r scale(KW$rainfall)[4]`\\ 
\color{red}{1} & `r scale(KW$temp)[5]` & `r scale(KW$rainfall)[5]`\\ 
\end{bmatrix}
\begin{bmatrix}
\hat{\beta_0} \\ \hat{\beta_1} \\ \hat{\beta_2}
\end{bmatrix}
$$


## Multiple continuous covariates 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?


## Multiple continuous covariates 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?

   - $\hat{\beta_0}$ is the predicted occupancy when both `scale(temp)` and `scale(rainfall)` are 0. This occurs when temperature and rainfall are at their mean values
   
   - $\hat{\beta_1}$ is the predicted change in occupancy for each 1 SD change in temperature

   - $\hat{\beta_2}$ is the predicted change in occupancy for each 1 SD change in rainfall 
   
   
## Multiple continuous covariates 

* It's hard to visualize a 3-D relationship, so let's plot our regression for one predictor while holding the other constant 

   - For the red line, temperature is held at the mean 
   
   - For the blue line, temperature is held at 1 SD above the mean
   
```{r, echo = FALSE, results = FALSE}
plot(KW$rainfall_scale,KW$creepers, ylab = "Percent occupancy", xlab = "Rainfall",main = "")

lines(seq(-2,2,0.5),predict(temp.rain.scale,newdata = data.frame(temp_scale = rep(0,9), rainfall_scale = seq(-2,2,0.5))),col="red")

lines(seq(-2,2,0.5),predict(temp.rain.scale,newdata = data.frame(temp_scale = rep(1,9), rainfall_scale = seq(-2,2,0.5))),col="blue")

```


## Multiple categorical covariates 

* What is the design matrix for this model: `creepers ~ factor(year) + factor(region)`? 


## Multiple categorical covariates 

* What is the design matrix for this model: `creepers ~ factor(year) + factor(region)`? 

```{r, echo = FALSE, results = TRUE}
year.region <- lm(creepers ~ yrfac + region, KW)
model.matrix(year.region)[1:15,]
```


## Multiple categorical covariates 

* What is the design matrix for this model: `creepers ~ factor(year) + factor(region)`? 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
\color{red}{1} & 0 & 0 & 0 & 0 & 0 & 0\\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 & 0 & 0\\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 & 0 & 0\\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 & 0 & 0\\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} & 0 & 0\\ 
\color{red}{1} & 0 & 0 & 0 & 0 & \color{red}{1} & 0\\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 & \color{red}{1} & 0\\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 & \color{red}{1} & 0\\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 & \color{red}{1} & 0\\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} & \color{red}{1} & 0\\ 
\color{red}{1} & 0 & 0 & 0 & 0 & 0 & \color{red}{1}\\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 & 0 & \color{red}{1}\\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 & 0 & \color{red}{1}\\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 & 0 & \color{red}{1}\\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} & 0 & \color{red}{1}\\ 
\end{bmatrix}
\begin{bmatrix}
\hat{\beta_0} \\ \hat{\beta_1} \\ \hat{\beta_2} \\ \hat{\beta_3} \\ \hat{\beta_4} \\ \hat{\beta_5} \\ \hat{\beta_6} 
\end{bmatrix}
$$


## Multiple categorical covariates 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?


## Multiple categorical covariates 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?

  - $\hat{\beta}_0$ is the prediction for year 1 in region 1 
  
  - $\hat{\beta}_1$, $\hat{\beta}_2$, $\hat{\beta}_3$, and $\hat{\beta}_4$ are the differences between years 2, 3, 4, and 5 (respectively) and year 1. These differences are the same across all regions 
  
  - $\hat{\beta}_5$ and $\hat{\beta}_6$ are the differences between regions 2 and 3 (respectively) and region 1. These differences are the same across all years 


## Multiple categorical covariates 

* Region 1 is red, region 2 is orange, and region 3 is green
   
* The distance from the red to the orange line is the effect of region 2 compared to region 1 (i.e., $\hat{\beta}_5$), and the distance from the red to the green line is the effect of region 3 compared to region 1 (i.e., $\hat{\beta}_6$)

```{r, echo = FALSE, results = FALSE}

plot(KW$year,KW$creepers, ylab = "Percent occupancy", xlab = "Year",main = "")

lines(c(1:5),predict(year.region, newdata = data.frame(yrfac =as.factor(c(1:5)),region = as.factor(rep(1,5)))),col = "red")

lines(c(1:5),predict(year.region, newdata = data.frame(yrfac =as.factor(c(1:5)),region = as.factor(rep(2,5)))),col = "orange")

lines(c(1:5),predict(year.region, newdata = data.frame(yrfac =as.factor(c(1:5)),region = as.factor(rep(3,5)))),col = "green")

```

## Categorical and continuous covariates 

* What is the design matrix for this model: `creepers ~ region + temp`? 


## Categorical and continuous covariates 

* What is the design matrix for this model: `creepers ~ region + temp`? 

```{r, echo = FALSE, results = TRUE}
region.temp <- lm(creepers ~ region + temp, KW)
model.matrix(region.temp)[1:15,]
```


## Categorical and continuous covariates 

* What is the design matrix for this model: `creepers ~ region + temp`? 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
\color{red}{1} & 0 & 0 & `r KW$temp[1]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[2]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[3]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[4]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[5]`\\ 
\color{red}{1} & \color{red}{1} & 0 & `r KW$temp[6]`\\ 
\color{red}{1} & \color{red}{1} & 0 & `r KW$temp[7]`\\ 
\color{red}{1} & \color{red}{1} & 0 & `r KW$temp[8]`\\ 
\color{red}{1} & \color{red}{1} & 0 & `r KW$temp[9]`\\ 
\color{red}{1} & \color{red}{1} & 0 & `r KW$temp[10]`\\ 
\color{red}{1} & 0 & \color{red}{1} & `r KW$temp[11]`\\ 
\color{red}{1} & 0 & \color{red}{1} & `r KW$temp[12]`\\ 
\color{red}{1} & 0 & \color{red}{1} & `r KW$temp[13]`\\ 
\color{red}{1} & 0 & \color{red}{1} & `r KW$temp[14]`\\ 
\color{red}{1} & 0 & \color{red}{1} & `r KW$temp[15]`\\ 
\end{bmatrix}
\begin{bmatrix}
\hat{\beta_0} \\ \hat{\beta_1} \\ \hat{\beta_2} \\ \hat{\beta_3} 
\end{bmatrix}
$$


## Categorical and continuous covariates 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?

  - $\hat{\beta}_0$ is the prediction for region 1 when temperature = 0
  
  - $\hat{\beta}_1$ and $\hat{\beta}_2$ are the differences between regions 2 and 3 (respectively) and region 1. These differences are the same regardless of temperature 

  - $\hat{\beta}_3$ is the predicted change in occupancy for each 1 degree change in temperature. This change is the same over all regions   


## Categorical and continuous covariates 

* Red is for region 1, orange is for region 2, green is for region 3 

* Each region's regression line will cross the y-axis at a different point, but the lines are forced to be parallel in a plot against temperature 

```{r, echo = FALSE, results = FALSE}
plot(KW$temp,KW$creepers, ylab = "Percent occupancy", xlab = "Temperature",main = "")

lines(seq(18,27,1),predict(region.temp,newdata = data.frame(temp = seq(18,27,1), region = as.factor(rep(1,10)))),col="red")

lines(seq(18,27,1),predict(region.temp,newdata = data.frame(temp = seq(18,27,1), region = as.factor(rep(2,10)))),col="orange")

lines(seq(18,27,1),predict(region.temp,newdata = data.frame(temp = seq(18,27,1), region = as.factor(rep(3,10)))),col="green")
```

## Categorical and continuous covariates 

* What is the design matrix for this model: `creepers ~ -1 + region + temp`? 


## Categorical and continuous covariates 

* What is the design matrix for this model: `creepers ~ -1 + region + temp`? 

```{r, echo = FALSE, results = TRUE}
region.temp.nI <- lm(creepers ~ -1 + region + temp, KW)
model.matrix(region.temp.nI)[1:15,]
```


## Categorical and continuous covariates 

* What is the design matrix for this model: `creepers ~ -1 + region + temp`? 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
\color{red}{1} & 0 & 0 & `r KW$temp[1]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[2]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[3]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[4]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[5]`\\ 
0 & \color{red}{1} & 0 & `r KW$temp[6]`\\ 
0 & \color{red}{1} & 0 & `r KW$temp[7]`\\ 
0 & \color{red}{1} & 0 & `r KW$temp[8]`\\ 
0 & \color{red}{1} & 0 & `r KW$temp[9]`\\ 
0 & \color{red}{1} & 0 & `r KW$temp[10]`\\ 
0 & 0 & \color{red}{1} & `r KW$temp[11]`\\ 
0 & 0 & \color{red}{1} & `r KW$temp[12]`\\ 
0 & 0 & \color{red}{1} & `r KW$temp[13]`\\ 
0 & 0 & \color{red}{1} & `r KW$temp[14]`\\ 
0 & 0 & \color{red}{1} & `r KW$temp[15]`\\ 
\end{bmatrix}
\begin{bmatrix}
\hat{\beta_0} \\ \hat{\beta_1} \\ \hat{\beta_2} \\ \hat{\beta_3}
\end{bmatrix}
$$


## Categorical and continuous covariates 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?

  - $\hat{\beta}_0$, $\hat{\beta}_1$, and $\hat{\beta}_2$ are the predictions for regions 1, 2, and 3 (respectively) when temperature = 0
  
  - $\hat{\beta}_4$ is the predicted change in occupancy for each 1 degree change in temperature. This change is the same over all regions  
  
  
## Interactions  

* What is the design matrix for this model: `creepers ~ temp*rainfall`? 


## Interactions 

* What is the design matrix for this model: `creepers ~ temp*rainfall`? 

```{r, echo = FALSE, results = TRUE}
temp.rainfall.intx <- lm(creepers ~ temp*rainfall, KW)
model.matrix(temp.rainfall.intx)[1:15,]
```


## Interactions 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
\color{red}{1} & `r KW$temp[1]` & `r KW$rainfall[1]` & `r KW$temp[1]*KW$rainfall[1]`\\ 
\color{red}{1} & `r KW$temp[2]` & `r KW$rainfall[2]` & `r KW$temp[2]*KW$rainfall[2]`\\ 
\color{red}{1} & `r KW$temp[3]` & `r KW$rainfall[3]` & `r KW$temp[3]*KW$rainfall[3]`\\ 
\color{red}{1} & `r KW$temp[4]` & `r KW$rainfall[4]` & `r KW$temp[4]*KW$rainfall[4]`\\ 
\color{red}{1} & `r KW$temp[5]` & `r KW$rainfall[5]` & `r KW$temp[5]*KW$rainfall[5]`\\ 
\color{red}{1} & `r KW$temp[6]` & `r KW$rainfall[6]` & `r KW$temp[6]*KW$rainfall[6]`\\ 
\color{red}{1} & `r KW$temp[7]` & `r KW$rainfall[7]` & `r KW$temp[7]*KW$rainfall[7]`\\ 
\color{red}{1} & `r KW$temp[8]` & `r KW$rainfall[8]` & `r KW$temp[8]*KW$rainfall[8]`\\ 
\color{red}{1} & `r KW$temp[9]` & `r KW$rainfall[9]` & `r KW$temp[9]*KW$rainfall[9]`\\ 
\color{red}{1} & `r KW$temp[10]` & `r KW$rainfall[10]` & `r KW$temp[10]*KW$rainfall[10]`\\ 
\color{red}{1} & `r KW$temp[11]` & `r KW$rainfall[11]` & `r KW$temp[11]*KW$rainfall[11]`\\ 
\color{red}{1} & `r KW$temp[12]` & `r KW$rainfall[12]` & `r KW$temp[12]*KW$rainfall[12]`\\ 
\color{red}{1} & `r KW$temp[13]` & `r KW$rainfall[13]` & `r KW$temp[13]*KW$rainfall[13]`\\ 
\color{red}{1} & `r KW$temp[14]` & `r KW$rainfall[14]` & `r KW$temp[14]*KW$rainfall[14]`\\ 
\color{red}{1} & `r KW$temp[15]` & `r KW$rainfall[15]` & `r KW$temp[15]*KW$rainfall[15]`\\ 
\end{bmatrix}
\begin{bmatrix}
\hat{\beta_0} \\ \hat{\beta_1} \\ \hat{\beta_2} \\ \hat{\beta_3} \end{bmatrix}
$$


## Interactions 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?

  - $\hat{\beta}_0$ is the prediction when both temperature and rainfall are 0 
  
  - $\hat{\beta}_1$ is the "main effect" of temperature - the predicted change in occupancy for a 1 degree change in temperature **when rainfall = 0** (when rainfall = 0, both $\hat{\beta}_2$ and $\hat{\beta}_3$ drop out)

  - $\hat{\beta}_2$ is the "main effect" of rainfall - the predicted change in occupancy for a 1 cm change in rainfall **when temperature = 0** (when temperature = 0, both $\hat{\beta}_1$ and $\hat{\beta}_3$ drop out)
  
  - $\hat{\beta}_3$ is the "interaction" between temperature and rainfall - the change in the effect of temperature and rainfall as the other changes

```{r, echo = FALSE, results = FALSE}

p1 <- predict(temp.rainfall.intx,newdata = data.frame(temp = 20, rainfall = 100))

p2 <- predict(temp.rainfall.intx,newdata = data.frame(temp = 20, rainfall = 101))
```


## Interactions 

* Red is for rainfall = 75 cm; orange = 85 cm; yellow = 95 cm; green = 105 cm

```{r, echo = FALSE, results = FALSE}
plot(KW$temp,KW$creepers, ylab = "Percent occupancy", xlab = "Temperature",main = "")

lines(seq(18,27,1),predict(temp.rainfall.intx,newdata = data.frame(temp = seq(18,27,1), rainfall = rep(75,10))),col="red")

lines(seq(18,27,1),predict(temp.rainfall.intx,newdata = data.frame(temp = seq(18,27,1), rainfall = rep(85,10))),col="orange")

lines(seq(18,27,1),predict(temp.rainfall.intx,newdata = data.frame(temp = seq(18,27,1), rainfall = rep(95,10))),col="yellow")

lines(seq(18,27,1),predict(temp.rainfall.intx,newdata = data.frame(temp = seq(18,27,1), rainfall = rep(105,10))),col="green")
```


## Nested effects 

* What is the design matrix for this model: `creepers ~ temp:region`? 


## Nested effects 

* What is the design matrix for this model: `creepers ~ temp:region`? 

```{r, echo = FALSE, results = TRUE}
temp.region.nest <- lm(creepers ~ temp:region, KW)
model.matrix(temp.region.nest)[1:15,]
```


## Nested effects 

* What is the design matrix for this model: `creepers ~ temp:region`? 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
\color{red}{1} & `r KW$temp[1]` & 0 & 0 \\ 
\color{red}{1} & `r KW$temp[2]` & 0 & 0 \\ 
\color{red}{1} & `r KW$temp[3]` & 0 & 0 \\ 
\color{red}{1} & `r KW$temp[4]` & 0 & 0 \\ 
\color{red}{1} & `r KW$temp[5]` & 0 & 0 \\ 
\color{red}{1} & 0 & `r KW$temp[1]` & 0 \\ 
\color{red}{1} & 0 & `r KW$temp[2]` & 0 \\ 
\color{red}{1} & 0 & `r KW$temp[3]` & 0 \\ 
\color{red}{1} & 0 & `r KW$temp[4]` & 0 \\ 
\color{red}{1} & 0 & `r KW$temp[5]` & 0 \\ 
\color{red}{1} & 0 & 0 & `r KW$temp[1]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[2]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[3]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[4]`\\ 
\color{red}{1} & 0 & 0 & `r KW$temp[5]`\\ 
\end{bmatrix}
\begin{bmatrix}
\hat{\beta_0} \\ \hat{\beta_1} \\ \hat{\beta_2} \\ \hat{\beta_3} 
\end{bmatrix}
$$


## Nested effects 

* What is the interpretation of each of the $\boldsymbol{\hat{\beta}}$?

  - $\hat{\beta}_0$ is the prediction when temperature is 0. This is the same in every region  
  
  - $\hat{\beta}_1$ , $\hat{\beta}_2$ , $\hat{\beta}_3$ is the predicted change in occupancy for a 1 degree change in temperature in regions 1, 2, and 3, respectively

## Nested effects 

* Let's plot this relationship

   - red is region 1, orange is region 2, green is region 3 
   
   - these lines will intersect at the y-axis 
   
```{r, echo = FALSE, results = FALSE}
plot(KW$temp,KW$creepers, ylab = "Percent occupancy", xlab = "Temperature",main = "")

lines(seq(18,27,1),predict(temp.region.nest,newdata = data.frame(temp = seq(18,27,1), region = as.factor(rep(1,10)))),col="red")

lines(seq(18,27,1),predict(temp.region.nest,newdata = data.frame(temp = seq(18,27,1), region = as.factor(rep(2,10)))),col="orange")

lines(seq(18,27,1),predict(temp.region.nest,newdata = data.frame(temp = seq(18,27,1), region = as.factor(rep(3,10)))),col="green")

```


## Summary 

Our model is defined by a design matrix $X$

To build and understand our model (e.g., interpret the vector $\hat{\beta}$) we must understand $X$  

There is tremendous flexibility in building models to allow us to make the inference we want to make




