---
title: "Intro to mixed effects models"
subtitle: "Analysis of Ecological and Environmental Data<br>QERM 514"
author: "Mark Scheuerell"
date: "4 May 2026"
output:
  ioslides_presentation:
    wide: true
    css: ../lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Goals for today

* Understand types of random effects structures

> - Understand how random effects are estimated

> - Understand restricted maximum likelihood

> - Understand approaches to make inference from mixed models


## General types of models

| Errors | Single random process | Multiple random processes |
|:-------|:--------------:|:------------------:|
| Normal Errors | Linear Model (LM) | Linear Mixed Model (LMM) |
| Multiple Forms of Errors | Generalized Linear Model (GLM) | Generalized Linear Mixed Model (GLMM) |


## Multiple random processes | Linear mixed models 

We are now moving into linear models with *multiple* random processes

> * the first still describes our model *errors* (as before)

> * the other(s) describe the *random effect(s)* of some factor(s)


## What do we mean by random effects? 

Let's start with *fixed effects*, which are what we've been estimating


## Fixed effects

We use fixed effects for 

* Continuous predictors 


## Fixed effects

We use fixed effects for 

* Continuous predictors 

* Categorical predictors (factors) when 

   * we are interested in the specific levels (eg, "N", "P", "N & P", "Control")  


## Fixed effects 

We use fixed effects for: 

* Continuous predictors 

* Categorical predictors (factors) when 

   * we are interested in the specific levels (eg, "N", "P", "N & P", "Control")  

   * the factors don't come from a larger population (eg, "Age 1", "Age 2", "Age 3+")


## Fixed effects 

We use fixed effects for: 

* Continuous predictors 

* Categorical predictors (factors) when

   * we are interested in the specific levels (eg, "N", "P", "N & P", "Control")  

   * the factors don't come from a larger population (eg, "Age 1", "Age 2", "Age 3+")
   
   * we only have a few levels (eg, "Plot 1", "Plot 2", "Plot 3")


## Fixed vs. random effects? 

We use random effects for

* factor levels sampled from a distribution (eg, we randomly chose 12 study plots)


## Fixed vs. random effects? 

We use random effects for

* factor levels sampled from a distribution (eg, we randomly chose 12 study plots)

* correlated errors when data are

   - nested (eg, plants within plots & multiple plots sampled)


## Fixed vs. random effects? 

We use random effects for

* factor levels sampled from a distribution (eg, we randomly chose 12 study plots)

* correlated errors when data are  

   - nested (eg, plants within plots & multiple plots sampled)

   - time series (eg, productivity of captive birds over many years) 


## Fixed vs. random effects? 

We use random effects for

* factor levels sampled from a distribution (eg, we randomly chose 12 study plots)

* correlated errors when data are  

   - nested (eg, plants within plots & multiple plots sampled)

   - time series (eg, productivity of captive birds over many years) 

   - spatial (eg, multiple ponds under study) 


## Model for means

Imagine we are interested in modeling the mass of fish measured in several different lakes

We have 3 hypotheses about the variation in fish sizes

1. differences in mass are due mostly to individual fish with no differences among lakes


## Model for means

Imagine we are interested in modeling the mass of fish measured in several different lakes

We have 3 hypotheses about the variation in fish sizes

1. differences in mass are due mostly to individual fish with no differences among lakes

2. differences in mass are due mostly to *specific* factors that differ among lakes


## Model for means

Imagine we are interested in modeling the mass of fish measured in several different lakes

We have 3 hypotheses about the variation in fish sizes

1. differences in mass are due mostly to individual fish with no differences among lakes

2. differences in mass are due mostly to *specific* factors that differ among lakes

3. differences in mass are due mostly to *general* factors that are shared among lakes


## Model for means

Our first model simply treats all of the fish $i$ in the different lakes $j$ as one large group

$$
y_{ij} = \mu + \epsilon_{ij} \\
\epsilon_{ij} \sim \text{N}(0, \sigma^2_{\epsilon}) \\
$$

where $\mu$ is the mean mass of fish across *all* lakes & our primary interest is the size of $\sigma_{\epsilon}^2$


## Model for means

In essence, we are *pooling* all of fish from the different lakes together so we can drop the $j$ subscript

$$
y_{ij} = \mu + \epsilon_{ij} \\
\epsilon_{ij} \sim \text{N}(0, \sigma^2_{\epsilon}) \\
\Downarrow \\
y_{i} = \mu + \epsilon_{i} \\
\epsilon_{i} \sim \text{N}(0, \sigma^2_{\epsilon})
$$



## Model for means

Our second model separates all of the fish $i$ into groups based on the *specific* lake $j$ from which they were caught

$$
y_{ij} = \mu + \alpha_j + \epsilon_{ij} \\
\epsilon_{ij} \sim \text{N}(0, \sigma^2_{\epsilon})
$$

<br>

where $\alpha_j$ is the *specific* effect of lake $j$


## Model for means

Here there is *no pooling* of fish from different lakes and the $j$ subscript tells us about a *specific* lake

$$
y_{ij} = \mu + \alpha_j + \epsilon_{ij} \\
\epsilon_{ij} \sim \text{N}(0, \sigma^2_{\epsilon})
$$



## Model for means

Our last model treats differences in fish mass among lakes as similar to one another (correlated)

$$
y_{ij} = \mu + \alpha_j + \epsilon_{ij} \\
\epsilon_{ij} \sim \text{N}(0, \sigma^2_{\epsilon}) \\
\alpha_j \sim \text{N}(0, \sigma^2_{\alpha})
$$

<br>

where $\alpha_j$ is the effect of lake $j$ as though it were *randomly* chosen


## Model for means

The degree of correlation among lakes $(\rho)$ is determined by the relative sizes of $\sigma^2_{\alpha}$ and $\sigma^2_{\epsilon}$ 

$$
y_{ij} = \mu + \alpha_j + \epsilon_{ij} \\
\epsilon_{ij} \sim \text{N}(0, \sigma^2_{\epsilon}) \\
\alpha_j \sim \text{N}(0, \sigma^2_{\alpha}) \\
\Downarrow \\
\rho = \frac{\sigma^2_{\alpha}}{\sigma^2_{\alpha} + \sigma^2_{\epsilon}}
$$


## Model for means

Here we could say that the lakes are *partially pooled* together by formally addressing correlations among lakes 

$$
y_{ij} = \mu + \alpha_j + \epsilon_{ij} \\
\epsilon_{ij} \sim \text{N}(0, \sigma^2_{\epsilon}) \\
\alpha_j \sim \text{N}(0, \sigma^2_{\alpha})
$$

with

$$
\rho = \frac{\sigma^2_{\alpha}}{\sigma^2_{\alpha} + \sigma^2_{\epsilon}}
$$


## Model for means

```{r dist_of_means, fig.width=5, fig.height=4.5, fig.align='center'}
## normal pdf + constant
kk <- 1
npdf <- function(x, k = kk) {
  k + dnorm(x)
}
## set plot area
par(mai = c(0.1, 0.1, 0.1, 0.1),
    omi = c(0.6, 0, 0, 0))
## grand mean
plot(c(-8, 8), c(2, 2), type = "l", lwd = 2, ylim = c(0,2.4),
     xaxt = "n", yaxt = "n", bty = "n",
     xlab = "", ylab = "")
abline(v = 0, lty = "dashed")
points(0, 2, pch = 16, col = "blue", cex = 1.5)
text(-8, 2.4, "no effects", pos = 4, cex = 1.2)
text(8, 2.4, "complete pooling", pos = 2, cex = 1.2)
## random means
curve(npdf, -8, 8, lwd = 2, add = TRUE, col = "gray",
      xaxt = "n", yaxt = "n", bty = "n",
      xlab = expression(alpha[j]), ylab = "")
lines(c(-8, 8), c(kk, kk), lwd = 2,
     xaxt = "n", yaxt = "n", bty = "n")
points(rnorm(7), rep(kk, 7), pch = 16, col = "blue", cex = 1.5)
text(-8, 1.4, "random effects", pos = 4, cex = 1.2)
text(8, 1.4, "partial pooling", pos = 2, cex = 1.2)
## fixed means
lines(c(-8, 8), c(0, 0), lwd = 2,
     xaxt = "n", yaxt = "n", bty = "n")
points(runif(7, -8, 8), rep(0, 7), pch = 16, col = "blue", cex = 1.5)
text(-8, 0.4, "fixed effects", pos = 4, cex = 1.2)
text(8, 0.4, "no pooling", pos = 2, cex = 1.2)
## x-axis
mtext("0", 1, line = 0.5, cex = 1.3, adj = c(1,1)/2)
mtext(expression(alpha[italic(j)]), 1, line = 2.5, cex = 1.5)
```


## Fish mass across lakes

```{r sim_ancova, fig.width=6, fig.height=4.5, fig.align='center'}
set.seed(514)
## number of lakes
nl <- 12
## number of fish per lake
nf <- sample(seq(2, 24), nl)
## lake ID's
ID <- vector("list", nl)
for(i in 1:nl) {
  ID[[i]] <- rep(i, nf[[i]])
}
IDs <- unlist(ID)
## total number of fish
tf <- sum(sapply(nf, sum))
## fish lengths
fl <- sapply(nf, rnorm, mean = 4, sd = 0.8)
lfl <- unlist(fl)
## y-intercepts
b0 <- rnorm(nl, -4, 0.4)
## slopes
b1 <- rnorm(nl, 1.5, 0.4)
## errors
ee <- sapply(nf, rnorm, mean = 0, sd = 0.3)
## fish mass
fm <- vector("list", nl)
for(i in 1:nl) {
  fm[[i]] <- b0[i] + b1[i] * fl[[i]] + ee[[i]]
}
lfm <- unlist(fm)

## colors
clr <- viridis::plasma(nl, alpha = 0.7, begin = 0.2, end = 0.8)

## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass by lake
plot(rep(1, nf[[1]]), fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = c(1, nl), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "Lake ID",
     xaxt = "n", cex.lab = 1.2)
axis(1, seq(nl))
for(i in 2:nl) {
  points(rep(i, nf[[i]]), fm[[i]], pch = 16,  col = clr[i])
}
```


## Fish mass across lakes

Simple model with complete pooling

```{r complete_pool, echo = TRUE}
## log of fish mass (lfm) as grand mean
m1 <- lm(lfm ~ 1)
```


## Fish mass across lakes

```{r complete_pool_plot, fig.width=6, fig.height=4.5, fig.align='center'}
## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass by lake
plot(rep(1, nf[[1]]), fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = c(1, nl), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "Lake ID",
     xaxt = "n", cex.lab = 1.2)
axis(1, seq(nl))
for(i in 2:nl) {
  points(rep(i, nf[[i]]), fm[[i]], pch = 16,  col = clr[i])
}
abline(h = coef(m1), lty = "dashed")
```


## Fish mass across lakes

Fixed effects model with no pooling across lakes

```{r no_pool, echo = TRUE, results='hold'}
## log of fish mass (lfm) with lake-level means
m2 <- lm(lfm ~ 1 + as.factor(IDs))
```


## Fish mass across lakes

```{r no_pool_plot, fig.width=6, fig.height=4.5, fig.align='center'}
## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass by lake
plot(rep(1, nf[[1]]), fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = c(1, nl), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "Lake ID",
     xaxt = "n", cex.lab = 1.2)
abline(h = coef(m1), lty = "dashed", col = "gray")
points(1, mean(fm[[1]]), pch = "-", cex = 4)
axis(1, seq(nl))
for(i in 2:nl) {
  points(rep(i, nf[[i]]), fm[[i]], pch = 16, col = clr[i])
  points(i, mean(fm[[i]]), pch = "-", cex = 4)
}
```


## Fish mass across lakes

Random effects model with partial pooling across lakes

```{r partial_pool, echo = TRUE, results='hold', message=FALSE}
## load lme4 package
library(lme4)
## log of fish mass (lfm) with lake-level effects
m3 <- lmer(lfm ~ 1 + (1|IDs))
```


## Fish mass across lakes

```{r partial_pool_plot, fig.width=6, fig.height=4.5, fig.align='center'}
## random intercepts
m3_coef <- unlist(coef(m3))

## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass by lake
plot(rep(1, nf[[1]]), fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = c(1, nl), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "Lake ID",
     xaxt = "n", cex.lab = 1.2)
abline(h = coef(m1), lty = "dashed", col = "gray")
points(1, mean(fm[[1]]), pch = "-", cex = 4,  col = "gray")
points(1, m3_coef[1], pch = "-", cex = 4)
axis(1, seq(nl))
for(i in 2:nl) {
  points(rep(i, nf[[i]]), fm[[i]], pch = 16,  col = clr[i])
  points(i, mean(fm[[i]]), pch = "-", cex = 4,  col = "gray")
  points(i, m3_coef[i], pch = "-", cex = 4)
}
```


## Shrinkage of group means

In fixed effects models, the group means are

$$
\alpha_j = \bar{y} - \mu
$$

<br>

In random effects models, the group means "shrink" towards the mean

$$
\alpha_j = (\bar{y} - \mu) \left( \frac{\sigma^2_{\alpha}}{\sigma^2_{\alpha} + \sigma^2_{\epsilon}} \right)
$$


# QUESTIONS?


## Fish mass across lakes

Let's return to our model for fish mass across different lakes

Now we want to include the effect of fish length as well


## Fish mass versus length

```{r length_mass, fig.width=5, fig.height=4.5, fig.align='center'}
## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass vs length
plot(fl[[1]], fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = range(unlist(fl)), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "log(fish length)",
     cex.lab = 1.2)
for(i in 2:nl) {
  points(fl[[i]], fm[[i]], pch = 16,  col = clr[i])
}
```


## A global regression model

Fish mass as a function of its length (no lake effects)

$$
y_{i} = \underbrace{\beta_0 + \beta_1 x_{i}}_{\text{fixed}} + \epsilon_{ij} 
$$

$\epsilon_{ij} \sim \text{N}(0,\sigma_\epsilon)$


## A global regression model

Fish mass as a function of its length (no lake effects)

```{r fit_regr, echo = TRUE}
## fit global regression model
a1 <- lm (lfm ~ lfl)
```


## A global regression model

```{r plot_regr_fit, fig.width=5, fig.height=4.5, fig.align='center'}
## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass vs length
plot(fl[[1]], fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = range(unlist(fl)), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "log(fish length)",
     cex.lab = 1.2)
for(i in 2:nl) {
  points(fl[[i]], fm[[i]], pch = 16,  col = clr[i])
}
## regr line
abline(coef(a1))
```


## A global regression model

```{r plot_regr_resid, fig.width=5, fig.height=4.5, fig.align='center'}
e1 <- f1 <- fm
## get lake-specific resids
for(i in 1:nl){
  f1[[i]] <- predict(a1, newdata = data.frame(lfl = fl[[i]]))
  e1[[i]] <- fm[[i]] - f1[[i]] 
}

## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass vs length
plot(f1[[1]], e1[[1]], pch = 16, las = 1, col = clr[1],
     xlim = range(unlist(f1)), ylim = range(unlist(e1)),
     ylab = "Residuals", xlab = "Fitted values",
     cex.lab = 1.2)
for(i in 2:nl) {
  points(f1[[i]], e1[[i]], pch = 16,  col = clr[i])
}
## zero line
abline(h = 0, lty = "dashed")
```


## A global regression model

```{r plot_regr_resid_ID, fig.width=5, fig.height=4.5, fig.align='center'}
## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass vs length
plot(rep(1, nf[[1]]), e1[[1]], pch = 16, las = 1, col = clr[1],
     xlim = c(1,12), ylim = range(unlist(e1)),
     ylab = "Residuals", xlab = "Lake ID",
     cex.lab = 1.2, xaxt = "n")
for(i in 2:nl) {
  points(rep(i, nf[[i]]), e1[[i]], pch = 16,  col = clr[i])
}
## zero line
abline(h = 0, lty = "dashed")
axis(1, seq(nl))
```


## Unique regression models

Fish mass as a function of its length for *each* lake

$$
y_{ij} = \underbrace{\beta_{0j} + \beta_{1j} x_{ij}}_{\text{fixed}} + \epsilon_{ij}
$$

$\epsilon_{ij} \sim \text{N}(0,\sigma_\epsilon)$


## Unique regression models

Fish mass as a function of its length for *each* lake

```{r fit_unique_regr, echo = TRUE}
## matrix for coefs
cf <- matrix(NA, nl, 2)
## fit regression unique to each lake
for(i in 1:nl) {
  cf[i,] <- coef(lm(fm[[i]] ~ fl[[i]]))
}
```


## Unique regression models

```{r plot_unique_regr, fig.width=5, fig.height=4.5, fig.align='center'}
## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass vs length
plot(fl[[1]], fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = range(unlist(fl)), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "log(fish length)",
     cex.lab = 1.2)
## regr line
abline(a = cf[1,1], b = cf[1,2], col = clr[i])
for(i in 2:nl) {
  points(fl[[i]], fm[[i]], pch = 16,  col = clr[i])
  ## regr line
  abline(a = cf[i,1], b = cf[i,2], col = clr[i])
}
```


## A linear mixed model

Fish mass as a function of its length for a *random* lake

$$
y_{ij} = \underbrace{\beta_{0j} + \beta_1 x_{ij}}_{\text{fixed}} + \underbrace{\alpha_{j}}_\text{random} + \epsilon_{ij} 
$$

$\epsilon_{ij} \sim \text{N}(0,\sigma_\epsilon)$

$\alpha_{j} \sim \text{N}(0,\sigma_\alpha)$


## A linear model (ANCOVA)

Fish mass as a function of its length and *random* lake

```{r fit_ancova_rdm_int, echo = TRUE}
## fit ANCOVA with fixed factor for length & rdm factor for lake
a2 <- lmer(lfm ~ lfl + (1|IDs))
```


## Fish mass versus length

```{r plot_ancova_rdm_fit, fig.width=5, fig.height=4.5, fig.align='center'}
## get coefs
ints <- coef(a2)$IDs[,1]

## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass vs length
plot(fl[[1]], fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = range(unlist(fl)), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "log(fish length)",
     cex.lab = 1.2)
## regr line
abline(a = ints[1], b = coef(a2)$IDs[1,2], col = clr[i])
for(i in 2:nl) {
  points(fl[[i]], fm[[i]], pch = 16,  col = clr[i])
  ## regr line
  abline(a = ints[i], b = coef(a2)$IDs[1,2], col = clr[i])
}
```


## A random effects model

Fish mass as a function of its length for a *random* fish *and* lake

$$
y_{ij} = (\beta_{0j} + \alpha_{j}) + (\beta_{1j} + \delta_j) x_{ij} + \epsilon_{ij} \\
y_{ij} = \underbrace{\beta_{0j} + \beta_{1j} x_{ij}}_\text{fixed} + \underbrace{\alpha_{j} + \delta_j x_{ij}}_\text{random} + \epsilon_{ij}
$$

$\epsilon_{ij} \sim \text{N}(0,\sigma_\epsilon)$

$\alpha_{j} \sim \text{N}(0,\sigma_\alpha)$

$\delta_{j} \sim \text{N}(0,\sigma_\delta)$


## A random effects model

Fish mass as a function of its length for a *random* fish *and* lake

```{r fit_ancova_rdm_both, echo = TRUE}
## fit ANCOVA with random effects for length & lake
a3 <- lmer(lfm ~ lfl + (lfl|IDs))
```


## A random effects model

```{r plot_ancova_rdm_both, fig.width=5, fig.height=4.5, fig.align='center'}
## get coefs
ints <- coef(a3)$IDs[,1]
slps <- coef(a3)$IDs[,2]

## set plot area
par(mai = c(0.9, 0.9, 0.1, 0.1),
    omi = c(0, 0, 0, 0))

## plot fish mass vs length
plot(fl[[1]], fm[[1]], pch = 16, las = 1, col = clr[1],
     xlim = range(unlist(fl)), ylim = range(unlist(fm)),
     ylab = "log(fish mass)", xlab = "log(fish length)",
     cex.lab = 1.2)
## regr line
abline(a = ints[1], b = slps[1], col = clr[i])
for(i in 2:nl) {
  points(fl[[i]], fm[[i]], pch = 16,  col = clr[i])
  ## regr line
  abline(a = ints[i], b = slps[i], col = clr[i])
}
```


## Model diagnostics

```{r re_diagnostics, fig.width=8, fig.height=3.5, fig.align='center'}
## set plot area
par(mai = c(0.9, 0.9, 0.4, 0.1),
    omi = c(0, 0, 0, 0),
    mfrow = c(1,3), cex.lab = 1.5)

## qq resids
qqnorm(residuals(a3), main = "QQ plot (residuals)", pch = 16)
qqline(residuals(a3))
## qq RE's
qqnorm(unlist(ranef(a3)), main = "QQ plot (RE's)", pch = 16)
qqline(unlist(ranef(a3)))
## resids vs fitted
plot(fitted(a3), residuals(a3), pch = 16,
     xlab = "Fitted", ylab = "Residuals",
     main = "Residuals vs fitted")
abline(h=0, lty = "dashed")
```


## Summary

* Think hard about your question and data

    - Are there groups or levels?
    
    - Are there temporal or spatial dimensions?   

<br>

* Decide what random effects make sense

