---
title: "Intro to mixed effects models"
subtitle: "Analysis of Ecological and Environmental Data<br>QERM 514"
author: "Mark Scheuerell"
date: "1 May 2020"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Reminder

__Your project plan is due today__


## Goals for today


## Forms of linear models
 
```{r model_diagram_lmm, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("black", "black", "black", "blue"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("black", "black", "black", "blue"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Why use linear mixed models?

* Ecological data are often messy, complex, and incomplete

* Data are often grouped by location, species, etc

* May have multiple samples from the same individual

* Often small sample sizes for some locations, species, etc


## Some terminology

**fixed factor** = *qualitative* predictor (e.g. sex)

**fixed effect** = *quantitative* change ("slope")

**random factor** = *qualitative* predictor whose levels are randomly
sampled from a population (eg, age)

**random effect** = *quantitative* change whose levels are randomly
sampled from a population


## A linear model (ANCOVA)

Fish mass as a function of its length and *specific* location

$$
y_{i,j} = \underbrace{\alpha + \beta x_{i,j} + \delta_{j}}_{\text{fixed}} + \underbrace{\epsilon_{i,j}}_{\text{random}} 
$$

$y_i$ is the log(mass) for fish *i* in location $j$

$x_i$ is the log(length) for fish *i* in location $j$

$\delta_j$ is the mean log(mass) of fish in location $j$

$\epsilon_{i,j} \sim \text{N}(0,\sigma_\epsilon)$


## A linear mixed model

Fish mass as a function of its length and *general* location

$$
y_{i,j} = \underbrace{\alpha + \beta x_{i,j}}_{\text{fixed}} + \underbrace{\delta_{j} + \epsilon_{i,j}}_{\text{random}} 
$$

$y_i$ is the log(mass) for fish *i* in location $j$

$x_i$ is the log(length) for fish *i* in location $j$

$\delta_j$ is the mean log(mass) of fish in location $j$

$\epsilon_{i,j} \sim \text{N}(0,\sigma_\epsilon) ~ \text{and} ~ \delta_{j} \sim \text{N}(0,\sigma_\delta)$


