---
title: "Course overview"
subtitle: "Analysis of Ecological and Environmental Data<br>QERM 514"
author: "Mark Scheuerell"
date: "30 March 2020"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## {.flexbox .vcenter .bigger}

<font size="10">
Who am I? 
</font>


## {.flexbox .vcenter .bigger}

<font size="10">
Who are you? 
</font>


## {.flexbox .vcenter .bigger}

<font size="10">
What is this course about? 
</font>


## Two major goals in ecology:

<font size="6">
1. Infer process from pattern   
</font>


## {.flexbox .vcenter}

<font size="10">
Process $\overset{?}{\Rightarrow}$ Pattern
</font>


## {.flexbox .vcenter}

<font size="10">
Pattern $= f($Process$)$
</font>


## {.flexbox .vcenter}

<font size="10">
Data $= f($Process$)$
</font>


## {.flexbox .vcenter}

<font size="10">
Data = Process + Noise
</font>


## Ecological data often have lots of noise

```{r info_noise_barchart}
dat <- matrix(c(20, 80, 90, 10), 2, 2)
par(mai = c(0.5, 1, 0.5, 0.5))
bp <- barplot(dat, las = 1, col = c("blue", "gray"), border = NA, axes = FALSE,
        ylab = "", names.arg = c("Ecology", "Finance"), cex.names = 1.7)
text(0.7, 10, adj = c(0.5, 0.5), "Process", col = "white", cex = 1.5)
text(0.7, 60, adj = c(0.5, 0.5), "Noise", col = "black", cex = 1.5)
text(1.9, 45, adj = c(0.5, 0.5), "Process", col = "white", cex = 1.5)
text(1.9, 95, adj = c(0.5, 0.5), "Noise", col = "black", cex = 1.5)
```

## {.flexbox .vcenter}

<font size="10">
Our challenge is to separate<br>the signal from the noise
</font>


## Two major goals in ecology:

<font size="6">
1. Infer process from pattern   
</font>

<font size="6">
2. **Make predictions**
</font>


## Ecological forecasting

<font size="6">
How will [some future scenario] affect<br>[some ecosystem service]?
</font>


## General approach

### Question $\rightarrow$ Data $\rightarrow$ Model $\rightarrow$ Inference $\rightarrow$ Prediction


## Learning objectives for the course

By the end of the quarter, students should be able to:

* Identify an appropriate statistical model based on the data and specific question

* Understand the assumptions behind a chosen statistical model

* Use **R** to fit a variety of linear models to data

* Evaluate data support for various models and select the most parsimonious model among them

* Use **R Markdown** to combine text, equations, code, tables, and figures into reports


## {.flexbox .vcenter .bigger}

<font size="10">
[__Course information__](https://qerm514.github.io/website)
</font>




