---
title: "Course overview"
subtitle: "Analysis of Ecological and Environmental Data<br>QERM 514"
author: "Mark Scheuerell"
date: "30 March 2026"
output:
  ioslides_presentation:
    widescreen: true
    css: ../lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center')
```

## {.flexbox .vcenter .bigger}

<font size="10">
Who am I? 
</font>


## My role as course instructor

> - Help you learn the material

> - Help you learn how to ask for help

> - Be a future resource


## {.flexbox .vcenter .bigger}

<font size="10">
Who are you? 
</font>


## Introduce yourself

1. Your degree program (MS, PhD)

2. Your school/department

3. Your area of study (a phrase or short sentence)


## {.flexbox .vcenter .bigger}

<font size="10">
What is this course about? 
</font>


## Two major goals in ecology:

<font size="6">
1. Infer process from pattern   
</font>


## {.flexbox .vcenter}

<font size="10">
Process $\overset{?}{\Rightarrow}$ Pattern
</font>


## {.flexbox .vcenter}

<font size="10">
Pattern $= f($Process$)$
</font>


## {.flexbox .vcenter}

<font size="10">
Data $= f($Process$)$
</font>


## {.flexbox .vcenter}

<font size="10">
Data = Process + Noise
</font>


## Ecological data often have lots of noise

```{r info_noise_barchart, fig.alt = "Barplot comparing the proportions of signal and noise in contrasting datasets between ecology and finance."}
dat <- matrix(c(20, 80, 90, 10), 2, 2)
par(mai = c(0.5, 1, 0.5, 0.5))
bp <- barplot(dat, las = 1, col = c("blue", "gray"), border = NA, axes = FALSE,
        ylab = "", names.arg = c("Ecology", "Finance"), cex.names = 1.7)
text(0.7, 10, adj = c(0.5, 0.5), "Process", col = "white", cex = 1.5)
text(0.7, 60, adj = c(0.5, 0.5), "Noise", col = "black", cex = 1.5)
text(1.9, 45, adj = c(0.5, 0.5), "Process", col = "white", cex = 1.5)
text(1.9, 95, adj = c(0.5, 0.5), "Noise", col = "black", cex = 1.5)
```

## {.flexbox .vcenter .hcenter}

<font size="10">
Our challenge is to separate<br>the signal from the noise
</font>


## Two major goals in ecology:

<font size="6">
1. Infer process from pattern   
</font>

<font size="6">
2. **Make predictions**
</font>


## Ecological forecasting

<font size="6">
How will [some future scenario] affect [some ecosystem service]?
</font>


## General approach

### Question $\rightarrow$ Data $\rightarrow$ Model $\rightarrow$ Inference $\rightarrow$ Prediction


## Forms of linear models

| Errors | Single random process | Multiple random processes |
|:-------|:---------------------:|:-------------------------:|
| Normal | Linear Model (LM)     | Linear Mixed Model (LMM)  |
| Non-normal | Generalized Linear Model (GLM) | Generalized Linear Mixed Model (GLMM) |


## Forms of linear models
 
```{r model_diagram, fig.height = 5, fig.width = 7, fig.alt = "Circle and arrow diagram showing the relationtships between general and generalized linear models with and without random effects."}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5)
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Learning objectives for the course

By the end of the quarter, students should be able to:

* Identify an appropriate statistical model based on the data and specific question

* Understand the assumptions behind a chosen statistical model

* Use **R** to fit a variety of linear models to data

* Evaluate data support for various models and select the most parsimonious model among them

* Use **R Markdown** to combine text, equations, code, tables, and figures into reports


# Software

## R Software | An environment for statistical computing and graphics

> - It's FREE!

> - Enormous developer & support communities

> - [Learn more here](https://www.r-project.org/about.html)


## R Software

* Most of R's current capabilities come from *packages*

> - We will use several different packages for fitting statistical models (eg, `nlme`, `lme4`)


## RStudio | An integrated development environment (IDE) for R


## &nbsp; {data-background=RStudio_desktop.png data-background-size=cover}


## RStudio highlights

* Syntax highlighting, code completion, smart indentation  
* Easily manage multiple working directories using projects  
* Interactive debugger to diagnose and fix errors quickly  
* Extensive package development tools  
* Built-in tools & GUIs for git & Markdown


## RStudio benefits

[Cheat sheets](https://rstudio.com/resources/cheatsheets/)

[Webinars, links, etc](https://resources.rstudio.com/)


## &nbsp; {data-background=RStudio_cheatsheet.png data-background-size=cover}


## R Markdown

Notebook interface to weave together text, equations, and code into nicely formatted output

Allows you to create and document fully reproducible workflows

Ideal framework for homework assignments!


## R Markdown

Supports dozens of static and dynamic output formats

* HTML  
* PDF  
* MS Word  
* HTML5 slides  
* Books  
* Shiny apps  
* Scientific articles  
* websites

[Gallery here](https://rmarkdown.rstudio.com/gallery.html)


## {.flexbox .vcenter .bigger}

<font size="10">
[__Course information__](https://qerm514.github.io/website)
</font>




