---
title: "Course Synthesis"
author: "Sarah Converse"
date: "6 June 2025"
output:
  ioslides_presentation
subtitle: Analysis of Ecological and Environmental Data<br>QERM 514

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(faraway)
library(MASS)
library(knitr)
library(VGAM)
library(pscl)
library(extraDistr)

```



```{css, echo = FALSE}

.title-slide hgroup h1 {
  font-size: 40px;
}
.title-slide hgroup h2 {
  font-size: 30px;
}
h2 {
  font-size: 40px;
}
slides > slide {
  font-size: 18px;
}
```


## Objectives for today

* Reflect on all we've covered this quarter 

* Hear some statistical "life lessons" to keep in mind as you move forward


## Course learning objectives


* By the end of the quarter, you should be able to:

   > + Identify an appropriate statistical model based on data and an analytical question

   > + Understand the assumptions behind a chosen statistical model and evaluate whether assumptions are reasonably met in a given situation

   > + Conceptualize, implement, and interpret statistical analyses based on the models covered in this course

   > + Fit models in a variety of R libraries and work with model outputs
   
   
## Why this course? 

> * There are patterns in our observations of the world that tell us about how the world works: pattern tells us about process

> * To better understand how systems work, we need to separate the noise from the pattern 

> * In ecology and environmental science, there is a relatively large amount of noise, but also meaningful pattern

> * We need to be able to make statements about what we’ve learned (and what we are uncertain about)


## What were our general topics? 

* Moving from a problem to a model for a broad class of models: 

   + linear models
   
   + mixed models
   
   + generalized linear models
   
   + generalized linear mixed models

* Moving from a model to inference 


## Model types 

| Errors | Single random process | Multiple random processes |
|:-------|:--------------:|:------------------:|
| Normal Errors | Linear Model (LM) | Linear Mixed Model (LMM) |
| Multiple Forms of Errors | Generalized Linear Model (GLM) | Generalized Linear Mixed Model (GLMM) |


## Linear models

```{r echo = FALSE, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("black", "blue", "black", "black"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("black", "blue", "black", "black"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Partitioning deviations

```{r partition_SS, echo = FALSE, fig.width = 8, fig.height = 3.5, fig.align="center"}
## set random seed for reproducibility
set.seed(514)

## sample size
nn <- 30
## some values for the predictor x
xx <- sort(runif(nn, 0, 20))
## intercept
beta_0 <- 1
## slope
beta_1 <- 0.5
## random errors (mean = 0, SD = 1)
epsilon <- rnorm(nn)
## observed data
yy <- beta_0 + beta_1*xx + epsilon

## fit a regression model
mm <- lm(yy ~ xx)
## get predicted values
yhat <- predict(mm)

## set plot area
par(mfrow = c(1,3),
    mai = c(0.5,0.3,0.5,0.2),
    omi = c(0, 0, 0, 0),
    cex = 0.7)

## SSTO - plot y and mean(y)
plot(xx, yy, type = "n",
     bty = "l", xaxt = "n", yaxt = "n",
     xlab = "",
     ylab = "")
segments(x0 = xx, y0 = yy, y1 = mean(yhat), col = "gray")
abline(h = mean(yy), lty = "dashed")
points(xx, yy, pch = 19)
mtext(expression(Total: ~ italic(y[i])-bar(italic(y))), side = 3, line = 1)
mtext(expression(italic(x)), 1, line=0.2, cex=1.1, at=max(xx))
mtext(expression(italic(y)), 2, line=0.3, cex=1.1, at=max(yy), las=1)
mtext(expression(italic(bar(y))), 2, line=0.3, cex=1.1, at=mean(yy), las=1)

## SSR - plot yhat and mean(yhat)
plot(xx, yy, type = "n",
     bty = "l", xaxt = "n", yaxt = "n",
     xlab = "",
     ylab = "")
segments(x0 = xx, y0 = yhat, y1 = mean(yhat), col = "gray")
points(xx, yhat, pch = 21, bg = "white")
abline(a = coef(mm)[1], b = coef(mm)[2])
abline(h = mean(yy), lty = "dashed")
points(xx, yhat, pch = 21, bg = "white")
mtext(expression(Model: ~ italic(hat(y)[i])-bar(italic(y))), side = 3, line = 1)
mtext(expression(italic(x)), 1, line=0.2, cex=1.1, at=max(xx))
mtext(expression(italic(y)), 2, line=0.3, cex=1.1, at=max(yy), las=1)
mtext(expression(italic(bar(y))), 2, line=0.3, cex=1.1, at=mean(yy), las=1)
mtext(expression(italic(hat(y))), 4, line=0.2, cex=1.1, at=max(yy), las=1)

## SSE - plot observed and yhat
plot(xx, yy, type = "n",
     bty = "l", xaxt = "n", yaxt = "n",
     xlab = "",
     ylab = "")
segments(x0 = xx, y0 = yhat, y1 = yy, col = "gray")
points(xx, yy, pch = 19)
abline(a = coef(mm)[1], b = coef(mm)[2])
mtext(expression(Error: ~ italic(y[i])-italic(hat(y)[i])), side = 3, line = 1)
mtext(expression(italic(x)), 1, line=0.3, cex=1.1, at=max(xx))
mtext(expression(italic(y)), 2, line=0.3, cex=1.1, at=max(yy), las=1)
mtext(expression(italic(hat(y))), 4, line=0.2, cex=1.1, at=max(yy), las=1)
```


## Partitioning sums-of-squares

The sums-of-squares have the same additive property as the deviations

$$
\underbrace{\sum (y_i - \bar{y})^2}_{SSTO} = \underbrace{\sum (\hat{y}_i - \bar{y})^2}_{SSR} + \underbrace{\sum (y_i - \hat{y}_i)^2}_{SSE}
$$


## Linear models in matrix form

$$
y_i = \beta_0 + \beta_1 x_{1,i} + \beta_2 x_{2,i} + e_i \\
\Downarrow \\
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_N
\end{bmatrix}
= 
\begin{bmatrix}
1 & x_{1,1} & x_{2,1} \\ 1 & x_{1,2} & x_{2,2} \\ \vdots & \vdots \\ 1 & x_{1,N} & x_{2,N} 
\end{bmatrix}
\begin{bmatrix}
\beta_0 \\ \beta_1 \\ \beta_2
\end{bmatrix}
+
\begin{bmatrix}
e_1 \\ e_2 \\ \vdots \\ e_N
\end{bmatrix} \\
\Downarrow \\
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{e}
$$


## Ordinary least squares

Rewriting our model, we have

$$
\mathbf{y} = \mathbf{X} \hat{\boldsymbol{\beta}} + \mathbf{e} \\
\Downarrow \\
\mathbf{e} = \mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}}
$$

so the sum of squared errors is

$$
\mathbf{e}^{\top} \mathbf{e} = (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}})^{\top} (\mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}})
$$


## Ordinary least squares

Minimizing the sum of squared errors leads to

$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^{\top} \mathbf{y} \\
\Downarrow \\
\hat{\mathbf{y}} = \mathbf{X} \hat{\boldsymbol{\beta}}
$$


## Inferential tools 

Tests 

$$
F = \frac{ \left( SSE_{\theta} - SSE_{\Theta} \right)  / (k_{\Theta} - k_{\theta})}{ SSE_{\Theta}  / (n - k_{\Theta})} \sim F_{k_{\Theta} - k_{\theta}, n - k_{\Theta}}
$$

<br> 

Confidence intervals 

$$
\hat{\mathbf{y}}^* \pm ~ t^{(\alpha / 2)}_{df} \sigma \sqrt{ {\mathbf{X}^*}^{\top} (\mathbf{X}^{\top} \mathbf{X})^{-1} \mathbf{X}^* } 
$$


## Diagnostics

```{r diagnostics, echo = FALSE, fig.width = 8, fig.height = 3.5, fig.align="center"}
xx <- rnorm(40)
yy <- 1 + 2*xx + rnorm(40)
mm <- lm(yy ~ xx)

## set plot area
par(mfrow = c(1,3),
    mai = c(0.9,0.4,0.5,0.1),
    omi = c(0, 0.5, 0, 0),
    cex = 1)

## e vs y_hat
plot(residuals(mm), fitted(mm), pch = 16, las = 1, xpd = NA,
     cex.lab = 1.5, xlab = expression(hat(italic(y))), ylab = "Residuals",
     main = "", cex.main = 1)
## QQ plot
qqnorm(residuals(mm), pch = 16, las = 1,
     cex.lab = 1.5,
     main = "", cex.main = 1)
qqline(residuals(mm))
## ACF
acf(residuals(mm),
     cex.lab = 1.5, xlab = expression(italic(x)), ylab = expression(y),
     main = "", cex.main = 1)
```


## Unusual observations

```{r outliers, echo = FALSE, fig.width = 8, fig.height = 3.5, fig.align="center"}
xr <- round(1:10 + rnorm(10, 0, 0.2), 1)
testdata <- data.frame(x = xr,
                       y = xr + rnorm(10))
mm <- lm(y ~ x, testdata)

p1 <- c(5.5,12)
m1 <- lm(y ~ x, rbind(testdata, p1))

p2 <- c(17,17)
m2 <- lm(y ~ x, rbind(testdata, p2))

p3 <- c(17,5.1)
m3 <- lm(y ~ x, rbind(testdata, p3))

## set plot area
par(mfrow = c(1,3),
    mai = c(0.9,0.4,0.5,0.1),
    omi = c(0, 0.5, 0, 0),
    cex = 1)

## plot examples
plot(y ~ x, rbind(testdata, p1), pch = 16, las = 1, xpd = NA,
     cex.lab = 1.5, xlab = expression(italic(x)), ylab = expression(y),
     main = "No leverage or influence", cex.main = 1)
points(5.5, 12, pch = 16, cex = 1.5, col ="red")
abline(mm)
abline(m1, lty=2, col ="red")
plot(y ~ x, rbind(testdata, p2), pch = 16, las = 1,
     cex.lab = 1.5, xlab = expression(italic(x)), ylab = expression(y),
     main = "Leverage but no influence", cex.main = 1)
points(17, 17, pch = 16, cex = 1.5, col ="red")
abline(mm)
abline(m2, lty=2, col ="red")
plot(y ~ x, rbind(testdata, p3), pch = 16, las = 1,
     cex.lab = 1.5, xlab = expression(italic(x)), ylab = expression(y),
     main = "Leverage and influence", cex.main = 1)
points(17, 5.1, pch = 16, cex = 1.5, col ="red")
abline(mm)
abline(m3, lty=2, col ="red")
```


## Bias-variance trade-off

```{r var_bias_trade-off, fig.align='center', fig.height=4.5, fig.width=5.5}
## create dummy variance and bias^2
a1 <- 1
b1 <- 0.2
a2 <- 3
b2 <- -0.25
xx <- seq(0,100)/10
## variance
prec <- exp(a1 + b1*xx)
## bias
bias <- exp(a2 + b2*xx)
## total
both <- prec + bias
mn <- both == min(both)
## plot them
par(mai = c(0, 0, 0, 0),
    omi = c(0.5,0.5,0.05,0.05),
    cex = 1.2)
## combination
plot(xx, both, type = "l", col="black", lwd=2, ylim=c(0,max(both)*1.05),
     xlab="", ylab="", xaxt="n", yaxt="n", xaxs="r", yaxs="i")
segments(x0 = xx[mn], y0 = 0, y1 = both[mn], lty = "dashed", col="dimgray")
## variance
lines(xx, prec, col="blue", lwd=2)
## bias
lines(xx, bias, col="red", lwd=2)
## labels
mtext("Model complexity", side = 1, line = 1, cex = 1.5)
mtext("Error", side = 2, line = 1, cex = 1.5)
text(10, max(both)*0.95, "Total", col="black", pos=2)
text(0, 5, "Variance", col="blue", pos=4)
text(10, 3, expression(Bias^2), col="red", pos=2)
text(xx[mn], both[mn]*1.05, "Optimum", pos=3, col="dimgray")
```


## Model selection 

In sample 

* Null hypothesis testing (backwards, forwards, and stepwise selection)

* Regularization (AIC, QAIC, BIC)

Cross validation

* K-fold 

* Leave-one-out 

Out of sample 


## Linear mixed models

```{r model_diagram_lmm, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("black", "black", "black", "blue"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("black", "black", "black", "blue"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## What are random effects? 

We use random effects for: 

* Factor levels sampled from a distribution (e.g., we randomly chose 12 study plots)

* Accounting for correlation in errors because data are:  

   - Nested (e.g., fish within schools, multiple schools measured)

   - Longitudinal (e.g., productivity of captive birds over many years) 

   - Spatial (e.g., multiple grassland plots under study) 


## Linear mixed model

We can extend the linear model to include both fixed and random effects

$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{Z} \boldsymbol{\alpha} + \mathbf{e} \\
~ \\
\mathbf{e} \sim \text{MVN}(\mathbf{0}, \sigma^2 \mathbf{I}) \\
~ \\
\boldsymbol{\alpha} \sim \text{MVN}(\mathbf{0}, \sigma^2 \mathbf{D})
$$


## Model selection for mixed models

1. Fit a global model with respect to fixed effects

2. Keep the fixed effects constant and search for random effects (use bootstrap LRT)

3. Keep random effects as is and fit different fixed effects (use AIC)


## Generalized linear models

```{r model_diagram_glm, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("blue", "black", "black", "black"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("blue", "black", "black", "black"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Generalized linear models

*Three components*

1. Distribution of the data $y \sim f_{\theta}(y)$

2. Link function $g(\eta)$

3. Linear predictor $\eta = \mathbf{X} \boldsymbol{\beta}$


## Logistic regression for binary response

1. Distribution of the data: $y \sim \text{Bernoulli}(p)$

2. Link function: $\text{logit}(p) = \log \left( \frac{p}{1 - p} \right) = \eta$

3. Linear predictor: $\eta = \mathbf{X} \boldsymbol{\beta}$


## Logistic regression for proportions

1. Distribution of the data: $y \sim \text{Binomial}(N, p)$

2. Link function: $\text{logit}(p) = \log \left( \frac{p}{1 - p} \right) = \eta$

3. Linear predictor: $\eta = \mathbf{X} \boldsymbol{\beta}$


## Overdispersion

The variance is larger than would be expected for a binomial random variable

Overdispersion generally arises in 2 ways related to IID errors

1) heterogeneity in $p$ across trials 

2) trials are not independent


## Overdispersion relative to binomial?  

| Model                   | Pros | Cons |
|:------------------------|:-----|:-----|
| binomial                | Easy | Underestimates variance |
| binomial with variance inflation | Easy; estimate of variance | Ad hoc |
| quasi-binomial                   | Easy; estimate of variance | No distribution for inference |
| beta-binomial                    | Solid theoretical foundation | Somewhat hard to implement |


## Poisson regression

1. Distribution of the data: $y \sim \text{Poisson}(\lambda)$

2. Link function: $\text{log}(\lambda) = \eta$

3. Linear predictor: $\eta = \mathbf{X} \boldsymbol{\beta}$


## Overdispersed relative to Poisson?

1. Distribution of the data: $y \sim \text{Negative binomial}(\lambda,\phi)$

2. Link function: $\text{log}(\lambda) = \eta$

3. Linear predictor: $\eta = \mathbf{X} \boldsymbol{\beta}$


## Zero-truncated data

Zero-truncated data cannot take a value of 0

Although somewhat rare in ecological studies, examples include

* number of seconds a whale is at the surface before diving

* herd size in elk

We can fit a zero-truncated Poisson or negative binomial 


## Zeros in ecological data

Lots of count data are *zero-inflated*

* The data contain more zeros than would be expected under a Poisson or negative binomial distribution

There are 2 general approaches for dealing with zero-inflated data

1) Zero-altered ("hurdle") models

2) Zero-inflated ("mixture") models


## Sources of zeros and approaches

| Source | Reason | Over-dispersion | Zero inflation | Approach |
|:-------|:-------|:--------------:|:--------------:|:--------:|
| Random | Sampling variability | No | No | Poisson |
|   |   | Yes | No | Neg binomial |
| Structural | Outside count process | No | Yes | ZAP or ZIP |
|   |   | Yes | Yes | ZANB or ZINB |


## Generalized linear mixed models

```{r model_diagram_glmm, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("black", "black", "blue", "black"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("black", "black", "blue", "black"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Generalized linear mixed model

GLMMs combine the flexibility of non-normal distributions (GLMs) with the ability to address correlations among observations and nested data structures (LMMs)

<br> 

<u>Good news</u>

* these extensions follow similar methods to GLMs and LMMs

<u>Bad news</u>

* these models are on the frontier of statistical research

* existing documentation is rather technical

* multiple approaches for fitting models; some with different results


## Generalized linear mixed model

Just like GLMs, GLMMs have three components:

1. Distribution of the data $f(y; \theta)$

2. Link function $g(\eta)$

3. Linear predictor $\eta$

For GLMMs, our linear predictor also includes random effects

$$
\eta = \mathbf{X} \boldsymbol{\beta} + \mathbf{Z} \boldsymbol{\alpha}
$$


## Common packages for fitting GLMMs 

* `MASS:glmmPQL` - offers PQL

* `lme4::glmer` - offers LaPlace approximation or GHQ, most widely used so lots of help available, negative binomial with `lme4::glmer.nb` 

* `glmmTMB::glmmTMB` - offers LaPlace approximation only, allows for a wider variety of distributions than `lme4`, and offers both zero-inflation and some simple correlation structures 


## Generalized additive models 

An additive model has a predictor that is a sum of smoothed (and possibly unsmoothed) terms: 

$y_i = \beta_0 + \Sigma_j s_j(x_{j,i}) + \epsilon_j$

<br> 

With smoothed terms, we have to decide how much smoothing is best 


## Measuring wiggliness 

First, consider the squared second derivative:

$\int_{\mathbb{R}}[f'']^2dx$

* This is the rate of change in the slope - more wiggliness translates to a higher rate of change in the slope 

<br> 

We use an estimate of this to penalize our likelihood: 

$\mathcal{L}_p(\beta)=log(\mathcal{L}) - \lambda\beta^TS\beta$

where $\lambda$ is a smoothness parameter indicating how much we penalize wiggliness - we can select an appropriate penalty using methods such as REML or cross-validation


## LMs $\rightarrow$ GLMMs

```{r echo = FALSE, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("black", "black", "black", "black"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("black", "black", "black", "black"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Statistics life-lessons 

Now you need to go out and apply what you’ve learned – to your own data, to colleagues’ data, and in your role as a practicing scientist (and therefore critic of scientific method)  

This is what I hope you’ll keep in mind as you go: 


## Lesson number 1 

A good analysis begins with a good question

A good question is... 

* An interesting question: it’s a question that relates to something that is relevant and important to basic or applied science

Thinking hard about the possible answers to our question is key: this is where good hypotheses come from  


## Platt (1964) 

_Strong inference is just the ...method of inductive inference that goes back to Francis Bacon. The steps are familiar...and are practiced, off and on, by every scientist. The difference comes in their systematic application:_ 

_1) Devising alternative hypotheses_ 

_2) Devising a crucial experiment (or several of them), with alternative possible outcomes, each of which will, as nearly as possible, exclude one or more of the hypotheses_ 

_3) Carrying out the experiment so as to get a clean result_ 

_1’)  Recycling the procedure_


## Lesson number 2 

A good analysis doesn’t have to be a complicated analysis 

* A clean, elegant analysis is always best when you can achieve it. Don’t make the analysis harder or more complicated than it needs to be 

* Keep the explanation clean and elegant too - the goal should be clarity and reproducibility 

## Lesson number 3 

Stay close to the data 

* Ideally we don’t want to model functions of the data – we want to model the data itself; that way we can capture the information, as well as the error, as truly and cleanly as possible 


## Lesson number 4

Know thy model! 

Both you and your reader should be quite clear on (at least) three things: 

* What your model (set) is

* Why it is your model (set)

* How inference extracted from your model (set) will lend insights into your question 


## Lesson number 5 

Assumptions matter 

* The strength of our inference is conditional on the degree to which we meet our assumptions. We need to be very clear, and allow our reader to be very clear, on our assumptions

* Sometimes you’ll pay a price for this (e.g., population indices) 

* But don’t give in! Fight the good fight!


## Lesson number 6 

Be your own (best) critic

* Look for the holes. Confess them. If you can’t fix them, that doesn’t mean there is no value in the analysis. Sometimes the best you can do is be honest about where the holes are. Maybe someone else down the road will figure out a way to fix those holes. But it’s your responsibility to point them out  


## Sagan (1996)

_And the second reason that I'm worried about [lack of scientific literacy in the population] is that science is more than a body of knowledge. It's a way of thinking. A way of skeptically interrogating the universe with a fine understanding of human fallibility...._ 

_If we are not able to ask skeptical questions, to interrogate those who tell us that something is true, to be skeptical of those in authority, then we're up for grabs for the next charlatan – political or religious – who comes ambling along. It's a thing that Jefferson laid great stress on. It wasn't enough, he said, to enshrine some rights in a Constitution or a Bill of Rights. The people had to be educated, and they had to practice their skepticism and their education. Otherwise we don't run the government—the government runs us._


## Lesson number 7

You will never know enough about analyzing data

Therefore you have to keep learning




