---
title: "Design Matrices - Part 1"
author: "Sarah Converse"
date: "21 April 2025"
output:
  ioslides_presentation
subtitle: Analysis of Ecological and Environmental Data<br>QERM 514

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(faraway)
library(MASS)
```

```{css, echo = FALSE}

.title-slide hgroup h1 {
  font-size: 40px;
}
.title-slide hgroup h2 {
  font-size: 30px;
}
h2 {
  font-size: 40px;
}
slides > slide {
  font-size: 20px;
}
```


## Objectives for today

* Review the matrix representation of linear models 

* Understand the significance of the design matrix in building and interpreting models 

* Understand the construction of design matrices for categorical and continuous predictors


## Linear models 

* Recall models for the height of tree $i$

* We've said that a model is defined by $\boldsymbol{X}$ 
  - What does $\boldsymbol{X}$ look like in each of these models?
  - Is there just one possible $\boldsymbol{X}$ for each of these models? 
  
$$
height_i = \beta_0 + \beta_1[site_i] + \epsilon_i\\ 			
height_i = \beta_0 + \beta_1[site_i] + \beta_2[species_i] + \epsilon_i\\ 
height_i = \beta_0 + \beta_1*age_i + \epsilon_i\\         
height_i = \beta_0 + \beta_1*age_i + \beta_2*rainfall_i + \epsilon_i\\
height_i = \beta_0 + \beta_1*age_i + \beta_2[species_i] + \epsilon_i\\
height_i = \beta_0 + \beta_1*age_i * \beta_2[species_i] + \epsilon_i\\
$$


## Linear models in matrix form

* To build and understand our model (e.g., interpret the vector $\boldsymbol{\beta}$) we must understand $\boldsymbol{X}$  

$$
\begin{bmatrix}
y_1 \\ y_2 \\ \vdots \\ y_n
\end{bmatrix}
= 
\begin{bmatrix}
1 & x_{11} & \cdots & x_{1p}\\ 1 & x_{21} & \cdots &  x_{2p}\\ \vdots & \vdots & \ddots & \vdots\\ 1 & x_{n1} & \cdots & x_{np}
\end{bmatrix}
\begin{bmatrix}
\beta_0 \\ \beta_1 \\ \vdots \\ \beta_p
\end{bmatrix}
+
\begin{bmatrix}
\epsilon_1 \\ \epsilon_2 \\ \vdots \\ \epsilon_n
\end{bmatrix}
$$


## Design matrices 

$\boldsymbol{X}$ is our design matrix. Recall that it is: 

* $n$ rows – one for each sample 

* $p$ columns – one for each parameter 

$$
\begin{bmatrix}
1 & x_{11} & \cdots & x_{1p}\\ 1 & x_{21} & \cdots &  x_{2p}\\ \vdots & \vdots & \ddots & \vdots\\ 1 & x_{n1} & \cdots & x_{np}
\end{bmatrix}
$$


## An example 

* We have the % of plots occupied by Kery's wallcreeper for each year from **1990 - 1994**, across 3 regions, with temperature and rainfall data for each year 

```{r, echo = FALSE, results = TRUE}
set.seed(16)

years <- 5
year <- rep(c(1990:1994)-1989,3)
temp <- rep(runif(5,15,29),3)
rainfall <- rep(runif(5,30,50),3)
region <- c(rep(1,5),rep(2,5),rep(3,5))

region.dummy <- matrix(NA,nrow=years*3,ncol=3); region.dummy[,1] <- c(rep(1,5),rep(0,10));region.dummy[,2] <- c(rep(0,5),rep(1,5),rep(0,5)); region.dummy[,3] <- c(rep(0,10),rep(1,5))
b0 <- 35 
b1 <- -1.5
b2 <- -0.2
b3 <- -0.01
b4 <- 1
b5 <- -2
sigma <- 3
creepers <- b0 + b1*year + b2*temp + b3*rainfall + b4*region.dummy[,2] + b5*region.dummy[,3] + rnorm(years*3,0,sigma) 

KW <- data.frame(creepers,year,temp,rainfall,region)
KW$creepers <- round(KW$creepers,dig=1)
KW$temp <- round(KW$temp,dig=1)
KW$rainfall <- round(KW$rainfall,dig=1)
KW$region <- as.factor(KW$region)

print(KW)

``` 


## An example

* Let's build a basic model 

```{r, echo = TRUE, results = TRUE}

sumary(trend.mod <- lm(creepers ~ year, KW))

```

## An example{.smaller} 

* What does our design matrix look like? 

```{r, echo = TRUE, results = TRUE}

model.matrix(trend.mod)[1:15,]
```

## An example

* To better understand the design matrix, let's think about how prediction works: 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
1 & 1 \\ 
1 & 2 \\
1 & 3 \\
1 & 4 \\
1 & 5 \\
1 & 1 \\
1 & 2 \\
1 & 3 \\
1 & 4 \\
1 & 5 \\
1 & 1 \\
1 & 2 \\
1 & 3 \\
1 & 4 \\
1 & 5 \\
\end{bmatrix}
\begin{bmatrix}
\hat{\beta}_0 \\ \hat{\beta}_1 
\end{bmatrix}
$$

## Prediction

$\hat{y}_1$= 1 * `r round(coef(trend.mod)[1],dig=3)` +
1 * `r round(coef(trend.mod)[2],dig=3)`\

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(trend.mod))),ceiling(max(KW$creepers,predict(trend.mod))))
plot(KW$year[1],predict(trend.mod)[1],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Prediction

$\hat{y}_1$= 1 * `r round(coef(trend.mod)[1],dig=3)` +
1 * `r round(coef(trend.mod)[2],dig=3)`\
$\hat{y}_2$= 1 * `r round(coef(trend.mod)[1],dig=3)` +
2 * `r round(coef(trend.mod)[2],dig=3)`

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(trend.mod))),ceiling(max(KW$creepers,predict(trend.mod))))
plot(KW$year[1:2],predict(trend.mod)[1:2],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Prediction

$\hat{y}_1$= 1 * `r round(coef(trend.mod)[1],dig=3)` +
1 * `r round(coef(trend.mod)[2],dig=3)`\
$\hat{y}_2$= 1 * `r round(coef(trend.mod)[1],dig=3)` +
2 * `r round(coef(trend.mod)[2],dig=3)`\
$\cdots$\
$\hat{y}_5$= 1 * `r round(coef(trend.mod)[1],dig=3)` +
5 * `r round(coef(trend.mod)[2],dig=3)`

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(trend.mod))),ceiling(max(KW$creepers,predict(trend.mod))))
plot(KW$year[1:5],predict(trend.mod)[1:5],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Design matrices{.smaller}

* What does the intercept do for us? 

```{r, echo = TRUE, results = TRUE}

sumary(trend.mod2 <- lm(creepers ~ -1 + year, KW))
```


## Omitting the intercept 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
1 \\ 
2 \\
3 \\
4 \\
5 \\
1 \\
2 \\
3 \\
4 \\
5 \\
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix}
\begin{bmatrix}
\hat{\beta}_1 
\end{bmatrix}
$$


## Omitting the intercept

* Again, to better understand this design matrix, let's think about how prediction works: 

$\hat{y}_1$= 1 * `r round(coef(trend.mod2)[1],dig=3)`\

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(trend.mod2))),ceiling(max(KW$creepers,predict(trend.mod2))))
plot(KW$year[1],predict(trend.mod2)[1],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Omitting the intercept

$\hat{y}_1$= 1 * `r round(coef(trend.mod2)[1],dig=3)`\
$\hat{y}_2$= 2 * `r round(coef(trend.mod2)[1],dig=3)`

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(trend.mod2))),ceiling(max(KW$creepers,predict(trend.mod2))))
plot(KW$year[1:2],predict(trend.mod2)[1:2],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Omitting the intercept

$\hat{y}_1$= 1 * `r round(coef(trend.mod2)[1],dig=3)`\
$\hat{y}_2$= 2 * `r round(coef(trend.mod2)[1],dig=3)`\
$\cdots$\
$\hat{y}_5$= 5 * `r round(coef(trend.mod2)[1],dig=3)`

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(trend.mod2))),ceiling(max(KW$creepers,predict(trend.mod2))))
plot(KW$year[1:5],predict(trend.mod2)[1:5],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Omitting the intercept 

Why does this happen? 

 * Think about what happens in $year = 0$: $\hat{y}_1$= 0 * `r round(coef(trend.mod2)[1],dig=3)`\


## Omitting the intercept 

Why does this happen? 

* Think about what happens in $year = 0$: $\hat{y}_1$= 0 * `r round(coef(trend.mod2)[1],dig=3)`\

  - This coding forces the regression line through the intercept 

  - Intercepts are the default assumption because things can (sometimes) go haywire if you don't have one 
  
```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(0,ceiling(max(KW$creepers,predict(trend.mod2))))
plot(KW$year[1:5],predict(trend.mod2)[1:5],xlim = c(0,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(0,predict(trend.mod2,data.frame(year = 0)),pch=8)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Design matrices

What happens when we scale continuous covariates? 

`scale("x")` = $\frac{X_i - \bar{(X)}}{sd(X)}$

```{r, echo = FALSE, result = FALSE}

KW$temp_scale <- scale(KW$temp)
``` 

New values have mean at 0 and variance = standard deviation = 1 

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))
plot(KW$temp,KW$temp_scale, xlab = "temp",ylab = "scale(temp)")
```


## Scaling covariates 

* Given this design matrix, where the $x_i$ are the scaled values of temperature: 
$$
\begin{bmatrix}
1 & `r round(KW$temp_scale[1],dig = 3)` \\ 
1 & `r round(KW$temp_scale[2],dig = 3)` \\ 
1 & `r round(KW$temp_scale[3],dig = 3)` \\ 
1 & `r round(KW$temp_scale[4],dig = 3)` \\ 
1 & `r round(KW$temp_scale[5],dig = 3)` \\ 
1 & `r round(KW$temp_scale[6],dig = 3)` \\ 
1 & `r round(KW$temp_scale[7],dig = 3)` \\ 
1 & `r round(KW$temp_scale[8],dig = 3)` \\ 
1 & `r round(KW$temp_scale[9],dig = 3)` \\ 
1 & `r round(KW$temp_scale[10],dig = 3)` \\ 
1 & `r round(KW$temp_scale[11],dig = 3)` \\ 
1 & `r round(KW$temp_scale[12],dig = 3)` \\ 
1 & `r round(KW$temp_scale[13],dig = 3)` \\ 
1 & `r round(KW$temp_scale[14],dig = 3)` \\ 
1 & `r round(KW$temp_scale[15],dig = 3)` \\ 
\end{bmatrix}
$$ 


## Scaling covariates

If I wanted a prediction of wallcreeper occupancy at the mean temperature, I would predict at the mean value, i.e., $x_i = 0$ 

My prediction equation would be $y_{\text{mean temp}} = 1 * \beta_0 + 0 * \beta_1 = \beta_0$ 


## Scaling covariates

```{r, echo = TRUE, results = TRUE}

sumary(trend.mod3 <- lm(creepers ~ temp_scale,KW))

```


## Scaling covariates

```{r, echo = FALSE, results = FALSE}

plot(KW$temp_scale,KW$creepers, xlab = "scale(temp)",ylab = "% occupied")
abline(a = coef(trend.mod3)[1], b = coef(trend.mod3)[2],col="red")
points(KW$temp_scale,predict(trend.mod3),col = "red")
abline(v = 0)

```


## Scaling covariates

Notice how scaling changes our interpretation of $\boldsymbol{\beta}$

* Without scaling: 

   - $\beta_0$ is the predicted value for occupancy when temperature (C) = 0 
   - $\beta_1$ is the predicted change in occupancy for a one-degree change in temperature
   
* With scaling: 
   - $\beta_0$ is the predicted value for occupancy when temperature (C) = the mean of observed temperature
   - $\beta_1$ is the predicted change in occupancy for a one-unit change in scaled temperature, which is equal to a change of 1 SD in observed temperature
   

## Design matrices 

* Our design matrices so far have enforced a linear relationship between year and $\hat{y}$

* What if we think that wallcreeper occupancy is changing from year to year, but we don't think the relationship is linear? 

  - In other words, what if we think of `year` as a factor, or categorical variable? 
  
  
## Coding factors 

* Let's try: `model.fac1 <- lm(creepers ~ as.factor(year), KW)`

```{r, echo = TRUE, result = TRUE}

KW$yrfac <- as.factor(KW$year)
sumary(model.fac1 <- lm(creepers ~ yrfac, KW))
``` 


## Coding factors{.smaller}

* What does our design matrix look like? 

```{r, echo = TRUE, results = TRUE}

model.matrix(model.fac1)[1:15,]
```


## Prediction

* Again, let's look at how prediction works: 

$$
\begin{bmatrix}
\hat{y_1} \\ \hat{y_2} \\ \hat{y_3} \\ \hat{y_4} \\ \hat{y_5} \\ \hat{y_6} \\ \hat{y_7} \\ \hat{y_8} \\ \hat{y_9} \\ \hat{y_{10}} \\ \hat{y_{11}} \\ \hat{y_{12}} \\ \hat{y_{13}} \\ \hat{y_{14}} \\ \hat{y_{15}} \\ \end{bmatrix}
= 
\begin{bmatrix}
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\end{bmatrix}
\begin{bmatrix}
\hat{\beta}_0 \\ \hat{\beta}_1 \\ \hat{\beta}_2 \\ \hat{\beta}_3 \\ \hat{\beta}_4   
\end{bmatrix}
$$


## Prediction

$\hat{y}_1$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
0 * `r round(coef(model.fac1)[2],dig=3)` +
0 * `r round(coef(model.fac1)[3],dig=3)` +
0 * `r round(coef(model.fac1)[4],dig=3)` +
0 * `r round(coef(model.fac1)[5],dig=3)`\

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(model.fac1))),ceiling(max(KW$creepers,predict(model.fac1))))
plot(KW$year[1],predict(model.fac1)[1],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Prediction

$\hat{y}_1$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
0 * `r round(coef(model.fac1)[2],dig=3)` +
0 * `r round(coef(model.fac1)[3],dig=3)` +
0 * `r round(coef(model.fac1)[4],dig=3)` +
0 * `r round(coef(model.fac1)[5],dig=3)`\
$\hat{y}_2$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
1 * `r round(coef(model.fac1)[2],dig=3)` +
0 * `r round(coef(model.fac1)[3],dig=3)` +
0 * `r round(coef(model.fac1)[4],dig=3)` +
0 * `r round(coef(model.fac1)[5],dig=3)`\

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(model.fac1))),ceiling(max(KW$creepers,predict(model.fac1))))
plot(KW$year[1:2],predict(model.fac1)[1:2],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Prediction

$\hat{y}_1$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
0 * `r round(coef(model.fac1)[2],dig=3)`+
0 * `r round(coef(model.fac1)[3],dig=3)`+
0 * `r round(coef(model.fac1)[4],dig=3)`+ 
0 * `r round(coef(model.fac1)[5],dig=3)`\
$\hat{y}_2$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
1 * `r round(coef(model.fac1)[2],dig=3)` +
0 * `r round(coef(model.fac1)[3],dig=3)` +
0 * `r round(coef(model.fac1)[4],dig=3)` +
0 * `r round(coef(model.fac1)[5],dig=3)`\
$\cdots$\
$\hat{y}_5$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
0 * `r round(coef(model.fac1)[2],dig=3)` +
0 * `r round(coef(model.fac1)[3],dig=3)` +
0 * `r round(coef(model.fac1)[4],dig=3)` +
1 * `r round(coef(model.fac1)[5],dig=3)`\

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

par(mai = c(1,1,0.5,0.5))

ylims <- c(floor(min(KW$creepers,predict(model.fac1))),ceiling(max(KW$creepers,predict(model.fac1))))
plot(KW$year[1:5],predict(model.fac1)[1:5],xlim = c(1,5),ylim = ylims,ylab = expression(hat(y)), xlab = "year",pch=19)
points(KW$year[1:15],KW$creepers[1:15],col="red")
```


## Coding factors

* What if I want to compare all other years to year 5? 

$$
\text{Matrix 1}=
\begin{bmatrix}
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\end{bmatrix}
\text{Matrix 2}=
\begin{bmatrix}
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\end{bmatrix}
$$


## Coding factors 

* I need a custom design matrix 

```{r, echo = FALSE, result = TRUE}

X.mat <- matrix(NA,nrow=15,ncol=5)
X.mat[,1] <- rep(1,15)
X.mat[,2] <- rep(c(1,0,0,0,0),3)
X.mat[,3] <- rep(c(0,1,0,0,0),3)
X.mat[,4] <- rep(c(0,0,1,0,0),3)
X.mat[,5] <- rep(c(0,0,0,1,0),3)

X.mat
```  


## Coding factors

```{r, echo = TRUE, result = TRUE}

sumary(model.fac2 <- lm(creepers ~ -1 + X.mat, KW))

```


## Coding factors 

* Let's compare the coefficients from model 1 and model 2 

```{r, echo = FALSE, result = TRUE}

print(round(coef(model.fac1),dig=3))
print(round(coef(model.fac2),dig=3))
```

* Notice: 

  - Model 1: 
$\hat{y}_1$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
0 * `r round(coef(model.fac1)[2],dig=3)` +
0 * `r round(coef(model.fac1)[3],dig=3)` +
0 * `r round(coef(model.fac1)[4],dig=3)` +
0 * `r round(coef(model.fac1)[5],dig=3)` = 
`r round(predict(model.fac1)[1],dig=3)`\

  - Model 2: 
$\hat{y}_1$= 1 * `r round(coef(model.fac2)[1],dig=3)` +
1 * `r round(coef(model.fac2)[2],dig=3)` +
0 * `r round(coef(model.fac2)[3],dig=3)` +
0 * `r round(coef(model.fac2)[4],dig=3)` +
0 * `r round(coef(model.fac2)[5],dig=3)` =
`r round(predict(model.fac2)[1],dig=3)`\


## Coding factors

$$
\text{Matrix 1}=
\begin{bmatrix}
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\end{bmatrix}
\text{Matrix 2}=
\begin{bmatrix}
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\end{bmatrix}
$$

## Coding factors

* These design matrices will produce the same predictions

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

plot(predict(model.fac1),predict(model.fac2),xlab = "Model 1 Predictions",ylab = "Model 2 Predictions",main = "",pch=19)
abline(a = 0, b = 1)

```


## Coding factors

* Why not just skip the intercept?  

$$
\text{Matrix 1}=
\begin{bmatrix}
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\end{bmatrix}
\text{Matrix 3}=
\begin{bmatrix}
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
0 & \color{red}{1} & 0 & 0 & 0 \\ 
0 & 0 & \color{red}{1} & 0 & 0 \\ 
0 & 0 & 0 & \color{red}{1} & 0 \\ 
0 & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
0 & \color{red}{1} & 0 & 0 & 0 \\ 
0 & 0 & \color{red}{1} & 0 & 0 \\ 
0 & 0 & 0 & \color{red}{1} & 0 \\ 
0 & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & 0 & 0 & 0 & 0 \\ 
0 & \color{red}{1} & 0 & 0 & 0 \\ 
0 & 0 & \color{red}{1} & 0 & 0 \\ 
0 & 0 & 0 & \color{red}{1} & 0 \\ 
0 & 0 & 0 & 0 & \color{red}{1} \\ 
\end{bmatrix}
$$


## Omitting the intercept  

```{r, echo = TRUE, result = TRUE}

sumary(model.fac3 <- lm(creepers ~ -1 + yrfac, KW))

```


## Omitting the intercept  

* Let's compare the coefficients from model 1 and model 3 

```{r, echo = FALSE, result = TRUE}

print(round(coef(model.fac1),dig=3))
print(round(coef(model.fac3),dig=3))
```

Notice: 

* Model 1: 
$\hat{y}_1$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
0 * `r round(coef(model.fac1)[2],dig=3)` +
0 * `r round(coef(model.fac1)[3],dig=3)` +
0 * `r round(coef(model.fac1)[4],dig=3)` +
0 * `r round(coef(model.fac1)[5],dig=3)` =
`r round(predict(model.fac1)[1],dig=3)`\

* Model 3: 
$\hat{y}_1$= 1 * `r round(coef(model.fac3)[1],dig=3)` +
0 * `r round(coef(model.fac3)[2],dig=3)` +
0 * `r round(coef(model.fac3)[3],dig=3)` +
0 * `r round(coef(model.fac3)[4],dig=3)` +
0 * `r round(coef(model.fac3)[5],dig=3)` = `r round(predict(model.fac3)[1],dig=3)`\

$~$
$~$

* Model 1: 
$\hat{y}_2$= 1 * `r round(coef(model.fac1)[1],dig=3)` +
1 * `r round(coef(model.fac1)[2],dig=3)` +
0 * `r round(coef(model.fac1)[3],dig=3)` +
0 * `r round(coef(model.fac1)[4],dig=3)` +
0 * `r round(coef(model.fac1)[5],dig=3)` =
`r round(predict(model.fac1)[2],dig=3)`\

* Model 3: 
$\hat{y}_2$= 0 * `r round(coef(model.fac3)[1],dig=3)` +
1 * `r round(coef(model.fac3)[2],dig=3)` +
0 * `r round(coef(model.fac3)[3],dig=3)` +
0 * `r round(coef(model.fac3)[4],dig=3)` +
0 * `r round(coef(model.fac3)[5],dig=3)` = `r round(predict(model.fac3)[2],dig=3)`\


## Omitting the intercept  

* These design matrices will produce the same predictions

```{r, echo = FALSE, results = FALSE, fig.height = 4, fig.width = 6}

plot(predict(model.fac1),predict(model.fac3),xlab = "Model 1 Predictions",ylab = "Model 3 Predictions",main = "",pch=19)
abline(a = 0, b = 1)

```


## Coding Factors

When we code a model with one or more factors, our predictions will be the mean for the relevant levels of the factor(s)

```{r, echo = TRUE, results = TRUE}

coef(model.fac3) #model coefficients 

unique(predict(model.fac3)) #unique predictions 

t(aggregate(KW$creepers, by = list(KW$year), FUN = "mean")$x) #data means 

```


## Design matrices 

* We've seen design matrices for continuous and categorical (factor) variables 

* We use indicator (or "dummy") variables to code factors: 
$X_{i,j} = 1$ if sample $i$ is in factor level $j$ and $X_{i,j} = 0$ otherwise

* Examples of categorical (factor) variables: 
  - Area 
  - Group
  - Sex
  - Species
  - Age
  - Year 

* Some predictors can be either continuous or categorical predictors (Age, Year) but the models will be very different in these two cases   


## Think, Pair, Share

* Why might I want a design matrix like the one below? 
  
$$
\begin{bmatrix}
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & \color{red}{-1} & \color{red}{-1} & \color{red}{-1} & \color{red}{-1} \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & \color{red}{-1} & \color{red}{-1} & \color{red}{-1} & \color{red}{-1} \\ 
\color{red}{1} & \color{red}{1} & 0 & 0 & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} & 0 & 0 \\ 
\color{red}{1} & 0 & 0 & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & 0 & 0 & \color{red}{1} \\ 
\color{red}{1} & \color{red}{-1} & \color{red}{-1} & \color{red}{-1} & \color{red}{-1} \\ 
\end{bmatrix}
$$


## Coding factors 

* The intercept will be equal to the grand mean (the mean of group means) when we have a design matrix like $\boldsymbol{X}$  

$$
\text{Means Matrix}=
\begin{bmatrix}
\color{red}{1} & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} \\ 
\color{red}{1} & \color{red}{-1} & \color{red}{-1} \\ 
\color{red}{1} & \color{red}{1} & 0 \\ 
\color{red}{1} & 0 & \color{red}{1} \\ 
\color{red}{1} & \color{red}{-1} & \color{red}{-1} \\ 
\end{bmatrix}
$$


## Means design matrix

The intercept will be equal to the grand mean (the mean of group means) when we have this type of design matrix 

Recognize that $\hat{y}_{f1} = \bar{y}_{f1}$

$$
\text{Eq. 1:} ~~ \hat{y}_1 = \bar{y}_1 = 1*\hat{\beta}_0 + 1*\hat{\beta}_1 + 0*\hat{\beta}_2 = \hat{\beta}_0 + \hat{\beta}_1 \\
$$

<br>

$$
\text{Eq. 2:} ~~ \hat{y}_2 = \bar{y}_2 = 1*\hat{\beta}_0 + 0*\hat{\beta}_1 + 1*\hat{\beta}_2 = \hat{\beta}_0 + \hat{\beta}_2 \\
$$

<br> 

$$
\text{Eq. 3:} ~~ \hat{y}_3 = \bar{y}_3 = 1*\hat{\beta}_0 + -1*\hat{\beta}_1 + -1*\hat{\beta}_2 = \hat{\beta}_0 - \hat{\beta}_1 - \hat{\beta}_2
$$

## Means design matrix

Show that $\beta_0 = \frac{\bar{y}_1 + \bar{y}_2 + \bar{y}_3}{3}$

* From Eq.1: $\hat{\beta}_1 = \bar{y}_1 - \hat{\beta}_0$

* From Eq.2: $\hat{\beta}_2 = \bar{y}_2 - \hat{\beta}_0$

* Substituting into Eq. 3: 

$\bar{y}_3 = \underline{\hat{\beta}_0 - (\bar{y}_1 - \hat{\beta}_0) - (\bar{y}_2 - \hat{\beta}_0)} = \underline{\hat{\beta}_0 - \bar{y}_1 + \hat{\beta}_0 - \bar{y}_2 + \hat{\beta}_0}$ 

$~~~~ = \underline{3*\hat{\beta}_0 - \bar{y}_1 - \bar{y}_2}$

$~~~~3*\hat{\beta}_0 = \bar{y}_3 + \bar{y}_1 + \bar{y}_2$

$\therefore \hat{\beta}_0 = \frac{\bar{y}_1 + \bar{y}_2 + \bar{y}_2}{3}$


## Means design matrix

* When a design is balanced (equal samples in each group), the grand mean is equal to the mean of all the data, i.e., $\beta_0 = \bar{y}$ when the design is balanced


## Summary

* We've seen how to code categorical variables and continuous variables

* We've learned that a single model can be coded in multiple different ways 

  - The predictions will be the same 
  
  - The sums-of-squares will be the same 
  
  - But the coefficients will have different interpretations
  
* Next we'll move on to some more complex models including: 

  - Multiple continuous variables 
 
  - Continuous and categorical variables 
 
  - Interactions and nesting 