---
title: "Generalized Additive Models, Part 2"
author: "Sarah Converse"
date: "4 June 2025"
output:
  ioslides_presentation
subtitle: Analysis of Ecological and Environmental Data<br>QERM 514
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(mgcv)
library(curl)
library(ggplot2)
library(gamair)
library(gratia)
library(here)


```


```{css, echo = FALSE}

.title-slide hgroup h1 {
  font-size: 40px;
}
.title-slide hgroup h2 {
  font-size: 30px;
}
h2 {
  font-size: 40px;
}
slides > slide {
  font-size: 20px;
}
```

## Acknowledgements 

Gavin Simpson and colleagues have produced a lot of useful teaching material on GAMs that I've made use of here. For example: 

* https://www.youtube.com/watch?v=sgw4cu8hrZM&t=598s

* https://www.youtube.com/watch?v=Ukfvd8akfco&t=905s


## Goals for today

* Understand some diagnostics

* Understand more about model selection techniques with GAMs

* Become familiar with some of the flexibility of GAMs


## Last time 

A generalized additive model is a generalization of a GLM. It has a predictor that is a sum of smoothed (and possibly unsmoothed) terms: 

$$
y_i = \beta_0 + \Sigma_j s_j(x_{j,i}) + \epsilon_j
$$

And with splines, our smoothed terms are a sum of weighted basis functions: 

$$
s(x)=\Sigma_k \beta_k b_k(x)
$$

## Model choices 

Based on understanding of your system: 

* error distribution 

* predictors 

* type of smoother(s)  

Check with diagnostics 

* <b> $k$ - the number of knots </b> 


## An example 

Chicago pollution and death rate (Peng et al.; NMMAPSdata package)

* deaths per day
* median particles <2.5 mg per $m^3$
* ozone ppb 
* temperature (fahrenheit)

```{r, echo = TRUE, library = TRUE}

data(chicago)

model1 <- mgcv::gam(death ~ s(pm25median) + 
                      s(o3median) + 
                      s(tmpd), 
                    family = "poisson", 
                    data=chicago, method = "REML")

```


## An example 

Smoothed Terms 

```{r, echo = FALSE, results = TRUE}

summary(model1)$s.table 

``` 


## GAM interpretation  

* Effective degrees of freedom (edf) - degree of non-linearity, but with upper bound set by $k$ 

* Reference degrees of freedom (Ref.df) - used for calculating test statistic 

* $p$-value - null hypothesis is a linear effect (this can include a flat linear effect, i.e., no effect) 

  * Biased low because they don't account for need to estimate $\lambda$ 

  * Works well in asymptopia 

  * Doesn't work for unpenalized smooths (random effects)


## Model checking 

* Effect of $k$ on edf (for median particles <2.5 mg per $m^3$)

```{r,echo = FALSE,results = TRUE}

record <- matrix(NA,nrow=20,ncol=3)
for(j in 1:20){
  model1 <- mgcv::gam(death ~ s(pm25median, k = j+3) + 
                      s(o3median) + 
                      s(tmpd), 
                    family = "poisson", 
                    data=chicago, method = "REML")
  
record[j,1] <- j+3
record[j,2] <- summary(model1)$s.table[1,1]
}

plot(record[,1],record[,2],xlab="k for pm2.5",ylab="effective degrees of freedom")

```


## Model checking 

* Check whether $k$ is adequate - can set it very large to start, but this increases computation time 

* Do some iterative checking with larger values of $k$ for multiple parameters, effective degrees of freedom should stabilize   

```{r, echo = TRUE, results = 'hide', eval = FALSE}

gam.check(model1) 

``` 


## Model checking 

With $k$ set to default for each variable: check $p$-values 

```{r, echo = FALSE, results = TRUE, fig.show='hide'}

set.seed(1)
model1 <- mgcv::gam(death ~ s(pm25median) + 
                      s(o3median) + 
                      s(tmpd), 
                    family = "poisson", 
                    data=chicago, method = "REML")

gam.check(model1) 

``` 


## Model checking  

With $k$ set to 20 for each variable 


```{r}
set.seed(1)
``` 


```{r, echo = TRUE, results = FALSE, fig.show='hide'}

model1 <- mgcv::gam(death ~ s(pm25median, k=20) + 
                      s(o3median, k=20) + 
                      s(tmpd, k=20), 
                    family = "poisson", 
                    data=chicago, method = "REML")

gam.check(model1) 

```


## Model checking  

With $k$ set to 20 for each variable: $p$-value above $\alpha$, and edf is not close to $k$ 

```{r, echo = FALSE, results = TRUE, fig.show='hide'}

model1 <- mgcv::gam(death ~ s(pm25median, k=20) + 
                      s(o3median, k=20) + 
                      s(tmpd, k=20), 
                    family = "poisson", 
                    data=chicago, method = "REML")

gam.check(model1, rep=500)

```


## Model checking  

* We need to check the other assumptions of our model as well 

* gam.check() also provides some basic diagnostic plots 

```{r, echo = FALSE, results = TRUE, fig.show='show'}

gam.check(model1) 

``` 


## Model selection 

* When we select $\lambda$ we are only penalizing wiggliness, thus only penalizing functions of $X$ with a second derivative that is >0

* Linear relationships can't be removed from the model this way  


## Model selection 

In gam(), we can use "select=TRUE" to apply penalties to the linear terms as well

```{r, echo = TRUE, results = FALSE}

model1 <- mgcv::gam(death ~ s(pm25median, k = 20) + 
                      s(o3median, k = 20) + 
                      s(tmpd, k = 20), 
                    family = "poisson", select = TRUE,  
                    data=chicago, method = "REML")
                    
```


## Model selection 

Look to see if edf is less than 1 

```{r, echo = FALSE, results = TRUE}

summary(model1)$s.table

```


## Model selection

```{r, echo = FALSE, results = TRUE}

gratia::draw(model1, scales = "fixed")

```


## Flexibility of GAMs 

* We have lots of options when fitting GAMs 

* This flexibility means we have to give very careful thought to what we're trying to achieve 

* Just because we can fit a GAM doesn't mean it will be interpretable or useful  


## Example  

Atmospheric carbon dioxide data (measured at the South Pole) since 1957 to ~1999 

```{r, echo = FALSE, results = TRUE}

data(co2s)

ggplot(co2s, aes(x = c.month, y = co2)) + geom_line()

```

## Example  

Using the cubic regression basis - better when we have a lot of wiggliness 

Make $k$ large (lots of potential wiggliness)


```{r, echo = TRUE, results = FALSE}

b <- gam(co2 ~ s(c.month, k=300, bs="cr"), data = co2s, method = 'REML')

``` 


## Example  

```{r, echo = FALSE, results = TRUE}

summary(b)

```


## Example  

Now try to predict from the model 

```{r, echo = FALSE, results = TRUE}

pd <- with(co2s, data.frame(c.month = 1:(nrow(co2s)+36)))
pd <- cbind(pd, predict(b, pd, se = TRUE))
pd <- transform(pd, upr = fit + (2*se.fit), lwr = fit - (2 * se.fit))


ggplot(pd, aes(x = c.month, y = fit)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
    geom_line(data = co2s, aes(c.month, co2), col = 'red') +
    geom_line(alpha = 0.5)

```

## Example  

We fit a model, but it isn't particularly interpretable or useful 

Let's distinguish within-year variation from the long-term trend 

* Fit a cyclic spline on month (1:12) and a cubic regression spline on cumulative month (1:507)

```{r, echo = TRUE, results = TRUE}

b2 <- gam(co2 ~ s(month, bs = "cc") + 
            s(c.month, bs = "cr", k = 300), 
            knots = list(month = c(0.5, 12.5)),
            data = co2s, method = 'REML')
```


## Example  

```{r, echo = FALSE, results = TRUE}

summary(b2)

```


## Example  

```{r, echo = FALSE, results = TRUE} 

nr <- nrow(co2s)
pd2 <- with(co2s, data.frame(c.month = 1:(nr+36),
                             month = rep(1:12, length.out=nr+36)))
pd2 <- cbind(pd2, predict(b2, pd2, se = TRUE))
pd2 <- transform(pd2, upr = fit + (2*se.fit), lwr = fit - (2 * se.fit))

ggplot(pd2, aes(x = c.month, y = fit)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
    geom_line(data = co2s, aes(c.month, co2), col = 'red') +
    geom_line(alpha = 0.5)
``` 


## Example  

* The second model seems more interpretable (there is predictable annual variation, but also a long-term pattern) and is more useful for prediction 

* Building GAMs is more complex than building GLMs because we have more choices 

* Requires thinking hard about the system and the question 


## Options for interactions 

In addition to splines on individual predictors, we can fit interactions 

* te(x1,x2) - main effects and interactions - separate smoothness parameter and separate marginal basis 

* ti(x1,x2) - interactions only - for when you have the main effect elsewhere in the model (for example, if you have multiple two-way interactions, a*b + a*c + b*c)

* s(x1,x2, bs = 'tp') - bivariate thin plate splines - single smoothness parameter, only use when units are the same (e.g., spatial smoother)

* f + s(x1, by = f) - when we want to smooth by different levels of a factor (be sure you include the factor as a fixed effect)

* s(x, f, bs = 'fs') - smooth function for each level of the factor, akin to random slopes 


## Mixed Models 

We can also fit random effects

* mgcv:gam() with bs = "re"

* mgcv:gamm() - nlme - with correlation structures 

* gamm4:gamm4() - lmer() or glmer()

Thus, these models are equivalent: 

* lme(y ~ 1 + (1|f), method = "REML")

* gam(y ~ s(f, bs = "re"), method = "REML")


## Summary 

* GAMs using splines offer tremendous flexibility 

* There is likely some tradeoff between flexibility and ease of understanding 

* Many of our relationships in ecology aren't linear 

* What are some examples? 


