---
title: "Introduction to linear models"
subtitle: "Analysis of Ecological and Environmental Data<br>QERM 514"
author: "Mark Scheuerell"
date: "6 April 2020"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Goals for today

* Identify whether a model is linear in the predictors


## A simple starting point

### Data = (Deterministic part) + (Stochastic part)


## Two questions to ask:

1. What is the distributional form of the random processes?

```{r error_dist_ex, fig.height = 4, fig.width = 7, fig.align = 'center'}
par(mfrow = c(1, 2), mai = c(0.5, 0.5, 0.5, 0.1), omi = rep(0, 4), bg = NA)

hist(rnorm(1e5), breaks = 50,
     yaxt = "n", xlab = expression(epsilon), main = "Normal")

hist(rnbinom(1e5, 5, 0.3), breaks = 50,
     yaxt = "n", xlab = expression(epsilon), main = "Negative binomial")
```


## Two questions to ask:

1. What is the distributional form of the random processes?

2. How many random processes are there?

$$
\text{abundance} = f(\text{species}, \text{population}, \text{time}) \\
$$


## Forms of linear models

| Errors | Single random process | Multiple random processes |
|:-------|:--------------:|:------------------:|
| Normal | Linear Model (LM) | Linear Mixed Model (LMM) |
| Non-normal | Generalized Linear Model (GLM) | Generalized Linear Mixed Model (GLMM) |


## Forms of linear models
 
```{r model_diagram, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5)
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Forms of linear models
 
```{r model_diagram_lm, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("black", "blue", "black", "black"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("black", "blue", "black", "black"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Linear models for size of fish

| Model | Description |
|:------|:-----------:|
| $\text{size}_i = \alpha + \beta \text{species}_i + \epsilon_i$ | 1-way ANOVA |
| $\text{size}_i = \alpha + \beta \text{species}_i + \beta_2 \text{site}_i + \epsilon_i$ | 2-way ANOVA |
| $\text{size}_i = \alpha + \beta \text{age}_i + \epsilon_i$ | simple linear regression |
| $\text{size}_i = \alpha + \beta \text{age}_i + \beta_2 \text{temperature}_i + \epsilon_i ~ ~$ | multiple regression |


## Forms of linear models
 
```{r model_diagram_glm, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("blue", "black", "black", "black"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("blue", "black", "black", "black"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## General approach

### Question $\rightarrow$ Data $\rightarrow$ <u>Model</u> $\rightarrow$ Inference $\rightarrow$ Prediction


## Linear models for size of fish

```{r lw_regr, fig.align = 'center', fig.height = 5, fig.width = 5}
## simulated data
set.seed(123)
nn <- 35
L10_length <- log(runif(nn, 100, 400), 10)
L10_mass <- -5.3 + 3.1 * L10_length + rnorm(nn, 0, 0.2)

par(mai = c(1, 1, 0.1, 0.1), omi = rep(0, 4), bg = NA)
plot(10^(L10_length), 10^(L10_mass), pch = 16, log = "xy",
     ylab = "Mass (g)", xlab = "Length (mm)", cex.lab = 1.5)
```


## Linear models for size of fish

Two simple choices:

$\text{log}_{10}(mass_i) = \alpha + \epsilon_i$

$\text{log}_{10}(mass_i) = \alpha + \beta ~ \text{log}_{10}(length_i) + \epsilon_i$


## Linear models for size of fish

In **R**, we fit these with `lm()`

$\text{log}_{10}(mass_i) = \alpha + \epsilon_i$

`lm(L10_mass ~ 1)` 

$\text{log}_{10}(mass_i) = \alpha + \beta ~ \text{log}_{10}(length_i) + \epsilon_i$

`lm(L10_mass ~ L10_length)` 


## General approach

### Question $\rightarrow$ Data $\rightarrow$ Model $\rightarrow$ <u>Inference</u> $\rightarrow$ Prediction


## Model 1 in R {.smaller}

```{r, echo = TRUE}
m1 <- lm(L10_mass ~ 1)
summary(m1)
```


## Model 2 in R {.smaller}

```{r, echo = TRUE}
m2 <- lm(L10_mass ~ L10_length)
summary(m2)
```


## General approach

### Question $\rightarrow$ Data $\rightarrow$ <u>Model</u> $\rightarrow$ Inference $\rightarrow$ Prediction


## Linear models for size of fish

```{r lw_regr, fig.align = 'center', fig.height = 5, fig.width = 5}
## simulated data
set.seed(123)
nn <- 35
L10_length <- log(runif(nn, 100, 400), 10)
L10_mass <- -5.3 + 3.1 * L10_length + rnorm(nn, 0, 0.2)

par(mai = c(1, 1, 0.1, 0.1), omi = rep(0, 4), bg = NA)
plot(10^(L10_length), 10^(L10_mass), pch = 16, log = "xy",
     ylab = "Mass (g)", xlab = "Length (mm)", cex.lab = 1.5)
```


## Linear models for size of fish

Two simple choices:

$\text{log}_{10}(mass_i) = \alpha + \epsilon_i$

$\text{log}_{10}(mass_i) = \alpha + \beta ~ \text{log}_{10}(length_i) + \epsilon_i$


## Linear models for size of fish

In **R**, we fit these with `lm()`

$\text{log}_{10}(mass_i) = \alpha + \epsilon_i$

`lm(L10_mass ~ 1)` 

$\text{log}_{10}(mass_i) = \alpha + \beta ~ \text{log}_{10}(length_i) + \epsilon_i$

`lm(L10_mass ~ L10_length)` 


## General approach

### Question $\rightarrow$ Data $\rightarrow$ Model $\rightarrow$ <u>Inference</u> $\rightarrow$ Prediction


## Model 1 in R {.smaller}

```{r, echo = TRUE}
m1 <- lm(L10_mass ~ 1)
summary(m1)
```


## Model 2 in R {.smaller}

```{r, echo = TRUE}
m2 <- lm(L10_mass ~ L10_length)
summary(m2)
```


## Which model should we choose?

We can an *information criterion* to select the one with the most *data support*

For example, Akaike's Information Criterion (AIC)

$$
\text{AIC} = -2 \text{LL} + 2 p
$$

where $\text{NLL}$ is the log-likelihood and $p$ is the number of *fitted* parameters

The model with the **lowest** AIC is considered the "best"


## Model selection

In **R** we can use the `AIC()` function

```{r, echo = TRUE}
AIC(m1, m2)
```


## Model selection

In **R** we can use the `AIC()` function

```{r, echo = TRUE}
AIC(m1, m2)
```

Model 2 has the lowest AIC and is therefore the most parsimonious


## General approach

### Question $\rightarrow$ Data $\rightarrow$ Model $\rightarrow$ Inference $\rightarrow$ <u>Prediction</u>


## Prediction of novel outcomes

Once we have a (reasonable) model, we can do some prediction

For example, what is the expected mass of a 250 mm fish?


## Prediction of novel outcomes

```{r lpredict, fig.align = 'center', fig.height = 5, fig.width = 5}
fit_mass <- predict(m2)

par(mai = c(1, 1, 0.1, 0.1), omi = rep(0, 4), bg = NA)
plot(10^(L10_length), 10^(L10_mass), pch = 16, log = "xy",
     ylab = "Mass (g)", xlab = "Length (mm)", cex.lab = 1.5)
lines(10^(L10_length), 10^(fit_mass))
```


## Prediction of novel outcomes

Alternatively, we can use the `predict()` function in **R**

```{r, echo = TRUE}
## prediction on the log10 scale
L10_mass_pred <- predict(m2, data.frame(L10_length = log(250, 10)))
## prediction in normal space
round(10^L10_mass_pred, 1)
```


## Uncertainty in our prediction?

```{r predict_int, fig.align = 'center', fig.height = 5, fig.width = 5, warning = FALSE}
fit_mass_2 <- predict(m2, interval = "prediction")

par(mai = c(1, 1, 0.1, 0.1), omi = rep(0, 4), bg = NA)
plot(10^(L10_length), 10^(L10_mass), pch = 16, log = "xy",
     ylab = "Mass (g)", xlab = "Length (mm)", cex.lab = 1.5)
lines(10^(L10_length), 10^(fit_mass_2[,"fit"]))
lines(10^(L10_length), 10^(fit_mass_2[,"upr"]), col = "blue")
lines(10^(L10_length), 10^(fit_mass_2[,"lwr"]), col = "blue")
```


## Uncertainty in our prediction?

More formally, we can use the `predict()` function in **R**

```{r, echo = TRUE}
## prediction on the log10 scale
L10_mass_pred <- predict(m2, data.frame(L10_length = log(250, 10)),
                         interval = "prediction")
## prediction in normal space
round(10^L10_mass_pred, 1)
```

