---
title: "Inference, Part 2"
author: "Sarah Converse"
date: "11 April 2025"
output:
  ioslides_presentation
subtitle: Analysis of Ecological and Environmental Data<br>QERM 514

---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(faraway)
```

```{css, echo = FALSE}

.title-slide hgroup h1 {
  font-size: 40px;
}
.title-slide hgroup h2 {
  font-size: 30px;
}
h2 {
  font-size: 40px;
}
slides > slide {
  font-size: 20px;
}
```


## Objectives for today

* Understand how $F$-tests are used to compare (nested) models

* Understand how to calculate confidence intervals


## Inferential methods

Once we've estimated the model parameters and their variance, we will want to draw conclusions from our analysis

One question we often ask is which model is "best"


## Our wallcreepers example

1. The **year** model: $creepers_i = \beta_0 + \beta_1*year_i$

2. The **year + temperature** model: $creepers_i = \beta_0 + \beta_1*year_i + \beta_2*temperature_i$

Is the complexity of model 2 necessary or is 1 good enough? 


## Hypothesis test to compare models

Recall our partitioning of sums-of-squares, where

$$
SSTO = SSR + SSE
$$

We might prefer the more complex model (call it $\Theta$) over the simple model (call it $\theta$) if 

$$
\frac{SSE_{\theta} - SSE_{\Theta}}{SSE_{\Theta}} > \text{a constant}
$$

Is the proportional decrease in the error enough to justify the additional parameter? 

Note that, by definition $SSE_{\Theta} \leq SSE_{\theta}$


## Hypothesis test to compare models

If $\Theta$ has $k_{\Theta}$ parameters and $\theta$ has $k_{\theta}$, we can scale this ratio to arrive at an $F$-statistic that follows an $F$ distribution

$$
F = \frac{ \left( SSE_{\theta} - SSE_{\Theta} \right)  / (k_{\Theta} - k_{\theta})}{ SSE_{\Theta}  / (n - k_{\Theta})} \sim F_{k_{\Theta} - k_{\theta}, n - k_{\Theta}}
$$


## $F$-distribution

```{r f_dist, fig.height=4, fig.width=6, fig.align="center"}
par(mai = c(0.5,0.5,0.1,0.1), omi = rep(0,4))
curve(df(x, df1=2, df2=1), from=0, to=4, lwd = 2, col = "orange",
      ylab = "", xlab = expression(italic(x)), las = 1)
curve(df(x, df1=4, df2=10), from=0, to=4, lwd = 2, col = "red", add = TRUE)
curve(df(x, df1=6, df2=25), from=0, to=4, lwd = 2, col = "blue", add = TRUE)
curve(df(x, df1=8, df2=50), from=0, to=4, lwd = 2, col = "purple", add = TRUE)
legend(3, 0.9, legend = c("df1 = 2, df2 = 1",
         "df1 = 4, df2 = 10",
         "df1 = 6, df2 = 25",
         "df1 = 8, df2 = 50"),
       col = c("orange", "red", "blue", "purple"),
       lwd = 1.1, bty = "n", xjust = 0.5, y.intersp = 1.2)
```


## $F$-test 

We can use the $F$-test to: 

* Test all the predictors ($H_0: \beta_1 = \cdots = \beta_p = 0$)

* Test one predictor ($H_0: \beta_i = 0| \text{other predictors in the model}$) 

* Test a subset (e.g., $H_0: \beta_i = \beta_j$) such as $y_i = \beta_0 + \beta_1 X1_i + \beta_2 X2_i$ vs $y_i = \beta_0 + \beta_1 (X1_i + X2_i)$

But notice in all of these tests, we can only test nested models. We can’t test, for example:  

* $y_i = \beta_0 + \beta_1 X1_i$ vs. $y_i = \beta_0 + \beta_2 X2_i$

We will address model selection more fully later in the quarter 


## Our wallcreepers example

1. The **year** model: $creepers_i = \beta_0 + \beta_1*year_i$

2. The **year + temperature** model: $creepers_i = \beta_0 + \beta_1*year_i + \beta_2*temperature_i$

Is the complexity of model 2 necessary or is 1 good enough? 

```{r, echo = FALSE}
set.seed(12)

years <- 16
year <- c(1990:2005)-1989
temp <- rnorm(16,0,2)
b0 <- 30 
b1 <- -0.8
b2 <- -0.2
sigma <- 3
creepers <- b0 + b1*year + b2*temp + rnorm(years,0,sigma)

data <- data.frame(year,creepers, temp)
```

```{r, echo = TRUE}

#run our models 
mod1 <- lm(creepers ~ year, data = data)
mod2 <- lm(creepers ~ year + temp, data = data)
```


## Our wallcreepers example

```{r,echo = TRUE}

#Calculate error SS for each model 
SSEm1 <- (t(creepers-predict(mod1)))%*%(creepers-predict(mod1))
SSEm2 <- (t(creepers-predict(mod2)))%*%(creepers-predict(mod2))

#Calculate our F-statistic and p-value
(F.st <- ((SSEm1 - SSEm2)/(3-2))/ ((SSEm2)/(years-3)))
pf(F.st, 1, 13, lower.tail = FALSE) 

``` 


## Our wallcreepers example

```{r,echo = TRUE}

#Check against model output - it matches! 
anova(mod2)

``` 


## Test of all predictors in a model

We can also test whether the collection of predictors in a model are better than the model of the mean 

$$
\Theta: \mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \mathbf{e} \\
\theta: \mathbf{y} = \boldsymbol{\mu} + \mathbf{e} \\
$$

We write the null hypothesis as

$$
H_0: \beta_1 = \beta_2 = \dots = \beta_k = 0
$$

and we would reject $H_0$ if $F > F^{(\alpha)}_{k_{\Theta} - k_{\theta}, n - k_{\Theta}}$


## Hypothesis test to compare models

$$
SSE_{\Theta} = \left( \mathbf{y} - \mathbf{X} \boldsymbol{\beta} \right)^{\top} \left( \mathbf{y} - \mathbf{X} \boldsymbol{\beta} \right) = \mathbf{e}^{\top} \mathbf{e} = SSE \\
SSE_{\theta} = \left( \mathbf{y} - \bar{y} \right)^{\top} \left( \mathbf{y} - \bar{y} \right) =  SSTO \\
\Downarrow \\
F = \frac{ \left( SSTO - SSE \right)  / (k - 1) } { SSE  / (n - k)}
$$


## Our wallcreepers example

```{r, echo = TRUE, size="small"}

#Run models 
mod0 <- lm(creepers ~ 1, data = data)
mod2 <- lm(creepers ~ year + temp, data = data)

#Get the SS 
SSTO <- (t(creepers-predict(mod0)))%*%(creepers-predict(mod0))
SSE.m2 <- (t(creepers-predict(mod2)))%*%(creepers-predict(mod2))

#Get F-statistic and p-value 
(F.st <- ((SSTO - SSE.m2)/(3-1))/ ((SSE.m2)/(years-3)))
pf(F.st, 2, 13, lower.tail = FALSE) 

```


## Our wallcreepers example

```{r, echo = TRUE}

#Check against model output - it matches! 
summary(mod2)$fstatistic

``` 


## Caveats about hypothesis testing

Null hypothesis testing can be overused 

* $p$-values are simply the probability of obtaining a test statistic as large or greater than that observed

* $p$-values are **not** weights of evidence

* "Critical" or "threshold" values against which to compare $p$-values must be chosen *a priori*

* Often the null hypothesis $H_0: \beta_1 = 0$ is patently silly 

* We will revisit this subject later 


## Think, pair, share  

Will you want to test nested models in your analysis, and what do you think about using null hypothesis tests to do so?  


## Confidence intervals for $\beta$

We often use confidence intervals (CI's) to express uncertainty in $\hat{\beta}_i$
  
  They take the form

$$
  100(1 - \alpha)\% ~ \text{CI}: \hat{\beta}_{i} \pm t_{n-p}^{(\alpha / 2)} \operatorname{SE}(\hat{\beta})
$$
  
  where here $\alpha$ is our *predetermined* Type-I error rate

The t-distribution is a symmetrical distribution much like the normal distribution but used because it is more conservative at smaller sample sizes 

As $n \rightarrow \infty$, the $t$-distribution $\rightarrow$ the normal distribution


## $t$-distribution

```{r t_dist, fig.height=4, fig.width=6, fig.align="center"}
x <- c(-5,5)
par(mai = c(0.5,0.5,0.1,0.1), omi = rep(0,4))
curve(dt(x, df=5), from=-5, to=5, lwd = 2, col = "orange",
      ylab = "", xlab = expression(italic(x)), las = 1,ylim = c(0,0.5) )
curve(dt(x, df=30), from=-5, to=5, lwd = 2, col = "red", add = TRUE)
curve(dt(x, df=100), from=-5, to=5, lwd = 2, col = "purple", add = TRUE)
legend(3, 0.3, legend = c("df = 5",
                          "df = 30",
                          "df = 100"),
       col = c("orange", "red", "blue", "purple"),
       lwd = 1.1, bty = "n", xjust = 0.5, y.intersp = 1.2)
```


## Bootstrap confidence intervals

The $t$-based confidence intervals we have described depend on the assumption of normality

The bootstrap method provides a way to construct CI's without this assumption

Bootstrap procedure:

1. Fit your model to the data  

2. Calculate $\mathbf{\epsilon} = \mathbf{y} - \mathbf{X} \hat{\boldsymbol{\beta}}$ 

3. Do the following *many* times:
    * Generate $\mathbf{\epsilon}^*$ by sampling *with replacement* from $\mathbf{\epsilon}$  
    * Calculate $\mathbf{y}^*$ = $\mathbf{X} \hat{\boldsymbol{\beta}} + \mathbf{\epsilon}^*$
    * Estimate $\hat{\boldsymbol{\beta}}^*$ from $\mathbf{X}$ & $\mathbf{y}^*$  

4. Select the $\tfrac{\alpha}{2}$ and $(1 - \tfrac{\alpha}{2})$ percentiles from the saved $\hat{\boldsymbol{\beta}}^*$


## Confidence interval for new predictions

Given a fitted model $\mathbf{y} = \mathbf{X} \hat{\boldsymbol{\beta}} + \mathbf{\epsilon}$, we might want to know the uncertainty around a new estimate $\mathbf{y}^*$ given some new predictor $\mathbf{X}^*$


## CI for the mean response

Suppose we wanted to estimate the uncertainty in the *average* response given by

$$
\hat{\mathbf{y}}^* = \mathbf{X}^* \hat{\boldsymbol{\beta}}
$$

Recall that the general formula for a CI on a quantity $z$ is

$$
100(1 - \alpha)\% ~ \text{CI}: \text{E}(z) ~ \pm ~ t^{(\alpha / 2)}_{df}\text{SD}(z)
$$

So we would have

$$
\hat{\mathbf{y}}^* ~ \pm ~ t^{(\alpha / 2)}_{df} \sqrt{\text{Var} \left( \hat{\mathbf{y}}^* \right)}
$$


## CI for the mean response

We can calculate the SD of our expectation as

$$
\begin{aligned}
\text{Var} \left( \hat{\mathbf{y}}^* \right) &= \text{Var} \left( \mathbf{X}^* \hat{\boldsymbol{\beta}} \right)  \\
  &= {\mathbf{X}^*} \text{Var}\left( \hat{\boldsymbol{\beta}} \right) {\mathbf{X}^*}^{\top} \\
  &= {\mathbf{X}^*} \left[ \sigma^2 (\mathbf{X}^{\top} \mathbf{X})^{-1} \right] {\mathbf{X}^*}^{\top} \\
  &\Downarrow \\
\text{SD} \left( \hat{\mathbf{y}}^* \right) &= \sigma \sqrt{ {\mathbf{X}^*} (\mathbf{X}^{\top} \mathbf{X})^{-1} {\mathbf{X}^*}^{\top}} 
\end{aligned}
$$


## CI for the mean response

So our CI on the mean response is given by

$$
\hat{\mathbf{y}}^* \pm ~ t^{(\alpha / 2)}_{df} \sigma \sqrt{ {\mathbf{X}^*} (\mathbf{X}^{\top} \mathbf{X})^{-1} {\mathbf{X}^*}^{\top}}  
$$


## CI for a specific response

What about the uncertainty in a *specific* prediction?

In that case we need to account for our additional uncertainty owing to the error in our relationship, which is given by

$$
\hat{\mathbf{y}}^* = \mathbf{X}^* \hat{\boldsymbol{\beta}} + \mathbf{\epsilon}
$$

## CI for a specific response

The SD of the new prediction is given by

$$
\begin{aligned}
\text{Var} \left( \hat{\mathbf{y}}^* \right) &= {\mathbf{X}^*} \text{Var}\left( \hat{\boldsymbol{\beta}} \right) {\mathbf{X}^*}^{\top} + \text{Var} \left( \mathbf{e} \right) \\
  &= {\mathbf{X}^*} \left[ \sigma^2 (\mathbf{X}^{\top} \mathbf{X})^{-1} \right] {\mathbf{X}^*}^{\top} + \sigma^2\\
  &= \sigma^2 \left( {\mathbf{X}^*} (\mathbf{X}^{\top} \mathbf{X})^{-1} {\mathbf{X}^*}^{\top} + 1 \right) \\
  &\Downarrow \\
\text{SD} \left( \hat{\mathbf{y}}^* \right) &= \sigma \sqrt{1 + {\mathbf{X}^*} (\mathbf{X}^{\top} \mathbf{X})^{-1} {\mathbf{X}^*}^{\top} } 
\end{aligned}
$$

## CI for a specific response

So our CI on the new prediction is given by

$$
\hat{\mathbf{y}}^* \pm ~ t^{(\alpha / 2)}_{df} \sigma \sqrt{1 + {\mathbf{X}^*} (\mathbf{X}^{\top} \mathbf{X})^{-1} {\mathbf{X}^*}^{\top} } 
$$

<br>

This is typically referred to as the *prediction interval*




