---
title: "Data transformations"
subtitle: "Analysis of Ecological and Environmental Data<br>QERM 514"
author: "Mark Scheuerell"
date: "17 April 2026"
output:
  ioslides_presentation:
    wide: true
    css: ../lecture_slides.css
header-includes:
  - \usepackage{amsmath,amssymb,xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Goals for today

* Identify possible transformations of the response variable when your errors have unequal variance or are skewed

> - Understand how to use common transformations and make inference from the resulting model

> - Understand that there are alternatives to transformation that we will use later


## Why would you transform?

We have made a number of assumptions about our models, which include

* the distribution of the errors (IID)

* linear relationship(s) between the response and predictor(s)

<br>

What can we do when these assumptions are not met?


## What can you transform?

It's possible to transform either or both sides of our models to

* achieve constant variance $\rightarrow$ transform $y$

* correct for skewness $\rightarrow$ transform $y$

* linearize the relationship $\rightarrow$ transform $x$ and/or $y$


## Types of transformations

The most common form is where $y^* = y^\lambda$

and $\lambda > 1$ (powers)

or $0 < \lambda < 1$ (roots)

<br>

For example

$\lambda = 2 \Rightarrow y^* = y^2$

$\lambda = \frac{1}{2} \Rightarrow y^* = \sqrt{y}$


## Types of transformations

One can also use inverses where $y^* = y^{-\lambda}$

and $\lambda > 1$ (powers)

or $0 < \lambda < 1$ (roots)

<br>

For example

$\lambda = 2 \Rightarrow y^* = \frac{1}{y^2}$

$\lambda = \frac{1}{2} \Rightarrow y^* = \frac{1}{\sqrt{y}}$


## Box-Cox transformation

The Box-Cox transformation is popular for stabilizing the variance of errors

It is defined as

$$
y^* = \frac{y^{\lambda} - 1}{\lambda}
$$

for all $y > 0$

## Box-Cox transformation

More specifically, because 

$$
\lim\limits_{\lambda \to 0} \frac{y^{\lambda} - 1}{\lambda} = \log(y)
$$

we instead use

$$
y^* = \left\{ 
\begin{matrix}
\frac{y^{\lambda} - 1}{\lambda} ~ \text{if} ~ \lambda \neq 0 \\
\log(y) ~ \text{if} ~ \lambda = 0
\end{matrix}
\right.
$$


## Box-Cox transformation

How does one actually _choose_ $\lambda$?

By using *profile likelihoods* (which we will see in a later lecture)

<br>

We'll use the `boxcox()` function from the **{MASS}** package


## Box-Cox transformation | An example

Let's return to the plant data from the Galapagos Archipelago where we modeled diversity as a function of island area

```{r model_gala, echo = TRUE, eval = FALSE}
## get data
data(gala, package = "faraway")
## fit regression model
mm <- lm(Species ~ Area, gala)
## estimate lambda
MASS::boxcox(mm)
```


## Box-Cox transformation

Here is the result of calling `boxcox(mm)`

```{r boxcox_gala, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
## get data
data(gala, package = "faraway")
## fit regression model
mm <- lm(Species ~ Area, gala)
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0.3, 0),
    cex = 1.1)
## plot profile & estimate lambda
llp <- MASS::boxcox(mm, lambda = seq(-100,100)/100)
lm <- llp$x[which.max(llp$y)]
mtext(side = 3, line = 0.5, at = lm, substitute(hat(lambda) == lm, list(lm = lm)))
```


## Box-Cox transformation

Okay, so how do we interpret $\lambda = 0.17$?

<br>

Box-Cox transformations work well, but sometimes we can do better with an approximation to $\lambda$


## Box-Cox transformation

General considerations

* The Box–Cox method gets upset by outliers

For example, if $\hat{\lambda}$ = 5 there is little rationale for such an
extreme transformation


## Box-Cox transformation

General considerations

* The Box–Cox method gets upset by outliers

* If some $y_i$ < 0, we can add a constant to all the $y$

This works if the constant is small, but it's a "hack"



## Box-Cox transformation

General considerations

* The Box–Cox method gets upset by outliers

* If some $y_i$ < 0, we can add a constant to all the $y$

* If the range in $y$ is small, then the Box–Cox transformation has little effect

Recall that linear models work well for *local* non-linear functions


## Alternative to Box-Cox

Consider the fecundity of a fish versus its length

```{r sim_fecund, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
set.seed(514)
nn <- 100
ll <- runif(nn, 30, 80)
ee <- rnorm(nn, 0, 3)
b0 <- 5
b1 <- 1.2
ff <- (b0 + b1*ll + ee)^2
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0, 0),
    cex = 1)
## plot data
plot(ll, ff, pch = 16, xpd = NA,
     ylab = "Number of eggs", xlab = "Length (cm)", cex.lab = 1.5)
```


## Alternative to Box-Cox

Here's the fit from a linear regression

```{r sim_fecund_fit, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
## fit regression model
mm <- lm(ff ~ ll)
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0, 0),
    cex = 1)
## plot data
plot(ll, ff, pch = 16, xpd = NA,
     ylab = "Number of eggs", xlab = "Length (cm)", cex.lab = 1.5)
abline(coef(mm))
```


## Alternative to Box-Cox

And here are the residuals from the fitted model

```{r sim_fecund_resids, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0, 0),
    cex = 1)
## plot data
plot(fitted(mm), resid(mm), pch = 16, las = 1, xpd = NA,
     ylab = "Residuals", xlab = "Estimated fecundity", cex.lab = 1.5)
abline(h = 0, lty = "dashed")
```


## Alternative to Box-Cox

This $\hat{\lambda}$ is really close to 0.5 which is a square-root transform $(x^\frac{1}{2} = \sqrt{x})$

```{r boxcox_fecund, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0.3, 0),
    cex = 1.1)
## plot profile & estimate lambda
llp <- MASS::boxcox(mm, lambda = seq(-50,150)/100)
lm <- llp$x[which.max(llp$y)]
mtext(side = 3, line = 0.5, at = lm, substitute(hat(lambda) == lm, list(lm = lm)))
```


## Square root transformation

Here's the fit from a linear regression to $\sqrt{y}$

```{r fecund_fit_sqrt, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
## fit regression model
ms <- lm(sqrt(ff) ~ ll)
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0, 0),
    cex = 1)
## plot data
plot(ll, sqrt(ff), pch = 16, xpd = NA,
     ylab = expression(sqrt(Number~of~eggs)), xlab = "Length (dm)", cex.lab = 1.5)
abline(coef(ms))
```


## Square root transformation

And here are the residuals from the fitted model

```{r fecund_resids_sqrt, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0, 0),
    cex = 1)
## plot data
plot(fitted(ms), resid(ms), pch = 16, las = 1, xpd = NA,
     ylab = "Residuals", xlab = expression(sqrt(Estimated~number~of~eggs)), cex.lab = 1.5)
abline(h = 0, lty = "dashed")
```


## Predictions from a transformed model

Using `predict()` will give fitted values $(\hat{y})$ on the transformed scale

```{r ex_trans_fits, echo = TRUE}
## expected sqrt(fecundity) for length = 5 dm
predict(ms, data.frame(ll = 5), interval = "confidence")
```


## Predictions from a transformed model

We need to include the back-transformation on `predict()`

$$
\begin{aligned}
\sqrt{\hat{y}}_i &= x_i \hat{\boldsymbol{\beta}} \\
&\Downarrow \\
\hat{y}_i &= (x_i \hat{\boldsymbol{\beta}})^2 \\
\end{aligned}
$$


## Back-transformed fit

Here are the fitted values and prediction interval on the natural scale

```{r fecund_fit_sqrt_back, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
## back-transformed predictions
tp <- predict(ms, data.frame(ll = seq(30,80)), interval = "prediction")
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0, 0),
    cex = 1)
## plot data
plot(ll, ff, pch = 16, xpd = NA,
     ylab = expression(Number~of~eggs), xlab = "Length (dm)", cex.lab = 1.5)
matlines(seq(30,80), (tp)^2, 
         lty = c(1, 2, 2), lwd = 2, 
         col = c("blue", "red", "red"))
```


## Transformed polynomials

Think back to an early lecture where we transformed a nonlinear polynomial into a linear model

$$
y_i = \beta_0 + \beta_1 x_{1,i} + \beta_2 \color{blue}{x_{2,i}^2} + e_i \\
\Downarrow \\
y_i = \beta_0 + \beta_1 x_{1,i} + \beta_2 \color{blue}{z_{2,i}} + e_i \\
\color{blue}{z_{2,i}} = \color{blue}{x_{2,i}^2}
$$

## Transformed polynomials

Polynomials are an easy way to model nonlinearities in data, such as

* Seasonal effects on primary productivity

* Temperature effects on growth of poikilotherms


## Ecological data

Many ecological observations only take positive values $(y > 0)$

* length or mass

* fecundity

* species counts/density

* concentration

* latency periods for infectious diseases

<br>

The distributions of these data also tend to be "long-tailed"


## Long-tailed data

Distribution of plant diversity data in the `gala` dataset

```{r gala_hist, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
data(gala, package="faraway")
rr <- range(gala$Species)
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0, 0),
    cex = 1.1)
## plot data
hist(gala$Species, breaks = seq(0, 450, 50),
     xlab = "Number of species", ylab = "Number of islands",
     main = "")
```


## Long-tailed data

These long-tailed data often follow a log-normal distribution

```{r gala_hist_ln, echo = FALSE, fig.width = 5, fig.height = 4, fig.align="center"}
## set plot area
par(mai = c(1.0,1.0,0.1,0.1),
    omi = c(0, 0, 0, 0),
    cex = 1.1)
## plot data
hist(log(gala$Species), breaks = seq(0, 7, 0.5),
     xlab = "log(number of species)", ylab = "Number of islands",
     main = "")
```


## Log transformation

A log-transformation is a really common way to deal with ecological data that are constrained to be positive

$$
y_i = \exp(\beta_0 + \beta_1 x_{i} + e_i) \\
\Downarrow \\
\log(y_i) = \beta_0 + \beta_1 x_{i} + e_i
$$


## Log-log transformation

Consider allometric scaling laws in ecology of the form

$$
y_i = \alpha \cdot x_i^\beta \cdot \delta_i
$$

<br>

For example, body mass as a function of length


## Log-log transformation

Log-log transformations are an easy way to linearize power models

$$
m_i = \alpha \cdot l_i^\beta \cdot \delta_i \\
\Downarrow \\
\log(m_i) = \log(\alpha) + \beta ~ \log(l_i) + \log(\delta_i) \\
\Downarrow \\
y_i = \beta_0 + \beta_1 x_i + e_i
$$


## Linear model for size of fish

The response and predictor are linear on the log-log scale

```{r lw_regr, fig.align = 'center', fig.height = 4, fig.width = 4}
## simulated data
nn <- 35
L10_length <- log(runif(nn, 100, 400), 10)
L10_mass <- -5.3 + 3.1 * L10_length + rnorm(nn, 0, 0.2)
## plot them
par(mai = c(1, 1, 0.1, 0.1), omi = rep(0, 4), bg = NA)
plot(10^(L10_length), 10^(L10_mass), pch = 16, log = "xy",
     ylab = "Mass (g)", xlab = "Length (mm)", cex.lab = 1.2)
```


## Summary

* Box-Cox is good to help ID a power/root, but the transformed variable can be hard to interpret

> - $\sqrt{y}$ is good for equalizing variance

> - $\log(y)$ is good for skewed data

> - $\log(y + a)$ with small $a$ is good for skewed data with some 0's

> - We will see later that there are model alternatives to transformations (GLMs)





