---
title: "Generalized Linear Mixed Models, Part 1"
author: "Sarah Converse"
date: "28 May 2025"
output:
  ioslides_presentation
subtitle: Analysis of Ecological and Environmental Data<br>QERM 514
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(MASS)
library(lme4)
library(glmmTMB)

```


```{css, echo = FALSE}

.title-slide hgroup h1 {
  font-size: 40px;
}
.title-slide hgroup h2 {
  font-size: 30px;
}
h2 {
  font-size: 40px;
}
slides > slide {
  font-size: 20px;
}
```

## Goals for today

* Understand the structural components of generalized linear mixed models

* Understand some of the methods for fitting GLMMs


## Forms of linear models
 
```{r model_diagram_glmm, fig.height = 5, fig.width = 7, fig.align = 'center'}
par(mai = rep(0, 4), omi = rep(0, 4), bg = NA)

## order: GLM, LM, GLMM, LMM
xx <- c(1, 2, 2, 3) * 10
yy <- c(2, 1, 3, 2) * 10

plot(c(7, 33), c(7, 33), type = "n", xlim = c(7, 33), ylim = c(7, 33),
     xaxt = "n", yaxt = "n", xlab = "", ylab = "",
     bty = "n")
symbols(xx, yy, circle = rep(2, 4), inches = 0.5, add = TRUE,
        lwd = 2, fg = c("black", "black", "blue", "black"), bg = "white")
text(xx, yy, c("GLM", "LM", "GLMM", "LMM"), cex = 1.5,
     col = c("black", "black", "blue", "black"))
## from LM to GLM
arrows(xx[2]-2, yy[2]+2, xx[1]+2, yy[1]-2, length = 0.2)
text(15, 14, "multiple forms of errors", pos = 2)
## from LM to LMM
arrows(xx[2]+2, yy[2]+2, xx[4]-2, yy[4]-2, length = 0.2)
text(25, 14, "multiple random processes", pos = 4)
## from GLM to GLMM
arrows(xx[1]+2, yy[1]+2, xx[3]-2, yy[3]-2, length = 0.2)
text(15, 26, "multiple random processes", pos = 2)
## from LMM to GLMM
arrows(xx[4]-2, yy[4]+2, xx[3]+2, yy[3]-2, length = 0.2)
text(25, 26, "multiple forms of errors", pos = 4)
```


## Generalized linear mixed model

GLMMs combine the flexibility of non-normal distributions (GLMs) with the ability to address correlations among observations and nested data structures (LMMs)

<br> 

## Generalized linear mixed model

GLMMs combine the flexibility of non-normal distributions (GLMs) with the ability to address correlations among observations and nested data structures (LMMs)

<u>Good news</u>

* these extensions follow similar methods to GLMs and LMMs

<u>Bad news</u>

* these models are on the frontier of statistical research

* existing documentation is rather technical

* multiple approaches for fitting models, some with different results


## Generalized linear mixed model

Just like GLMs, GLMMs have three components:

1. Distribution of the data: $f(y; \theta)$

2. Link function: $g(\eta)$

3. Linear predictor: $\eta$


## Generalized linear mixed model

Just like GLMs, we can assume a variety of data distributions:

* Gaussian - for normal outcomes 

* Bernoulli - for binary outcomes 

* binomial - for binomial outcomes

* beta binomial - for extra-binomial variation  

* gamma - for continuous positive outcomes  

* Poisson - for discrete positive outcomes  

* negative binomial - for extra-Poisson variation 

* zero-inflated Poisson and negative binomial

* and more... 

## Linear predictor for a GLM

We can write the linear predictor for GLMs as

$$
\eta = \beta_0 + \beta_1 x_1 + ... + \beta_k x_k \\
\Downarrow \\
\eta = \mathbf{X} \boldsymbol{\beta}
$$

where the $\beta_i$ are fixed effects of the covariates $x_i$


## Linear predictor for a GLMM

For GLMMs, our linear predictor also includes random effects

$$
\eta = \beta_0 + \beta_1 x_1 + \dots + \beta_k x_k  + \alpha_0 + \alpha_1 z_1 + \dots + \alpha_l z_l\\
\Downarrow \\
\eta = \mathbf{X} \boldsymbol{\beta} + \mathbf{Z} \boldsymbol{\alpha}
$$

where the $\beta_i$ are fixed effects of the covariates $x_i$


## Generalized linear mixed model

Survival of fish $s_{i,j}$ as a function of length $x_{i,j}$ in some location $j$

<br>

$$
\begin{aligned}
\text{data distribution:} & ~~~ y_{i,j} \sim \text{Binomial}(N_{i,j}, s_{i,j}) \\
\text{link function:} & ~~~ \text{logit}(s_{i,j}) = \text{log}\left(\frac{s_{i,j}}{1-s_{i,j}}\right) = \mu_{i,j} \\
\text{linear model:} & ~~~ \mu_{i,j} = \beta_0 + \beta_1x_{i,j} + \alpha_j\\
& ~~~ \alpha_j \sim \text{N}(0, \sigma_{\delta}^2)
\end{aligned}
$$


## Generalized linear mixed model

Fitting GLMMs is complex

* Why? Because GLMMs involve solving an integral with no analytical solution

Recall that we think of likelihoods in terms of the *observed data*

* But the random effects in our model are *unobserved* random variables, so we need to integrate them out of the likelihood

Best practices suggest we try to keep things simple - lots and lots of random effects become hard to handle 


## Likelihood for GLMMs

There is an integral in the likelihood: 

$$
\int f(Y|u)f(u)\partial u
$$

We want to compute the average likelihood across all possible values of the random effect $(u)$, weighted by their probabilities 

The challenge is computing this complex likelihood including the integral 

With LMMs, when $u \sim$ Normal and $\epsilon \sim$ Normal, there is an analytical solution, but often not true with non-normal errors – so we can’t easily get away from the integral 


## Approaches to fitting GLMMs

To avoid the integral, we will consider 3 methods that approximate the likelihood

They all have pros and cons so it's not possible to pick the "best"


## Penalized quasi-likelihood

Produces a linearized version of the response and then uses LMM methods to fit it 

<u>Pros</u>

* fast, flexible, and widely implemented

<u>Cons</u>

* not a true likelihood, making inference challenging 

* only asymptotically correct

* biased for binomial and Poisson with small samples


## Laplace approximation

Uses Laplace approximation for computing integrals of the form $\int e^{h(x)}\partial x$, which is a good approximation of the integral in a GLMM likelihood - we need to find only the maximum of the function $h(x)$ and the second derivative of $h(x)$ at the maximum   

<u>Pros</u>

* approximation of true likelihood rather than quasi-likelihood

* more accurate than PQL

<u>Cons</u>

* slower and less flexible than PQL

* may be impossible to compute for complex models


## Gauss-Hermite quadrature

Gauss-Hermite quadrature is an expansion of Laplace approximation where the integrand is evaluated at more than one point

<u>Pros</u>

* More accurate than Laplace

<u>Cons</u>

* Slow and computationally intense

* Limited to a few random effects (one in practice)


## Markov Chain Monte Carlo 

Sampling from the posterior distribution, as in MCMC, is an alternative

* This is what is done in Bayesian estimation 

* MCMC has the ability to handle much more complex random effects


## Practically, how will we fit these models?  

There are almost endless packages for doing this. Some of the most widely used (for frequentist estimation) are: 

* Penalized quasi-likelihood: `MASS::glmmPQL` 

* Laplace approximation: `lme4::glmer`, `lmer4::glmer.nb`, `glmmTMB::glmmTMB` 

* Gauss-Hermite quadrature: `lme4::glmer`, `lmer4::glmer.nb`  



## An example

Let's consider a long-term study of invasive brown treesnakes on Guam

* Introduced to the island shortly after WWII

* Voracious predators on native birds and other vertebrates

<br>

Our data consist of the number of eggs per female at 10 locations over 13 years

We are interested in the fixed effect of body size and the random effects of location and year


## Brown treesnakes

```{r echo = FALSE, results = FALSE, fig.height = 4.5, fig.width = 6, fig.align = 'center'}

#gravid snakes  
set.seed(3)
n <-  587
n.loc <- 10
n.yrs <- 13
grand.mean <- 1
slope <- 0.5

loc.eff <- rnorm(n.loc,0,1.1)
year.eff <- rnorm(n.yrs,0,0.5)
loc <- sample(c(1:n.loc),n,replace=TRUE)
year <- sample(c(1:n.yrs),n,replace=TRUE)
predictor <- runif(n,-2,2)
size <- predictor

#number of eggs as a function of body size, location random effect, year random effect 
mean <- exp(grand.mean + loc.eff[loc] + year.eff[year] + slope*size)

eggs <- rpois(587,mean)
egg.data <- data.frame(cbind(eggs, size, loc, year))

par(mai = c(1,1,0.5,0.5))
hist(eggs, breaks=12, cex.lab= 1.5, cex.axis=1.5, main = "", xlab = "Number of Eggs")

cntrl <- glmerControl(optimizer = "bobyqa", tol = 1e-4, optCtrl=list(maxfun=100000))

```

## Visualize the data 

Colors represent years 

```{r echo = FALSE, results = FALSE, fig.height = 4.5, fig.width = 6, fig.align = 'center'}

plot(eggs ~ size, data = egg.data, col = year)

```

## Visualize the data 

Colors represent locations

```{r echo = FALSE, results = FALSE, fig.height = 4.5, fig.width = 6, fig.align = 'center'}
plot(eggs ~ size, data = egg.data, col = loc)

```


## Let's fit some models 

We'll begin with the effects of body size and location only

* Using penalized quasi-likelihood 

```{r, echo = TRUE, results = TRUE}

start1 <- Sys.time()
m1.pql <- MASS::glmmPQL(eggs ~ size, random = ~ 1|loc, data = egg.data, 
                        family = poisson)
(elapsed1 <- Sys.time() - start1)

```

## Let's fit some models 

* Using LaPlace approximation

```{r, echo = TRUE, results = TRUE}

start2 <- Sys.time()
m1.lap <- lme4::glmer(eggs ~ size + (1|loc), data = egg.data, 
                      family = poisson, control = cntrl)
(elapsed2 <- Sys.time() - start2)
```

## Let's fit some models 

* Using Gauss-Hermite quadrature

```{r, echo = TRUE, results = TRUE}

start3 <- Sys.time()
m1.ghq <- lme4::glmer(eggs ~ size + (1|loc), data = egg.data, 
                      family = poisson, nAGQ = 40, control = cntrl)
(elapsed3 <- Sys.time() - start3)

``` 

## Brown treesnakes

Summary of the results from the 3 methods

```{r, echo = FALSE, results = TRUE}
library(insight)
tbl_smry <- cbind(summary(m1.pql)$tTable[, 1:2],
                  summary(m1.lap)$coefficients[, 1:2],
                  summary(m1.ghq)$coefficients[, 1:2])
tbl_smry <- rbind(tbl_smry,"location SD" = 
                    c(c(sqrt(as.numeric(VarCorr(m1.pql)[1,2])), NA),
                      c(sqrt(VarCorr(m1.lap)$loc[1,1]), NA),
                      c(sqrt(VarCorr(m1.ghq)$loc[1,1]), NA)))

colnames(tbl_smry) <- c("PQL", "SE", "Laplace", "SE", "GHQ", "SE")
round(tbl_smry, 3)
```

Simulated truth: 

Intercept = 1 

Size effect = 0.5  

Location SD = 1.1 

## Brown treesnakes

What if we also want to include the random effect of year?

## Laplace approximation

```{r, echo = TRUE, results = TRUE}

start2 <- Sys.time()
m2.lap <- glmer(eggs ~ size + (1|loc) + (1|year), data = egg.data, 
                family = poisson)
(elapsed2 <- Sys.time() - start2)

```

## Brown treesnakes

Summary of the results using LaPlace (practically not possible to fit crossed random effects in glmmPQL, GHQ not available with two random effects)

```{r, echo = FALSE, results = TRUE}
library(insight)
tbl_smry <- cbind(
                  summary(m2.lap)$coefficients[, 1:2])

tbl_smry <- rbind(tbl_smry,
                  "location SD" = c(sqrt(VarCorr(m2.lap)$loc[1,1]), NA),
                  "year SD" = c(sqrt(VarCorr(m2.lap)$year[1,1]), NA))

colnames(tbl_smry) <- c("Laplace", "SE")

round(tbl_smry, 3)
```

Simulated truth: 

Intercept = 1 

Size effect = 0.5  

Location SD = 1.1

Year SD = 0.5

## Summary of GLMM methods

Simpler models will work better 

LaPlace approximation likely to work best when we have multiple random effects 

With only one random effect, Gauss-Hermite quadrature likely to work best 

With very complex models, some things may not be possible in ML framework 

Common packages: 

* `MASS:glmmPQL` - offers PQL

* `lme4::glmer` - offers LaPlace approximation or GHQ, most widely used so lots of help available, negative binomial with `lme4::glmer.nb` 

* `glmmTMB::glmmTMB` - offers LaPlace approximation only, allows for a wider variety of distributions than `lme4`, and offers both zero-inflation and some simple correlation structures 


## Additional Resources  

https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html

