---
title: "Generalized Linear Mixed Models, Part 2"
author: "Sarah Converse"
date: "30 May 2025"
output:
  ioslides_presentation
subtitle: Analysis of Ecological and Environmental Data<br>QERM 514
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(lme4)
library(car)
library(DHARMa)
library(faraway)

```


```{css, echo = FALSE}

.title-slide hgroup h1 {
  font-size: 40px;
}
.title-slide hgroup h2 {
  font-size: 30px;
}
h2 {
  font-size: 40px;
}
slides > slide {
  font-size: 20px;
}
```

## Goals for today

* Understand some of the diagnostics available for evaluating GLMM fits

* Become familiar with some of the inferential tools for GLMMs


## Diagnostics

Diagnostics for GLMMs are similar to those for GLMs, but somewhat more limited

## Returning to our brown treesnake data 

Let's consider a long-term study of invasive brown treesnakes on Guam

* Introduced to the island shortly after WWII

* Voracious predators on native birds and other vertebrates

<br>

Our data consist of the number of eggs per female at 10 locations over 13 years

We are interested in the fixed effect of body size and the random effects of location and year


## Brown treesnakes

```{r echo = FALSE, results = FALSE}

#gravid snakes  
set.seed(3)
n <-  587
n.loc <- 10
n.yrs <- 13
grand.mean <- 1
slope <- 0.5

loc.eff <- rnorm(n.loc,0,1.1)
year.eff <- rnorm(n.yrs,0,0.5)
loc <- sample(c(1:n.loc),n,replace=TRUE)
year <- sample(c(1:n.yrs),n,replace=TRUE)
predictor <- runif(n,-2,2)
size <- predictor

#number of eggs as a function of body size, location random effect, year random effect 
mean <- exp(grand.mean + loc.eff[loc] + year.eff[year] + slope*size)

eggs <- rpois(587,mean)
egg.data <- data.frame(cbind(eggs, size, loc, year))

par(mai = c(1,1,0.5,0.5))
hist(eggs, breaks=12, cex.lab= 1.5, cex.axis=1.5, main = "", xlab = "Number of Eggs")

cntrl <- glmerControl(optimizer = "bobyqa", tol = 1e-4, optCtrl=list(maxfun=100000))

```

## Fit some models 


```{r echo = TRUE, results = TRUE}

m1.lap <- glmer(eggs ~ size + (1|loc), data = egg.data, 
                family = poisson, control = cntrl)

m2.lap <- glmer(eggs ~ size + (1|loc) + (1|year), data = egg.data, 
                family = poisson, control = cntrl)

```


## Goodness of fit

Recall our goodness of fit test based on the Pearson's $\chi^2$

$$
X^2 = \sum_{i = 1}^n \frac{(O_i - E_i)^2}{E_i} \sim \chi^2_{(n - 1)}
$$

<br>

where $O_i$ is the observed count and $E_i$ is the expected count


## Pearson's $\chi^2$ statistic

For a binomial distribution

$$
X^2 = \sum_{i = 1}^n \frac{(y_i - n_i \hat{p}_i)^2}{n_i \hat{p}_i}
$$

For a Poisson distribution

$$
X^2 = \sum_{i = 1}^n \frac{(y_i - \lambda_i)^2}{\lambda_i}
$$


## Goodness of fit

$H_0$: Our model is correctly specified

```{r, echo = TRUE, results = TRUE}
X2 <- sum((eggs - fitted(m2.lap))^2 / fitted(m2.lap))
pchisq(X2, df = n - length(coef(m2.lap)),lower.tail = FALSE)
```

The $p$-value is large so we do not reject $H_0$


## Model diagnostics

* Check for heteroscedasticity

```{r echo = TRUE, results = FALSE, fig.width=8, fig.height=3.5, fig.align='center'}

plot(m2.lap, sqrt(abs(resid(.))) ~ fitted(.), type = c("p", "smooth")) 

```


## Diagnostics 

* Use `DHARMa::simulateResiduals` to further visualize residuals 

```{r echo = FALSE, results = FALSE, fig.width=8, fig.height=3.5, fig.align='center'}

simulationOutput <- simulateResiduals(m2.lap, n = 1000)
plot(simulationOutput)

``` 

## Model diagnostics

* QQ plot for random effects 

```{r echo = FALSE, results = FALSE, fig.width=8, fig.height=3.5, fig.align='center'}
## set plot area
par(mai = c(0.9, 0.9, 0.4, 0.1),
    omi = c(0, 0, 0, 0),
    mfrow = c(1,3), cex.lab = 1.5)

## qq RE's
qqnorm(unlist(ranef(m2.lap)), main = "QQ plot (RE's)", pch = 16)
qqline(unlist(ranef(m2.lap)))

```


## Leverage

For other models, we can calculate the leverages to evaluate potentially extreme values in predictor space

For GLMMs, however, the leverages depend on the estimated variance-covariance matrices of the random effects

We can fit a GLM and evaluate leverage of the fixed effects 

```{r echo = FALSE, results = FALSE, fig.width=8, fig.height=3.5,fig.align='center'} 

m.glm <- glm(eggs ~ size, data = egg.data, family = poisson)

halfnorm(hatvalues(m.glm))
```

## Cook's Distance

For other models, we can calculate Cook's distances to identify potentially influential data points

For GLMMs, however, the Cook's distances involve derivatives of the likelihood with respect to the random effects (this is an active area of research)

Again, as a partial measure, we can look at the Cook's distances for the generalized linear model  

```{r echo = FALSE, results = FALSE, fig.width=8, fig.height=3.5,fig.align='center'}

halfnorm(cooks <- cooks.distance(m.glm))

large.cooks <- sort(cooks,decreasing=TRUE)[1:2]       
```

## Display large Cook's distances (from GLM)

```{r echo = FALSE, results = TRUE, fig.width=8, fig.height=3.5,fig.align='center'}

egg.data[as.numeric(names(large.cooks)),]

```

Large snakes with lots of eggs 


## Inference for fixed effects {.smaller}

We can evaluate our fixed effects with AIC - as long as we keep the random effects unchanged  

```{r echo = TRUE, results = TRUE}

m1.null <- glmer(eggs ~ (1 | loc) + (1 | year), data = egg.data, family = poisson)

AIC(m1.null)
AIC(m1.lap)

```


## Inference for random effects {.smaller}

We can test the significance of the random effects using a bootstrap LRT 

A significant p-value leads us to reject the null hypothesis that the simpler model is adequate 

```{r echo = TRUE, results = TRUE}

m2.lap <- glmer(eggs ~ size + (1|loc) + (1|year), data = egg.data, family = poisson)
m1.lap <- glmer(eggs ~ size + (1|loc), data = egg.data, family = poisson)
lrstat <- 2*(logLik(m2.lap)-logLik(m1.lap)) 

cntrl <- glmerControl(check.conv.singular = .makeCC(action = "ignore",  tol = 1e-4))
b.lrstat <- numeric(100)
for(i in 1:100){
  y <- unlist(simulate(m1.lap))
  bnull <- update(m1.lap, y ~ ., control=cntrl)
  balt <- update(m2.lap, y ~ ., control=cntrl)
b.lrstat[i] <- as.numeric(2*(logLik(balt)-logLik(bnull)))
}
(p <- mean(b.lrstat > lrstat))

```

## Inference

For confidence intervals, we will again use bootstrapping

* We will cover this in lab  


## What if we have evidence of overdispersion?  

As with GLMs, we can handle overdispersion in binomial or Poisson data  

Let's check our snake model applied to a new data set

```{r echo = FALSE, results = FALSE}
egg.data2 <- egg.data 
egg.data2$eggs <- rnbinom(n, mu = mean, size = 5) 
```


```{r, echo = TRUE, results = TRUE}
new.m2.lap <- glmer(eggs ~ size + (1 | loc) + (1 | year), 
                    family = poisson, data = egg.data2)

k <- length(coef(new.m2.lap)) + length(ranef(new.m2.lap))
pchisq(deviance(new.m2.lap), k, lower.tail = FALSE)

```

## Brown treesnakes | Negative binomial

Fitting a negative binomial model 

```{r, echo = TRUE, results = FALSE}
new.m2nb.lap <- glmer.nb(eggs ~ size + (1 | loc) + (1|year), 
                        data = egg.data2, control = cntrl)
```

## Brown treesnakes {.smaller}

```{r, echo = FALSE, results = TRUE}
summary(new.m2nb.lap)
```

## Summary 

* Diagnostics for GLMMs are an art rather than a science 

* Use LRTs to test random effects (or just leave them in)

* Using robust inferential methods, like bootstrapped confidence intervals 


